---
title: "Regression tables with tab_reg()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression tables with tab_reg()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option
library(dplyr)

# As in the introduction vignette, tables render as tabxplor's real html tables (the recommended
# everyday setting); the shared stylesheet is emitted once by tab_css() below, tooltips kept off.
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

options(cli.num_colors = 256)
set_color_palette(theme = "light")
```

```{r, echo = FALSE, results = "asis"}
cat(tab_css(theme = "light"))
```

```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

For the most common regression models, `tab_reg()` builds a **regression table** that looks and behaves like a `tabxplor` cross-table: one row per predictor level, significance stars, colours that grey out non-significant effects, and the same Excel, html or markdown exports. You give it a data frame, an **outcome** (`dependent`) and some **predictors**, and it tries to pick the right kind of model from the outcome's type. Its distinctive feature is `empirical = TRUE`, which shows the **observed / crude / empirical effect right next to the model's adjusted one**, so you can see what "controlling for the other variables" actually changed.

We use a formatted version of the `forcats::gss_cat` data, from the US General Social Survey:

```{r}
gss_simple <- gss_cat_data_formatting()
```


## The outcome's type chooses the model and the observed quantity to compare with

You rarely set `family` by hand — `tab_reg()` mostly detects it: 

| Outcome | Detected model | Default effect measure |
|:--------|:---------------|:---------------|
| 2-level factor | binomial (logistic) | odds ratio (OR) |
| numeric (continuous) | gaussian (linear) | coefficient β |
| count | poisson | incidence-rate ratio (IRR) |
| 3+ level unordered factor | multinomial | one OR column per category vs. the reference |
| 3+ level ordered factor | ordinal (proportional odds) | cumulative OR |

That last column is only the **default**: each family can report several measures, and on several contrasts. The full grid, and how to ask for another one, is in [What a model column reports](#what-a-model-column-reports-family-effect-measure) below.

With `empirical = TRUE`, each model column is joined by a crude/observed companion column showing the *observed, unadjusted (univariable)* effect — the effect you would see with no controls at all, "all things being unequal", for that predictor.

One rule covers every case: **the observed effect is the effect of that predictor alone**. With a categorical predictor that is exactly the observed contrast between levels (a percentage difference, an odds ratio computed from the raw percentages). With a *continuous* predictor there is no such shortcut, so it is the univariable slope — which assumes the effect is linear on the model's scale, something worth checking (with `cut()`, or splines) before trusting it.

- 2 levels logistic (binomial) → observed %, and observed odd-ratios
- 3+ levels logistic (multinomial) → the observed ORs are shown as a tooltip on the model cells in html exports
- gaussian (linear) → group means and their difference
- poisson (counts) → observed rate and observed rate ratio

A continuous predictor has no levels, so only its **effect** column is filled; the base column (`Obs_%`, `Obs_mean`, `Obs_rate`) stays empty, and its distribution — mean, standard deviation, and the mean in each outcome group — is shown in the html tooltip instead.

On **weighted** data, one thing is worth knowing from the start: an observed column is always measured exactly like the model column beside it, so its confidence interval accounts for the weighting (and, under a `survey` design, for the whole design). A `tab()` cross-table does not, by default — add `design_effect = TRUE` there if you want its percentages directly comparable with these. See [Weighted and survey data](#weighted-and-survey-data) below.


## Logistic regression (a binary factor)

When the outcome is a two-level factor, here "married" versus "not married", `tab_reg()` chooses a **binomial** family to fit a logistic regression and reports **odds ratios** (the reference level of each predictor shows the neutral value `1`). 

- Like in every regression model, the effect of a predictor level on the dependent variable reads "all the others chosen predictors being equals".
- Colors read like any `tabxplor` table: an odds ratio above 1 (blue) means *more likely to be married* than the reference level, below 1 (red) means *less likely*; stars and colors both flag significance.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"))
```

For a simple logistic regression, `empirical = TRUE` add **raw percentages** and **crude unmodelised odds-ratios** for each predictor.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), empirical = TRUE)
```

The raw percentages `Obs_%` are a summary of the base observed results everything in the model is derived from :"28% of black americans are married, compared to 51% for white americans"

- The plain difference from reference is colored (28% - 51% = -23%).
- Cells that are non-significantly different from the reference are greyed out (based on a Newcombe confidence interval for differences of proportions).

The **modelised odds-ratios** `Model_OR` directly compared to the **observed odds-ratios** `Obs_OR` :

- Comparing the two tells you what adjustment the model did: if a predictor's **model** OR is much closer to 1 than its **crude** OR, the raw association was largely explained by the other predictors.
- Here, "all things being unequal", black americans have 2.68 lower odds to be married than white americans. "All things being equal" (more precisely : income, age and religion being equal), black americans still have 2.4 lower odds to be married than white americans. The result holds, it’s not explained by income differences or religious differences.

The **observed odds-ratios** `Obs_OR` are the same as those you can calculate from percentages only in a crosstable :

- Colors and significance stars use a **Woolf OR confidence interval** that matches what is done in the regression model. 
- The population of the table must match the complete-cases population of the model, filtering out indivivuals with `NA` at any involved variable.

```{r}
gss_simple |> 
  dplyr::filter(dplyr::if_all(all_of(c("race", "age", "rincome", "relig")), ~ !is.na(.) )) |> 
  tab(race, married, pct = "row", na = "drop", 
    display = "{or}", ref = "first", color = "odds_ratio", color_signif = "grey_non_signif"
   )
```

### Average marginal effects (AME) of a logistic regression

Another way to interpret a logistic regression is to use **average marginal effects** (AME) with `effect = "marginal"` :

- Instead of directly reading the modelised odd-ratio we can infer, from it, the average difference of percentage of each level compared to the reference.
- Here `Model_AME` reads like this : "Comparing Black and White respondents who are alike in income, age and religion, being Black is associated with a marriage rate 19.8 points lower, on average."

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), effect = "marginal", empirical = TRUE)
```

Then we can use `empirical = TRUE` to compare this new modelised quantity to the **raw, observed difference of percentages** : 

- In the sample (`Obs_diff`), we observe a rate of marriages 23 points lower for black americans compared to white american (`28% - 51% = -23%`).
- The modelised difference `Model_AME` is very close, `-19.8%`, so very little of this difference between black and white can be explained by age, income or religion.

We can also compare the raw percentage `Obs_%` with the **model-adjusted probability** — the value in parentheses in the `Model_AME (adjusted %)` column :

- Adjusted percentage, which is a result of the model, reads this way : "if the whole sample kept its income/age/religion distribution but everyone were Black, an estimated 31.5% would be married, vs 51.3% if everyone were White". 
- Blacks marry at 30.9% in this sample (observed); standardizing their income/age/religion to the population mix moves that only to 31.5% — so almost none of the Black–White gap is accounted for by differences in those three variables; it persists after adjustment.

We can, again, get the same empirical differences and confidence intervals with a plain table : 

```{r}
dependent  <- "married"
predictors <- c("race", "age", "rincome", "relig")
factor_predictors <- c("race", "rincome", "relig")
gss_simple |>
  dplyr::filter(dplyr::if_all(all_of(c(predictors, dependent)), ~ !is.na(.) )) |> 
  tab(all_of(factor_predictors), all_of(dependent), pct = "row", na = "drop", 
      color = "difference", ref = 1, color_signif = "grey_non_signif", stars = TRUE, 
      ci_method = c(diff = "wald")
      ) |>
  dplyr::mutate(diff = set_display(`01-Married`, "diff"))
```



### Risk ratios: "how many times more likely?"

An odds ratio is **not** a "times more likely". The two are close only when the outcome is **rare**; as
soon as it is common — above roughly 10%, which covers most survey questions — the odds ratio sits much
further from 1 than the ratio of probabilities, so reading `Model_OR = 0.41` as "41% as likely to be
married" overstates the gap. There is a second reason to care: an odds ratio is *non-collapsible*, which
means it moves when you add a covariate **even if that covariate is not a confounder** — so comparing
`Model_OR` across nested models is not valid. Risk ratios do not have that problem.

tabxplor gives you two ways to get a **risk ratio** (RR), and on real data they agree closely.

The first keeps the familiar logistic model and just asks for the effect as a ratio, with
`effect = "marginal", measure = "ratio"`. Each cell shows the risk ratio and, in parentheses, the same adjusted predicted
probability as the AME column:

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"),
        effect = "marginal", measure = "ratio", empirical = TRUE)
```

`Model_RR` now reads plainly: "comparing Black and White respondents alike in income, age and religion,
Black respondents are about 0.61 times as likely to be married" — and `Obs_RR` is the same quantity
computed straight from the data, with no model. The cell is coherent by construction: the reference
probability multiplied by the risk ratio gives the level's probability (`51.3% × 0.61 ≈ 31.5%`), which is
the multiplicative counterpart of the additive AME identity above.

The second route changes the model itself: with a **binary outcome**, `family = "poisson"` fits a
*modified Poisson* regression (Zou 2004), whose coefficients are risk ratios directly:

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"),
        family = "poisson", empirical = TRUE)
```

This is strictly opt-in — a binary outcome still defaults to logistic — and tabxplor tells you when it
switches. Which to choose? `effect = "marginal", measure = "ratio"` keeps the well-behaved logistic fit and can never
predict a probability above 100%; `family = "poisson"` is the form epidemiologists will recognise, and it
is unaffected by the caveat that a marginal ratio is standardized to the sample at hand (so under
`split_var` each group standardizes to its own subpopulation). Neither is for prediction: with the
Poisson route the log link is unbounded, so fitted probabilities can exceed 1.

For readers who want the technical detail: standard errors are handled consistently in both cases.
Fitting a Poisson likelihood to a 0/1 outcome is a deliberate misspecification, so the naive standard
errors would be too wide; tabxplor replaces them with the robust **Huber–White sandwich**, via
`survey::svyglm()` — which means the design-based variance when you supply survey weights, and the
equivalent of `HC0` otherwise. The observed companion follows the same estimand throughout: `Obs_RR` is
the crude risk ratio with a Katz interval, never the crude odds ratio. The modified Poisson needs a
reasonable sample (n of at least about 100), and since a quasi-likelihood has no true likelihood its
footer reports N and a Wald-vs-null test instead of AIC/BIC.

## Colouring what adjustment did

`empirical = TRUE` puts the crude effect next to the modelled one, but comparing them cell by cell is
tedious on a real table. `color = "adjustment"` colours the **gap** between the two, so a whole table of
"what did adjusting change?" reads at a glance. Put it on the *background* channel and the text keeps
showing the effect size, so one look answers both questions:

```{r}
tab_reg(gss_simple, "married", c("race", "rincome", "relig"),
        empirical = TRUE, color = c(TRUE, "adjustment"))
```

Text colour = how strong the adjusted odds ratio is. Background = how far it sits from the observed
one: one pole means adjustment **strengthened** the effect, the other that it **attenuated** it. The
first threshold is ×1.1, the classic "a 10% change in the estimate is worth noticing" convention; the
rest are ×1.25, ×1.5 and ×2 (change them with `set_color_breaks(adj_ratio = ...)`). An *additive*
effect gets an additive ladder instead: percentage points for a marginal effect on a probability
(`±2 / ±5 / ±10 / ±20`), and standard deviations of the outcome for a linear β
(`±0.05 / ±0.1 / ±0.2 / ±0.4`) — so the same model reads the same way whether the outcome is recorded
in hours, minutes or days. You do not have to ask
for `empirical` separately — the colour needs the observed effect, so it turns it on for you.

The direction is measured **from the null**, not up or down. A protective effect (OR below 1) that
adjustment pulls toward 1 gets the same colour as a risky effect pulled toward 1, which is what you
want to read: "the other variables explained part of this association" is one statement, whichever
side of 1 the effect is on.

The same gap is readable as a number when you want the exact values, in the cell or on hover.
`display =` writes the template at build time (see *Reading the table* below); on a finished table
`set_display()` does the same — the columns that have no observed counterpart simply keep their
plain display:

```{r}
tab_reg(gss_simple, "married", c("race", "rincome"), empirical = TRUE) |>
  set_display("{or} (obs {obs})")
```

### One warning about odds ratios

An odds ratio changes when you adjust it **even when there is nothing to adjust for**. This is called
*non-collapsibility*: it is a property of the odds ratio itself, not a sign of confounding. In a
simulation where the extra variable is independent of the exposure — so there is strictly no
confounding — the crude odds ratio still moved by about **8%** once adjusted, while the risk ratio moved
by 0.3% and the marginal effect by essentially nothing. Eight percent is about the size of the first
colour step, so on the odds-ratio scale a faint background colour may be arithmetic rather than
confounding. The legend of an exported table says so.

The comparisons that *are* clean are the ones this vignette has already introduced: `effect = "marginal"`
and `effect = "marginal", measure = "ratio"` (marginal effects), `measure = "ratio"` on a binary outcome (risk ratios),
and the beta of a linear regression. On those scales the gap is confounding by the variables you added
— nothing else. If reading confounding is the point of your table, prefer one of them:

```{r}
tab_reg(gss_simple, "married", c("race", "rincome", "relig"),
        effect = "marginal", measure = "ratio", empirical = TRUE, color = c(TRUE, "adjustment"))
```

### Is the gap bigger than noise?

A background colour tells you the model *moved* an effect. It does not, on its own, tell you the move
is real: with 20 000 respondents a tiny shift is solid, with 300 a large one may be luck. `color_signif`
answers that, exactly as it does everywhere else in `tabxplor` — add `"grey_non_signif"` and a coloured
background now means "the model really changed this effect":

```{r}
tab_reg(gss_simple, "married", c("race", "rincome", "relig"),
        family = "poisson", empirical = TRUE,
        color = c(TRUE, "adjustment"), color_signif = "grey_non_signif")
```

Here `measure = "ratio"` on a binary outcome gives **risk ratios** (the modified Poisson), one of the clean scales from the
previous section. Text colour = how strong the adjusted risk ratio is. Background = how far it sits
from the observed one, greyed out when that distance could be chance. Hovering a cell in the html table
gives the exact numbers: the observed effect, the size of the gap, its confidence interval and its
p-value.

**One example is worth the whole explanation.** Read the `race` block of that table, comparing each
group with white respondents, before and after holding income and religion equal:

| scale | Black vs White | Other vs White |
|:------------------------------|---------------:|---------------:|
| odds ratio, crude → model | 0.41 → 0.39 | 0.90 → 1.01 |
| risk ratio, crude → model | 0.59 → 0.59 | 0.95 → 1.01 |
| did the risk ratio really move? | **no** (p = 0.5) | **yes** (p < 0.001) |

Two different stories, and the test separates them. For Black respondents nothing moves: their much
lower marriage rate is *not* explained by income or religion — it survives the adjustment untouched.
For the "Other" group the crude 5% deficit disappears once income and religion are held equal, and the
gap between the two numbers is far too big to be luck: that difference *was* explained by them.

Notice that the odds ratios move too — including for Black respondents, where the risk ratio says
nothing happened. That is the non-collapsibility of the previous section, showing up as a phantom
"change". It is why `color_signif` is **not applied to plain odds ratios**: `tab_reg()` leaves those
colours descriptive and says so once, pointing at `effect = "marginal"`, `effect = "marginal", measure = "ratio"` or
`family = "poisson"`.

The greying earns its keep on smaller samples. With 20 000 respondents almost every visible move is
real; on 2 000 the same table keeps one background colour out of three:

```{r}
set.seed(1)
small <- gss_simple[sample(nrow(gss_simple), 2000), ]
tab_reg(small, "married", c("race", "rincome", "relig"),
        family = "poisson", empirical = TRUE,
        color = c(TRUE, "adjustment"), color_signif = "grey_non_signif")
```

### Three ways to get this wrong

**"The two confidence intervals overlap, so nothing changed."** They almost always overlap, because the
model and the observed effect are computed from the *same people* and move together (a correlation of
0.5 to 0.9). Reading the overlap would declare "no change" essentially always. The test in the table
does the right thing instead: it measures the difference directly, which is why its interval is much
narrower than you would guess from the two printed ones.

**"A significant gap proves confounding."** It proves the two numbers differ by more than noise. Why
they differ is your reading, not the table's — and on a non-collapsible scale part of the difference is
arithmetic, which is the whole point of the previous section.

**"Every coloured cell counts."** Each cell is tested on its own, with no correction for the number of
tests. In a table with seven comparisons, about one in five will show a spurious coloured cell. Read
the pattern — a whole block of predictors shifting the same way — rather than one isolated cell.

### For the record: what exactly is tested

The null is that the modelled and observed effects are equal, on the effect's own scale: the log-ratio
for a risk ratio, an odds ratio or an incidence-rate ratio, the plain difference for a beta or a
marginal effect. That is the scale the colour already folds around, so the test and the colour cannot
drift apart.

The two estimates come from the same rows, so they are correlated and their difference has a smaller
standard error than either alone. The correct variance is the sampling variance of the difference of
their *influence functions* — what Stata calls seemingly-unrelated estimation (Weesie 1999; Mize, Doan
& Long 2019 is the sociological statement). Two things make it exact rather than approximate here:
every observed effect `tabxplor` prints *is* the coefficient of a saturated one-predictor model, so its
influence function is a closed-form expression whose standard error is the interval already shown; and
with survey weights or a design the variance goes through `survey`'s own linearization, so strata,
clusters and finite-population corrections are respected. The reference distribution is the normal
(z), which is mildly conservative in small samples.

The three policies read that same interval:

| `color_signif` | what the background shows |
|----------------|---------------------------|
| `"ignore"` (default in `tab()`) | the size of the gap, coloured whether or not it is significant |
| `"grey_non_signif"` (default in `tab_reg()`) | the same, greyed unless the gap's interval excludes "no change" |
| `"guaranteed_effect"` | the **floor** of the gap: "adjustment moved this effect by at least ×1.1" (on a difference scale, "by at least 2 points") |

**Stars and colour do not say the same thing, and the combination is the useful part.** The stars keep
reading each estimate's own p-value — "is this effect different from 1?" — while the colour reads the
gap's. All four combinations occur in one small table:

| level | stars | p (effect) | p (gap) | coloured |
|---|---|---|---|---|
| Black | \*\*\* | 0.0000 | 0.0058 | no — the gap is below the first threshold |
| Catholic | — | 0.2960 | 0.0000 | **yes** |
| Muslim | \* | 0.0570 | 0.5585 | no |
| None | \*\*\* | 0.0000 | 0.0000 | yes |

Read *starred and uncoloured* as a **robust** effect — adjustment barely touched it — and *unstarred
and coloured* as an effect that **only the crude table showed**. Catholic is the instructive row: no
effect worth reporting, but adjustment moved it a lot, and significantly.

Two more details. The direction is measured from the null, so a protective effect pulled toward 1
colours like a risky one pulled toward 1. And the test needs both estimates computed on the same
people: at `effect = "at_reference"`, or under `na = "drop_by_model"` when a compared model drops different
missing rows than the observed columns, the comparison is not made at all. (By default it always is —
see just below.)

## Grouped-binomial outcomes (a summed score)

When the outcome is a **summed score** — how many of several yes/no items a respondent answered one way — you model the number of "successes" out of a fixed number of items with `trials =`. `tab_reg()` then fits `cbind(score, trials - score)` as a binomial, so the odds ratios read on the *per-item* probability.

This is the natural model for a **multiple-answer survey question**. We take the `tea` data and its six "where do you drink tea?" items, summed into one score by `score_from_lv1()` — see *Multiple-answer questions* in `vignette("tabxplor")` for that preparation:

```{r}
data("tea", package = "FactoMineR")
tea_where_vars <- c("home", "work", "tearoom", "friends", "resto", "pub")

tea <- tea |>
  tibble::as_tibble() |>
  dplyr::mutate(dplyr::across(   # the "yes" answer must be each factor's first level
    dplyr::all_of(c(tea_where_vars, "Sport")),
    ~ (if (grepl("^Not", levels(.)[1])) forcats::fct_rev(.) else .) |>
      forcats::fct_relabel(~ gsub(".", " ", ., fixed = TRUE))
  )) |>
  score_from_lv1("tea_where", vars_list = tea_where_vars)
```

```{r}
tab_reg(tea, "tea_where", c("sex", "SPC", "Sport"),
        family = "binomial", trials = length(tea_where_vars))
```

Each odds ratio now reads *for any one place*: `1/1.44` for men means that, for any of the six places, men are about 1.44 times less likely than women to drink tea there. The model treats the six items as interchangeable trials, so one odds ratio covers them all — which is exactly the assumption to keep in mind before using it.

Grouped-binomial (like Poisson) models report a Pearson **dispersion** check in the footer, flagging over-dispersed counts.

## Ordinal and nominal outcomes (a factor with 3+ levels)

An **ordered** factor outcome is fit as a proportional-odds (cumulative) logistic model:

```{r}
tab_reg(gss_simple, "rincome", c("race", "age", "relig"))
```

A nominal outcome with three or more unordered levels is fit as one **multinomial** logistic model, giving one odds-ratio column per outcome category versus the reference category (also called relative risks ratios) :

```{r}
tab_reg(gss_simple, "party3", c("race", "age", "rincome", "relig"))
```

Relative risk ratios may be quite difficult to read because they are relative to **two** reference levels : not only the reference chosen for the predictor, but also the reference level chosen for the dependent variable. It’s specially difficult when it’s difficult to find a good reference level corresponding to the most common situation (like "married" for matrimonial status).

Most of the time, average marginal effects (AME) are easier to interpret, because they make the second reference level disappear, and directly modelise, for each level of the dependent variable, difference of percentages of each predictor compared to their reference level (a less abstract quantity than odds ratios). 

```{r}
tab_reg(gss_simple, "party3", c("race", "age", "rincome", "relig"), effect = "marginal", empirical = TRUE) # |> tab_export()
```

With `empirical = TRUE`, the crude/observed/unajusted differences and raw percentages appear in html tooltips, available at mouse hover of a cell.


## Every outcome now has an observed counterpart

`empirical = TRUE` answers one question — **"how much of this association survives once I adjust for the rest?"** — and until recently it went quiet on exactly the outcomes where that question bites hardest. It now works for every kind of outcome, and the rule behind it is a single sentence:

> The observed effect is **the model's own effect, fitted with one predictor**.

That is worth saying plainly, because it is what makes the comparison fair. The crude number is not "a percentage" and the modelled one "a coefficient" — they are the *same quantity*, computed the same way, on the *same people* (the same complete cases), and shown on the same scale. Only the list of predictors differs. So the distance between them measures adjustment, and nothing else.

The "same people" part is the default, not a hope: `na = "drop_by_outcome"` gives every model of an outcome one complete-case population, so a compared model cannot be fitted on rows the observed columns do not cover. If you ask for `na = "drop_by_model"` instead, a model on a different population gets **no** observed effect at all — better an empty cell than a "gap" that is really listwise deletion.

A word on how to read it. The colour says how much two numbers differ, and the greying whether that difference exceeds noise. It does not say "this is a confounder": the 10% first threshold is a convention, not a decision rule, and part of an odds-ratio gap is arithmetic rather than confounding (see the annex). "This effect was **attenuated** by adjustment" is the safe reading; "explained by" is not.

### An ordered outcome: what changes when you adjust

Income here is an **ordered** factor, so it is fit as a proportional-odds model. `Obs_cumOR` is the same model with one predictor at a time:

```{r}
tab_reg(gss_simple, "rincome", c("race", "relig"),
        empirical = TRUE, color = c(TRUE, "adjustment"))
```

Read a row left to right: the observed cumulative odds ratio, then the modelled one, with the **background** colouring the gap between them. Being Black is associated with lower income (`1/1.54` observed), and adjusting for religion barely moves it (`1/1.51`) — the association is not explained by religion. Compare Buddhist/Hinduist, where the effect *grows* once race and the other groups are held constant: adjustment can strengthen an association as easily as it can wash one away.

### A nominal outcome: the crude number rides in the cell

A multinomial model already spends one column per outcome category, so a second set of columns would double the table. The observed effect is printed **in the cell instead**, in parentheses:

```{r}
tab_reg(gss_simple, "party3", c("race", "relig"), family = "multinomial",
        empirical = TRUE, color = c(TRUE, "adjustment"))
```

`0.41 (0.41)` means "modelled 0.41, observed 0.41" — nothing to see. `0.08 (0.10)` means adjustment pulled the effect a little toward 1. The footer says which number is which, and hovering a cell (in an HTML export) adds the crude percentages behind it.

### A summed score

With `trials =`, the observed columns are the average share of "yes" answers expressed two ways: `Obs_mean` is the plain **mean score** — here the average number of places out of six — and `Obs_OR` the odds ratio of the summed items.

```{r}
tab_reg(tea, "tea_where", c("sex", "SPC", "Sport"),
        family = "binomial", trials = length(tea_where_vars),
        empirical = TRUE, color = c(TRUE, "adjustment"))
```

### How to read it — and how not to

Three habits are worth acquiring.

**A gap is a clue, not a verdict.** "Adjusting for religion halved this effect" is a sentence about your model, not about the world: it holds for *these* covariates, measured *this* way. A different set of controls gives a different gap.

**Do not read a gap as "the true effect minus the biased one".** Both numbers are estimates. If the crude one is based on few people it will wander, and the gap wanders with it. That is what the greying is for (below).

**Watch the scale, especially with odds ratios.** An odds ratio changes when you add a predictor *even if that predictor has nothing to do with the exposure* — a purely mathematical property called non-collapsibility. So on the odds-ratio columns above, part of every gap is arithmetic, not confounding, and the colours stay purely descriptive: they show the size of the move without claiming it is significant. If you want a gap you can test, ask for a collapsible measure — a risk ratio (`measure = "ratio"`) or a marginal effect (`effect = "marginal"`), which is also the friendlier quantity to explain:

```{r}
tab_reg(gss_simple, "party3", c("race", "relig"), family = "multinomial",
        effect = "marginal", empirical = TRUE, color = c(TRUE, "adjustment"),
        color_signif = "grey_non_signif")
```

Now the cell reads "modelled effect in percentage points, observed one in parentheses", and a background that stays grey means the two are not distinguishable from noise.

### For the record: which paths carry a test, and with which standard error

The `"adjustment"` colour always shows the **size** of the gap. Whether it also *tests* it depends on one property of the effect measure, and the table decides for you:

| what you asked for | observed effect | gap tested? |
|---|---|---|
| `effect = "marginal"` — binomial, gaussian, multinomial, ordinal | yes | **yes** |
| `effect = "marginal"` on a Poisson outcome | not paired (the crude effect is a rate ratio) | no |
| `measure = "ratio"` on a binary outcome (risk ratios) | yes | **yes** |
| Poisson incidence-rate ratios, linear β | yes | **yes** |
| odds ratios — binomial, multinomial, ordinal cumulative | yes | no (non-collapsible) |
| a compound `formula =` | no | no |
| a *weighted* 3+ level outcome with `effect = "marginal"` | yes | no (`marginaleffects` has no method) |

Where there is no test, `color_signif` is ignored for that channel and the colours read descriptively — the table never pretends to a significance it does not have.

Where there **is** one, the two estimates were fitted on the same rows, so they are correlated and the standard error of their difference is smaller than `sqrt(SE₁² + SE₂²)` would suggest. tabxplor uses the sampling variance of the difference of the two **influence functions** (Weesie 1999; Mize, Doan & Long 2019), design-based when there are survey weights. For a 3+ level outcome the modelled leg is the score of the multinomial or cumulative logit times its own variance matrix, and the observed leg is a closed form — the same one that produces the crude interval the table prints. The three `color_signif` policies then read that one interval exactly as they do everywhere else in tabxplor: `"ignore"` colours the gap regardless, `"grey_non_signif"` greys it unless the interval excludes "no change", and `"guaranteed_effect"` colours the *floor* of the gap ("adjustment moved this by at least ×1.1").

Two limits, stated rather than hidden. Each cell is tested on its own, with no multiple-comparison correction, so read the pattern rather than the one bright cell. And the test uses the robust (sandwich) variance on both sides — which is the right thing for comparing two differently-specified estimators, but means it can differ by a few percent from the descriptive interval printed beside it.


## Linear regression (a numeric outcome)

A continuous outcome gives plain linear regression coefficients (here we set `family` explicitly, because an integer like `age` is ambiguous — it could also be modelled as a count):

```{r}
tab_reg(gss_simple, "age", c("race", "marital", "relig", "rincome"), family = "gaussian")
```

In the case of a linear regression, the **empirical/observed counterpart of the model coefficient** for a categorical predictor is simply the **difference of means** : here, the difference of mean age, per level of the predictor, compared to the reference level. 
```{r}
tab_reg(gss_simple, "age", c("race", "marital", "relig", "rincome"), family = "gaussian", empirical = TRUE)
```

The empirical mean ages, and differences of mean ages from reference, can be computed in a simple table with : 

```{r}
tab(gss_simple, "race", "age", pct = "row", digits = 2, na = "drop",
    color = "difference", ref = 1,  ci_method = c(mean_diff = "student")
) |> 
  mutate(diff = set_display(age, "diff"))
# student : confidence intervals with pooled variance, to match those computed by linear regression
```


## Poisson regression (a count outcome)

```{r}
tab_reg(gss_simple, "tvhours", c("race", "marital", "relig", "rincome"), family = "poisson")
```

An **incidence-rate ratio** (IRR) of 1.5 means "50% more hours of TV per day". Unweighted Poisson models automatically use dispersion-scaled (quasi-Poisson) standard errors, so over-dispersed counts get honest, wider intervals. Concretely: with an over-dispersed outcome, `family = "poisson"` returns CIs and p-values **identical to `family = "quasipoisson"`** and emits a warning saying so (the footer reports the dispersion); at equidispersion (≈ 1) the scaling is a no-op and the result matches a plain `glm(family = poisson)` — so a comparison to a hand-fit Poisson `glm` never surprises you.

In the case of a poisson regression, the **empirical/observed counterpart of the model’s exponentiated coefficient** for a categorical predictor is the **ratio of means** : here, the ratio of average TV hours compared to the reference level.

```{r}
tab_reg(gss_simple, "tvhours", c("race", "marital", "relig", "rincome"), family = "poisson", empirical = TRUE)
```

The empirical average TV hours per day, and differences of rate of television watching from reference, can be computed in a simple table with : 

```{r}
tab(gss_simple, "race", "tvhours", pct = "row", digits = 2, na = "drop",
    color = "ratio", ref = 1,  ci_method = c(mean_ratio = "quasipoisson")
) |> 
  mutate(IRR = set_display(tvhours, "ratio"))
# default method for confidence intervals is robust ratio of means with unequal variable ;
#  we use "quasipoisson" to match those computed by quasi-poisson regression 
#  (assumption : variance is proportional to mean). 
```

## What a model column reports : `family` × `effect` × `measure`

This section is the reference behind every table above. Three arguments decide what a model column contains, and they answer three separate questions: **`family`** — which model is fitted; **`effect`** — which contrast is reported; **`measure`** — on which scale it is expressed.

Two of them are rarely typed. `family` is detected from the outcome, and `measure` falls back to the one that family is usually read on — an odds ratio for a logistic model, an incidence-rate ratio for a Poisson one, a plain coefficient for a linear one. What is left to choose is `effect`, and it is the choice worth understanding.

**One rule underpins all three, and it is what differs from `tab()`.** `effect` and `measure` change the **model that is fitted**; `display` only changes what the cell **shows**. In a cross-table every measure comes from the same counts, so asking for a ratio rather than a difference is a display choice. In a regression it is not: a risk ratio is a different fit from an odds ratio. So changing `display` never changes a number, while changing `effect` or `measure` may change every number in the column.

### The three `effect` values, in plain words

| `effect`        | the question it answers                                             | its name in the literature            |
|-----------------|---------------------------------------------------------------------|---------------------------------------|
| `"coefficient"` | two people alike on all other predictors: what does this one do?    | conditional effect (base model result)|
| `"marginal"`    | if everyone switched it, how far would the outcome move on average? | average marginal effect (AME)         |
| `"at_reference"`| the same, but for one person at the reference profile               | marginal effect at reference (MER)    |

By hand you would get them with `coef(glm(...))` (exponentiated when the measure is a ratio), with `marginaleffects::avg_comparisons(model)`, and with the same call on a one-row profile where every other predictor sits at its reference level or its mean. `"coefficient"` is the default, and the one every textbook prints.

**When do the three differ?** In a linear model, not at all: the coefficient, the average marginal effect and the effect at the reference profile are the same number (unless the predictor was fitted with a curve — see `shape =` below). In a logistic, Poisson, multinomial or ordinal model the link is curved, so the effect of a predictor depends on where you stand on the curve. A coefficient then answers "for a person like this one", and an average marginal effect answers "for this population, as it actually is" — which is why an AME is usually the friendlier number to report, and a coefficient the more comparable one across studies.

**`effect = "at_reference"`** evaluates the effect at the reference profile — the abstract individual combining the reference level of every predictor — instead of averaging it over the sample. For a multinomial model it also unlocks a contrast the other two do not have: the odds ratio of each outcome category *versus all the others* at that profile.

```{r, eval = FALSE}
tab_reg(gss_simple, "married", c("race", "age"), effect = "at_reference")
```

### The grid: what each combination builds

Every cell below is the column header the combination produces. *(default)* marks the measure the family falls back on when you say nothing.

| `family`    | `measure`                  | `"coefficient"`         | `"marginal"` | `"at_reference"`          |
|-------------|----------------------------|-------------------------|--------------|---------------------------|
| gaussian    | `"difference"` *(default)* | `Model_β`               | `Model_AME`  | `Model_MER`               |
| gaussian    | `"ratio"`                  | `Model_RoM`             | `Model_RoM`  | `Model_RoM`               |
| gaussian    | `"odds_ratio"`             | not defined             | —            | —                         |
| binomial    | `"odds_ratio"` *(default)* | `Model_OR`              | —            | —                         |
| binomial    | `"ratio"`                  | `Model_RR`              | `Model_RR`   | `Model_RR`                |
| binomial    | `"difference"`             | `Model_RD`              | `Model_AME`  | `Model_MER`               |
| poisson     | `"ratio"` *(default)*      | `Model_IRR`             | `Model_RoM`  | `Model_RoM`               |
| poisson     | `"difference"`             | —                       | `Model_AME`  | `Model_MER`               |
| poisson     | `"odds_ratio"`             | not defined             | —            | —                         |
| multinomial | `"odds_ratio"` *(default)* | `Model_OR`              | —            | `Model_OR` (vs. the rest) |
| multinomial | `"ratio"`                  | —                       | `Model_RR`   | `Model_RR`                |
| multinomial | `"difference"`             | —                       | `Model_AME`  | `Model_MER`               |
| ordinal     | `"odds_ratio"` *(default)* | `Model_OR` (cumulative) | —            | —                         |
| ordinal     | `"ratio"`                  | —                       | `Model_RR`   | `Model_RR`                |
| ordinal     | `"difference"`             | —                       | `Model_AME`  | `Model_MER`               |

The three states are the ones `reg_measures()` prints, and they are not the same refusal:

- a header — **available**: the call builds it;
- **not defined** — the quantity is not a thing for that outcome, whatever anyone implements (an odds ratio needs a probability to take the odds of, and a count has none);
- `—` — **not offered**: tabxplor does not build it, and the error message lists what this outcome does offer.

A fourth state exists only at run time: a link that does not converge on *your* data. The call says so, and for the risk difference it falls back to the linear probability model.

`family = "quasipoisson"` has exactly the grid of `poisson` — it changes the variance assumption, not what is estimated. And `measure = "log"` is not a fifth column: it is the family's default measure shown un-exponentiated, on the model's own link scale (`Model_β`).

### What the header words mean

| header      | the quantity                           | how you would get it by hand                                                 |
|-------------|----------------------------------------|------------------------------------------------------------------------------|
| `Model_β`   | a coefficient on the model's own scale | `coef(glm(...))`                                                             |
| `Model_OR`  | odds ratio                             | `exp(coef(glm(y ~ ., binomial)))`                                            |
| `Model_RR`  | risk ratio                             | modified Poisson, robust SE (Zou 2004), or a ratio of adjusted probabilities |
| `Model_IRR` | incidence-rate ratio                   | `exp(coef(glm(y ~ ., poisson)))`                                             |
| `Model_RoM` | ratio of means                         | log-link pseudo-Poisson, robust SE (Santos Silva & Tenreyro 2006)            |
| `Model_RD`  | risk difference, in points             | `glm(y ~ ., binomial("identity"))`, robust SE                                |
| `Model_AME` | average marginal effect                | `marginaleffects::avg_comparisons(model)`                                    |
| `Model_MER` | the same, at the reference profile     | the same call on a one-row reference profile                                 |

The two ratio routes are worth naming, because both are standard and neither is a `glm()` default: `measure = "ratio"` on a **binary** outcome is the modified Poisson (a Poisson likelihood on 0/1 data, made honest by a sandwich variance — Zou 2004), which is what epidemiologists use to get a risk ratio without the convergence trouble of a log-binomial fit; `measure = "ratio"` on a **numeric** outcome is the same trick one family over, a log-link Poisson pseudo-likelihood with robust standard errors (Santos Silva & Tenreyro 2006), giving a ratio of means.

### With `effect = "coefficient"`, the measure chooses the model

This is a property of the **coefficient** row alone, and it is worth being precise about. At `effect = "marginal"` or `"at_reference"`, every measure is computed from the family's own model — the logit for a binary outcome, the log link for a count — and only the last step, the averaging, differs. At `effect = "coefficient"` the measure *is* the coefficient, so asking for one the family does not report by default **changes the link, and therefore the fit**. Exactly three combinations do that; everything else on the coefficient row is the same fit read on its own scale. Each buys a directly readable number at the price of an assumption:

| `family` and `measure` | what is fitted | what the coefficient assumes |
|---|---|---|
| binomial · `"odds_ratio"` *(default)* | logistic regression | a constant odds ratio across profiles |
| binomial · `"ratio"` | modified Poisson, robust SE | a constant risk ratio |
| binomial · `"difference"` | identity-link binomial, robust SE | a constant risk difference |
| gaussian · `"difference"` *(default)* | linear regression | a constant mean difference |
| gaussian · `"ratio"` | log-link Poisson pseudo-ML, robust SE | E(y) = exp(xβ), so a constant ratio of means |
| poisson · `"ratio"` *(default)* | Poisson, quasi-Poisson SEs | a constant rate ratio |
| ordinal · `"odds_ratio"` *(default)* | proportional-odds model | one odds ratio, shared by every cut |
| multinomial · `"odds_ratio"` *(default)* | multinomial logit | a constant odds ratio per category |

What the literature says about the three that change the link:

- **The risk ratio is a well-established route, not a workaround.** `measure = "ratio"` on a binary outcome fits Zou's modified Poisson, which exists precisely to avoid the convergence failures of a log-binomial fit and is standard practice in epidemiology. Its own limits are different in kind from the risk difference's: fitted risks are not bounded above by 1, and in small or sparse samples both the estimate and the sandwich standard error are biased — penalised variants exist for that case.
- **The risk difference is the fragile one.** An identity link is bounded on neither side, so it can predict impossible risks and can simply fail to converge. When it does, tabxplor fits the linear probability model instead and the footer says so — the two target the same quantity, but they are different estimators and agree only if the model is right.
- **The ratio of means is consistent whenever the mean function is right.** Poisson pseudo-likelihood does not claim your outcome is a count: it is a device for the log link, and its robust standard error does not require the Poisson variance to be correct. Nothing, however, rescues a wrong mean function — and the outcome must be non-negative.
- **The default is not assumption-free either.** A logistic coefficient assumes a constant odds ratio, an ordinal one assumes the same odds ratio at every cut (the footer's Brant test checks it), and the odds ratio is non-collapsible, so it moves when a covariate is added even where there is nothing to confound.

**Which route to take.** If a risk ratio or a risk difference is what you want to *report*, the safer default is to leave the model on its family's own scale and ask for the measure through `effect = "marginal"` instead. Three reasons, none of them about taste: the logit always converges and can never predict a probability outside 0–100 %, while the log and identity links can do both; a marginal effect imposes no constant-effect assumption on the scale you report, since it averages the effect the fitted model actually implies at each respondent's own profile; and the marginal rows all run on the family's own model, so changing which measure you show never changes what was estimated.

That said, the two are different questions rather than two spellings of one. A coefficient is a **conditional** effect — "two people alike on every other predictor" — and a marginal effect a **population average**. Take the coefficient route when the conditional quantity is what you want: to line up with a published conditional estimate, or when "the effect is the same size everywhere on this scale" is the claim you actually mean to make.

### Caveats worth knowing

- **A binary outcome with `family = "poisson"` is not a count model.** tabxplor rewrites it to `family = "binomial", measure = "ratio"` — the modified Poisson, whose `exp(coef)` is a *risk* ratio, not an incidence-rate ratio — and says so. Naming the measure rather than the distribution is the clearer spelling.
- **`measure = "ratio"` on a numeric outcome stops on a negative outcome.** A ratio of means is not defined there; the call suggests modelling `log()` of a positive outcome, or `measure = "difference"`.
- **`Model_RD` and `Model_AME` are both percentage points, and are not the same number.** `Model_RD` is a conditional risk difference from an identity-link fit, `Model_AME` a marginal one standardised over the logistic fit (see above).
- **`measure = "log"` needs something to take the log of.** It logs the family's default measure, so it is refused where that default is already additive (a linear coefficient, an average marginal effect). Spell `"log_odds"`, `"log_risk"` or `"log_rate"` to log a chosen one instead.
- **Odds ratios are non-collapsible**, so `color = "adjustment"` colours the gap on an OR column but never tests it (see *How to read it — and how not to*, above). Ask for `measure = "ratio"` or `effect = "marginal"` if you want a gap the test can read.
- **Two combinations have no observed companion.** At `effect = "at_reference"` the model is conditional on one profile while the observed columns stay marginal over the whole sample, so the two are shown but not compared — `color = "adjustment"` and `{obs}` stay empty. And a **Poisson `effect = "marginal"`** is additive while its crude counterpart is a rate ratio: the two are not on the same scale, so they are not paired. Use `effect = "coefficient"` to compare them.
- **What needs an extra package, and what has no method.** `effect = "at_reference"` goes through **marginaleffects**; a *survey-weighted* multinomial or ordinal model has no marginal-effects method at all, so use the default `effect = "coefficient"` there. And `display` is ignored on a marginal-effects column, which already folds the adjusted prediction into the cell.

### Asking your own outcome

`reg_measures()` prints this grid for one outcome of your own data, with its per-cell status — it reads the very table `tab_reg()` resolves against, so what it prints is what the function does:

```{r}
reg_measures(gss_simple, "married")
```

The same list is in `?tab_reg`, section *"Which estimands each outcome offers"*, generated from the same source.


## Weighted and survey data

Passing a weight column with `wt =` switches estimation to a **survey design** (`survey::svyglm()`), which gives correct *design-based* standard errors. The framework `tabxplor` uses is simple: the **estimate is weighted**, and the model's standard errors come from the survey design, so unequal weights widen the confidence intervals honestly. The crude `Obs_*` companions are on the same basis, always — which is what makes them comparable with the model column beside them.

```{r, eval = FALSE}
# `weight` is a survey weight column in your data -- the common case:
tab_reg(data, "dependent", c("pred1", "pred2"), wt = "weight")
```

There are only **two ways to hand tabxplor your weights**, and that is deliberate. Either you give `wt =` a weight column, or — when your file also carries strata, clusters, a finite-population correction or calibration — you build the design once with the [survey](https://CRAN.R-project.org/package=survey) package and pass **the design itself** as `data`:

```{r, eval = FALSE}
library(survey)
d <- svydesign(ids = ~psu, strata = ~stratum, weights = ~w, data = my_survey, nest = TRUE)
tab_reg(d, "dependent", c("pred1", "pred2"), empirical = TRUE)
```

Everything then follows that design, in one regime: the `Model_*` coefficients and their intervals (`survey::svyglm()`), the average marginal effects, the per-standard-deviation scaling of numeric predictors, the model-vs-observed gap test — **and the observed `Obs_*` companion columns**, whose intervals are built on the design variance of each cell. That last point matters for the comparison this vignette is about: the model column and the observed column beside it are now measured the same way, so a difference between them is about *adjustment*, not about two different notions of uncertainty.

Two honest limits. A comparison *between two cells* (an observed odds ratio, a risk difference) ignores the design covariance between them, so it lands a few percent either side of the exact answer — against the 15–25 % it was out by when those columns were computed as if the sample were simple random. And a 3+ level outcome (nominal, ordinal) has no observed *column*: its crude value is folded into the model cell as a point, with no interval to make design-based.

Without a design, the observed columns' intervals already account for the weighting exactly (the flat survey design, `ids = ~1`) — a correction for unequal weighting only, blind to clustering and calibration.

**Why a weighted `tab_reg()` is not a weighted `tab()`.** `tab_reg()` cannot choose: its observed columns have to be measured like the model column beside them, and that one is design-based by construction (`survey::svyglm()`). `tab()` can, and keeps the descriptive convention — a weighted estimate on the raw sample size — as its default. Each table's footer says which it used, and `tab(design_effect = TRUE)` is what brings the two into line. In the vocabulary of `vignette("tabxplor")`: `tab_reg()` is always at **level 2** (or level 3 under a design), while `tab()` starts at level 1.


## Comparing several models

Pass a **named list** of predictor sets instead of a vector to fit and show several models side by side. A comparison Likelihood ratio test is one more footer key: `stats = "compare_baseline"` tests each model against the first (name another with `stats = c(compare_baseline = "M2")`), `stats = "compare_sequential"` tests each against the previous one:

```{r}
tab_reg(gss_simple,
        "married",
        list("Race only"    = "race",
             "+ age"        = c("race", "age"),
             "+ party"      = c("race", "age", "party3")),
        stats = "compare_sequential")
```

## The same model within sub-populations

`tab_vars =` is the regression analogue of `tab()`'s `tab_vars`: the same model is fitted **within each level** of a grouping variable, and the per-group tables are stacked into one grouped table. It answers "does this effect hold in every subgroup?". With just one dependent variable, it uses `tab_spread()` internally to lay the groups out as side-by-side columns for an easy comparison : 

```{r}
tab_reg(gss_simple, "married", c("race", "rincome"), tab_vars = "year")
```

Reading a row across the groups is the same chore as reading a model against its observed effect, and
it has the same answer: `color = "between_groups"` colours how far each group's effect sits from the
**first** group's, on the same row. Where the model-comparison test below says "these models differ"
once for the whole model, this says *which effects* differ between groups — a per-predictor reading of
what statisticians call effect modification.

Here the groups are separate people, so the difference between two of their effects can be tested — and
it is, so `color_signif` works as everywhere else. Add `"grey_non_signif"` and a background colour then
means "this really is a different effect, not just noise":

```{r}
tab_reg(gss_simple, "married", c("race", "rincome"), tab_vars = "party3",
        color = c(TRUE, "between_groups"), color_signif = "grey_non_signif")
```

Text colour = how strong the effect is inside that group. Background = how far it sits from the first
group's effect on the same row, greyed out when the difference could be chance. The first group is the
baseline and stays blank (nothing is compared to itself); reorder the levels with
`forcats::fct_relevel()` to pick another one. Hovering a cell in the html table gives the exact numbers:
the other group's effect, the size of the gap, its confidence interval and its p-value.

The footer gained a line at the same time: **one test per predictor**, asking "does this predictor act
differently between the groups?" for all its levels at once. That is the aggregated version of the same
question, and it is the one to quote — because it is asked once, not once per cell. You can ask for it
alone, without the colours, with `stats = c("n", "interaction")`.

### Three ways to get this wrong

**A coloured cell is a difference in *that measure*, not proof that the cause differs.** Groups can
differ because the outcome is simply rarer in one of them, or more variable — and then their effects
differ on every scale, odds ratios, risk ratios and marginal effects alike. There is no scale that
escapes this, so the reading has to be careful rather than clever.

**So read the pattern, not the single cell.** If a whole column is shifted the same way, the groups
probably differ in level or in variability — that is a property of the group, not of the predictor. If
**one row** stands out while the others sit still, that is what genuine effect modification looks like.
The colours are useful precisely because they make that shape visible at a glance. Adding
`empirical = TRUE` shows each group's base rate (`Obs_%`), which usually settles the question.

**Each cell is tested on its own, with no correction for the number of tests.** In a table with seven
comparisons, about one table in five will show a spurious coloured cell. The footer line is free of
this: it asks one question per predictor.

### For the record: what exactly is tested

The two groups are disjoint samples, so the two estimates are independent and the standard error of
their difference is `sqrt(SE_A² + SE_B²)` — the textbook test for a difference between two independent
estimates (Altman & Bland 2003). Both standard errors are read back from the confidence intervals the
table already prints, so the test and the printed intervals cannot disagree. The difference is measured
on the effect's own scale — the log-ratio for an odds ratio, a risk ratio or an incidence-rate ratio,
the plain difference for a beta or a marginal effect — and compared to a normal (z) threshold, which is
mildly conservative in small samples.

The three policies read that same interval:

| `color_signif` | what the background shows |
|----------------|---------------------------|
| `"ignore"` (default in `tab()`) | the size of the gap, coloured whether or not it is significant |
| `"grey_non_signif"` (default in `tab_reg()`) | the same, greyed unless the gap's interval excludes "no difference" |
| `"guaranteed_effect"` | the **floor** of the gap: "these two effects differ by at least ×1.1" (on a difference scale, "by at least 2 points") |

Two details worth knowing. The significance **stars** in the cells keep reading each estimate's own
p-value — "is this effect different from 1?" — not the gap's; the gap's p-value is in the tooltip.
And the footer's aggregated test is a likelihood-ratio test (an F test for linear and quasi models, a
design-based Wald test when you supply survey weights — the same rule as the model comparison below) computed on
one extra pooled model, on the **coefficients**: with `effect = "marginal"` the cells show marginal effects
while the footer tests the coefficients, two related but distinct questions, and the line says so.

## Fine-tuning the display, and scaling a predictor

A few arguments change *what each cell shows*, or *how a predictor is scaled* — without changing the model itself.

**`display`** folds a second quantity into each estimate cell. `"ci"` shows the confidence interval beside every estimate (any family):

```{r}
tab_reg(gss_simple, "married", c("race", "age"), display = "ci")
```

For a logistic model you can instead fold the model-adjusted predicted probability (`"prob"`) or the average marginal effect (`"ame"`) into the odds-ratio cell (these two need the `marginaleffects` package). Note the difference: `display = "ame"` *keeps* the odds-ratio column and adds the AME beside it, whereas `effect = "marginal"` (seen above) turns the **whole** column into an AME.

**`multiplier`** chooses the **unit** a continuous predictor's effect is reported per. One unit is rarely a readable amount: a one-year change in `age` barely moves the odds, so its odds ratio sits near 1, never crosses a colour threshold, and the row reads as "no effect" — while a whole standard deviation of age multiplies the odds by about 0.66. So the default is **per one standard deviation**, and the row label says which:

```{r}
tab_reg(gss_simple, "married", c("race", "age"))
```

Pass a single value to change it for every continuous predictor, or a named vector to override chosen ones — `"sd"`, `"2sd"` (roughly bottom to top of the distribution), or a number of units. `multiplier = 1` gives the raw per-one-unit effect:

```{r}
tab_reg(gss_simple, "married", c("race", "age"), multiplier = c(age = 10))
```

Everything scales together — the estimate, its interval, the observed `Obs_*` companion and the model-versus-observed comparison — and the p-value never changes. Two things worth knowing: because the default is not 1, a continuous predictor's `Model_OR` will **not** match `exp(coef(glm(...)))` unless you ask for `multiplier = 1`; and the standard deviation is measured once, on the predictors' complete cases, so the same variable keeps the same unit across several outcomes, compared models and `split_var` groups.


## Reading the footer statistics

The footer summarises model fit; which statistics appear depends on the family:

- **N** — the number of observations used.
- **LR vs null** — a likelihood-ratio test of the whole model against the intercept-only model (is the model worth anything?).
- **McFadden R²** — a pseudo-R² for binomial and poisson models (higher = the predictors explain more; values are much smaller than a linear R²).
- **AIC / BIC** — information criteria for comparing models on the same data (lower = better; BIC penalises complexity more).
- **Overall association** — one test per predictor: *is this variable associated with the outcome at all?* A block of stars against a reference category cannot answer that, which is why a multi-level factor needs more than its cells.
- Linear models (gaussian) add the usual **R² / adjusted R² / F / σ**; count models add the **Pearson dispersion (φ)**.

For weighted models a reduced set is reported (a Wald test against the null, Nagelkerke / Cox–Snell pseudo-R², a Rao–Scott AIC), because the likelihood-based quantities do not apply to a design-based fit.

## Checking the model

The same footer carries five **model checks**, and there is nothing to learn beyond the five nouns: each names an assumption, and the parenthesis names the instrument that measured it. Three of them cost nothing and are printed by default; the two that fit a model — **Linearity** and **Proportionality** — are asked for by name, or all at once with `stats = "all"`.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), stats = c("n", "linearity", "dispersion", "influence", "collinearity"))
```

Read them in the order they print, which is the order of what each one threatens:

- **Linearity** — *is this predictor's effect really one straight line?* One row per continuous predictor. This is the one to look at first, because a wrong shape does not stay in its own row. In the model above, letting `age` curve instead of run straight moves the top income category's odds ratio by about a quarter, and flips another income level's verdict at the 5 % threshold — while nothing in the table hints at it.
- **Proportionality (Brant)** — for an ordinal outcome, *is one cumulative odds ratio enough for every cut?*
- **Dispersion (robust/model SE)** — *are the standard errors wide enough?* About 1 means the family's variance assumption holds; 1.4 means the intervals should be about 40 % wider than they are. For a count outcome, `family = "quasipoisson"` fixes exactly that, and the row then returns to about 1 (while the Pearson dispersion φ still reports the over-dispersion itself). Under a survey design it simply says how much the design mattered — the intervals are already the design's.
- **Influence (max dfbetas)** — *does one respondent carry the result?* The largest amount, in standard errors, by which dropping a single respondent moves any coefficient. With thousands of respondents this is normally reassuringly small, and that reassurance is the point. Careful: influence is not the same as being an outlier — a surprising answer from an otherwise ordinary respondent moves nothing.
- **Collinearity (max VIF)** — *can the data tell these predictors apart?* Above about 5, two predictors are measuring much the same thing and their intervals are inflated. Collinearity biases nothing, so this is the one check that is a caution rather than a problem. (It needs the `car` package.)

Two things worth knowing. A curvature test on one predictor can pick up **another** predictor's wrong shape when the two are strongly correlated — one more reason to read the collinearity row beside it. And with a large survey sample almost any diagnostic *p*-value ends up significant, which is why three of the five report a magnitude instead.

### The shape a continuous predictor is fitted with

The linearity row tells you *whether* one straight line is enough. The little curve drawn inside a continuous predictor's own row label tells you *what shape the data actually has* — ten bins of the outcome against the predictor, on the model's own scale, with no model in it:

```{r}
tab_reg(gss_simple, "married", c("race", "age"), family = "binomial")
```

`age` climbs steeply and then flattens: one slope is a poor summary of that. `shape =` is how you fix it, without leaving the table. The most readable answer is usually to cut the predictor into groups — it then becomes an ordinary factor, so it gets one odds ratio per group, its own observed companion, counts and colours, and the shape becomes visible in the printed numbers themselves:

```{r}
tab_reg(gss_simple, "married", c("race", "age"), family = "binomial",
        shape = c(age = "quintiles"), empirical = TRUE)
```

The parsimonious alternative keeps one variable and adds a curvature term, so `age` takes two rows: the slope at the mean, and `age²`, which says whether the slope flattens (below 1) or accelerates (above 1) as you move away from it.

```{r}
tab_reg(gss_simple, "married", c("race", "age"), family = "binomial",
        shape = c(age = "quadratic"))
```

`shape = c(x = "log")` and `"sqrt"` are the other two — diminishing returns, the shape income data usually has. Everything else keeps working: the observed `Obs_*` companion is fitted with the same shape, so the model-versus-observed comparison still compares like with like, and the linearity row disappears for a predictor you have already cured.

(A `poly()` or spline basis is deliberately never emitted: the marginal-effects engine silently returns zero for those. If you reach one through a `formula`, a warning says so.)

## Plots

### The results: `forest_plot()`

`forest_plot()` draws the finished table — every effect with its confidence interval, its stars and its colour, one panel per model column:

```{r, eval = FALSE}
t <- tab_reg(gss_simple, "married", c("race", "rincome"), family = "binomial")
forest_plot(t)
```

It **reads the table and never re-fits anything**, so the figure cannot disagree with the numbers you printed: each point is the value in the cell, each gridline is one of your colour breaks (move them with `set_color_breaks()` and the axis moves too), and each point's colour is the cell's colour. It returns an ordinary `ggplot`, so `+ ggplot2::labs(...)` and `ggsave()` work as usual.

The three ways of reading significance become three things you can see:

| `color_signif` | in the table | in the plot |
|---|---|---|
| `"ignore"` | colour by the size of the effect | where the point sits, relative to the null line |
| `"grey_non_signif"` | grey unless the interval excludes the null | whether the whisker crosses the null line |
| `"guaranteed_effect"` | colour the interval bound nearest the null | how far the near end of the whisker is from it |

### Observed versus modelled, drawn honestly

With `empirical = TRUE`, each modelled effect carries its observed (crude) counterpart, and the plot draws it as a **hollow point with a bracket** — the margin of error *of the difference between the two*:

```{r, eval = FALSE}
t <- tab_reg(gss_simple, "married", c("race", "rincome"),
             family = "poisson", empirical = TRUE)      # risk ratios
forest_plot(t)
```

Read it as one question: **is the filled point outside the bracket?** If it is, adjustment moved the effect by more than noise — and that is exactly the test the table performs, to the last digit.

It is deliberately *not* two intervals side by side. Comparing two intervals by whether they overlap is a known mistake (Schenker & Gentleman 2001), and it is worse here, because the crude and the adjusted estimate are computed on the same people and are therefore correlated: the right interval is the one *around the difference*, which is what the bracket is. `observed = "ci"` draws the classic two-interval figure if you want it anyway.

A dotted connector with no bracket means the gap could not be tested — most often a conditional odds ratio, which moves under adjustment even with nothing to adjust for (see *Three ways to get this wrong* above).

### The model checks: `reg_check_plots()`

```{r, eval = FALSE}
reg_check_plots(t, gss_simple)                       # every panel the family allows
reg_check_plots(t, gss_simple, check = "linearity")  # just one
```

The same five nouns as the footer rows, one panel each, faceted across every model in the table. These are a **teaching** companion, not a decision tool: every verdict they illustrate is already a footer row, in every export, with no plotting package installed. They exist to show what a violation looks like. `theme = "print"` gives a greyscale version for a thesis appendix; you can also pass a plain fitted model instead of a table.

Note the opposite contracts: `forest_plot()` never re-fits (the results are in the table), `reg_check_plots()` always does (residuals are not).

## Where to go next

- `tab_reg()` is also available **without writing R code**, as a point-and-click **Regressions** analysis in the [jamovi](https://www.jamovi.org/download) module — install *tabxplor* from jamovi's module library (see `vignette("tabxplor")`).
- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `?tab_reg` for every argument (grouped by purpose), and `?tab_reg` &rarr; *Details* for the modelling choices.
- `survey::svydesign()` for complex designs (strata, clusters, calibration, finite-population corrections). Replicate-weight designs (`svrepdesign()`) are not supported.
- **Out of scope**, and best fitted with their own packages: survival / Cox models, mixed (multilevel) models, and pooling over multiply-imputed datasets.
- On comparing a crude and an adjusted coefficient, the canonical reference is Clogg, Petkova & Haritou (1995, *AJS* 100(5)), with Allison's comment in the same issue — what `color = "adjustment"` computes, generalised to GLMs, survey designs and marginal effects. For nested logit models, the KHB decomposition (Karlson, Holm & Breen 2012) separates the part of the change that is confounding from the part that is rescaling.
