---
title: "Programming with tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tabxplor)
library(dplyr)

options(cli.num_colors = 256)
set_color_style(type = "text", theme = "light")
```

```{r, echo = FALSE, include = FALSE}
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

This vignette is for users who want to **program with** `tabxplor` — write their own helpers, pull the underlying numbers out, or reshape a table cell by cell. If you only want to make and read tables, `vignette("tabxplor")` is the place to start.

Every numeric column of a `tabxplor` table is a single vector of class **`tabxplor_fmt`** ("formatted number"). It is a [`vctrs` record](https://vctrs.r-lib.org/reference/new_rcrd.html): behind the one value you see printed, each cell stores **all the data needed to compute the displayed number, its format and its colour** — counts, percentages, differences, confidence-interval bounds, and so on. Because it is a proper vector, it survives every `dplyr` verb, and you can read or rewrite any of its fields.

```{r}
gss_simple <- gss_cat_data_formatting()
tabs <- tab(gss_simple, race, marital, pct = "row", color = "diff")
```

## Getting the plain numbers out

The quickest way to recover the underlying numbers as ordinary numeric vectors is `get_num()`, which extracts whatever field is currently displayed:

```{r}
tabs %>% mutate(across(where(is_fmt), get_num))
```

To get the character strings instead (formatted, but without colours), use `format()`:

```{r, eval = FALSE}
tabs %>% mutate(across(where(is_fmt), format))
```

Individual fields are read most simply with `$` on the fmt column (see "Reading and changing fields" below):

```{r}
tabs$Married$pct
```

## The fields of a cell

A `tabxplor_fmt` cell carries **18 fields** (many are `NA` when the relevant quantity was not requested). The user-facing ones are:

| Field | Meaning |
|:------|:--------|
| `n`         | unweighted count (integer) |
| `wn`        | weighted count |
| `pct`       | percentage |
| `mean`      | mean (numeric column variables) |
| `diff`      | difference from the total / reference cell |
| `ratio`     | ratio to the reference (the "×2 rule"; relative risk) |
| `ci_inf`, `ci_sup` | confidence-interval bounds |
| `pvalue`    | per-cell significance p-value (feeds the stars) |
| `or`        | odds ratio / relative-risk ratio |
| `ctr`       | contribution to the chi-squared (`color = "contrib"`) |
| `var`       | variance (numeric columns; chi-squared variance with `pct`) |
| `tot_n`     | the cell's own base — the count its percentage is computed on |
| `digits`    | number of decimals to display (per cell) |
| `display`   | which field is shown (per cell) |
| `in_totrow`, `in_tottab`, `in_refrow` | is the cell in a total row / total table / reference row (logical) |

See the whole underlying data frame of a column with `vctrs::vec_data()`:

```{r}
vctrs::vec_data(tabs$Married)
```

## Reading and changing fields

Read a field with `$` (the friendliest way), or `vctrs::field()`:

```{r, eval = FALSE}
tabs$Married$pct
tabs %>% mutate(across(where(is_fmt), ~ .$pct))
tabs %>% mutate(across(where(is_fmt), ~ vctrs::field(., "pct")))
```

Confidence intervals are stored as their two bounds, the `ci_inf` and `ci_sup` fields:

```{r}
ci_tab <- tab(gss_simple, race, marital, pct = "row", ci = "cell")
ci_tab$Married$ci_inf
ci_tab$Married$ci_sup
```

Switch the displayed field with `set_display()`:

```{r, eval = FALSE}
tabs %>% set_display("diff")
tabs %>% mutate(across(where(is_fmt), ~ set_display(., "diff")))
```

To change a field, the easiest route is `dplyr::mutate()` **on the fmt vector itself** — a `tabxplor_fmt` behaves like a little data frame of its fields. For example, round every cell to two decimals:

```{r, eval = FALSE}
tabs %>% mutate(across(where(is_fmt), ~ mutate(., digits = 2L)))
```

A fuller example: turn a means table's variance into a standard deviation and add it as a new, un-coloured column shown to one decimal:

```{r, eval = FALSE}
tab_num(gss_simple, race, c(age, tvhours), digits = 1L) %>%
  mutate(across(
    c(age, tvhours),
    ~ mutate(., var = sqrt(var), display = "var", digits = 1L) %>% set_color("no"),
    .names = "{.col}_sd"
  ))
```

## Total rows, reference rows and columns

Helper predicates let you act on structural parts of a table: `is_totrow()` / `is_tottab()` / `is_refrow()` (cell-level, logical vectors), and `is_totcol()` / `is_refcol()` (column-level). They are what keeps totals in place when you re-order or when you write conditional formatting:

```{r, eval = FALSE}
# more decimals on the total row than on the body:
tab(gss_simple, race, marital, race, pct = "row") %>%
  mutate(across(
    where(is_fmt),
    ~ if_else(is_totrow(.), mutate(., digits = 1L), mutate(., digits = 2L))
  ))
```

## Column attributes

Besides its per-cell fields, each `fmt` column carries a few **column-level attributes**, read and set with `get_*` / `set_*` (or `is_*` / `as_*` for the logical ones):

- `type` — `get_type()` / `set_type()`: `"n"`, `"mean"`, `"row"`, `"col"`, `"all"`, `"all_tabs"` (or `"coef"` for regression betas). It drives which calculations `tab_*` functions perform.
- `color` — `get_color()` / `set_color()`: the colour measure of the column (`""`, `"no"`, `"diff"`, `"ratio"`, `"contrib"`, `"OR"`).
- `col_var` — `get_col_var()`: the name of the column variable (a table can hold several).
- `comp_all` — `get_comp_all()`: with `tab_vars`, is the comparison reference the sub-table (`FALSE`) or the whole table (`TRUE`)?
- `totcol` / `refcol` — `is_totcol()` / `is_refcol()`: is this a total column / a reference column?

## Building cells from scratch

`fmt()` constructs an `fmt` vector directly from numbers — handy for tests, or to add a hand-computed column. Everything the display and colour engine needs can be supplied:

```{r}
fmt(n = c(10L, 20L, 30L), pct = c(0.1, 0.2, 0.7), display = "pct", digits = 0L)
```

## Step by step, under the hood

`tab()` is a wrapper around `tab_many()`, which itself chains a small pipeline you can call yourself for maximum control: `tab_prepare()` (data cleaning) → `tab_plain()` / `tab_num()` (the count / mean core) → `tab_pct()` → `tab_ci()` → `tab_chi2()`. These building blocks are exported (the percentage / interval / test steps are now thin wrappers over one internal aggregate core), so you can assemble a bespoke table or a new helper function on top of them.

## See also

- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `vignette("tabxplor-reg")` — regression tables.
- `?fmt`, `?tab_num`, `?tab_ci` for the reference documentation of the type and its building blocks.
