---
title: "Programming with tabxplor"
description: >
  Programming with the fmt cell — know what a table is, reach any field or attribute, and write your own pipelines.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Messages and warnings are off for every chunk: the teaching notes tabxplor prints (an
# auto-detected family, an over-dispersion caveat) are explained in the prose where they
# matter, and repeated under every table they only clutter it. Re-enable one with
# `message = TRUE` on the chunk that needs it.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
# The shape table a continuous predictor draws under the footer is not this vignette's subject.
options(tabxplor.shape_table = "no")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option
library(dplyr)

# Tables render as tabxplor's real html tables (the recommended everyday setting); the shared
# stylesheet is emitted once by tab_css() below, and the hover tooltips are kept off here.
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

# Console outputs (vectors, fields...) keep their terminal colors, turned to html by fansi.
options(cli.num_colors = 256)
set_color_palette(theme = "light")
```

```{r, echo = FALSE, results = "asis"}
# The website carries a light/dark switch and tab_css("auto") follows it; a shipped vignette
# is always read on a light page, so there it stays light.
cat(tab_css(theme = if (Sys.getenv("IN_PKGDOWN") == "true") "auto" else "light"))
```

```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
# Escape the three HTML specials before fansi turns the ANSI codes into markup.
esc_html <- function(x) gsub(">", "&gt;", gsub("<", "&lt;", gsub("&", "&amp;", x, fixed = TRUE),
                                               fixed = TRUE), fixed = TRUE)
# fansi is Suggests-only, so the ANSI -> html step degrades: without it the escape codes are
# stripped and the output is handed on uncoloured, which is what a check run with no Suggests gets.
ansi_html <- if (requireNamespace("fansi", quietly = TRUE)) {
  function(x) fansi::sgr_to_html(x = esc_html(x), warn = FALSE)
} else {
  function(x) esc_html(gsub("\033\\[[0-9;]*m", "", x))
}
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         ansi_html(x),
         '</code></pre>')
})
# A cli message or warning is its own kind of condition, so knitr routes each through its own hook,
# not `output`: without these two it would land in the collapsed source block, ANSI codes and all.
for (hook in c("message", "warning")) {
  knitr::knit_hooks$set(stats::setNames(list(function(x, options) {
    paste0('<pre class="r-output"><code>',
           ansi_html(x),
           '</code></pre>')
  }), hook))
}
```

*Une version française de ce document est disponible : [Programmer avec tabxplor](https://bricenocenti.github.io/tabxplor/articles/tabxplor-programming-fr.html).*

This vignette is for users who want to **program with** `tabxplor` — write their own helpers, pull the underlying numbers out, or reshape a table cell by cell. If you only want to make and read tables, `vignette("tabxplor")` is the place to start.

Every numeric column of a `tabxplor` table is a single vector of class **`tabxplor_fmt`** ("formatted number"). It is a [`vctrs` record](https://vctrs.r-lib.org/reference/new_rcrd.html): behind the one value you see printed, each cell stores **all the data needed to compute the displayed number, its format and its colour** — counts, percentages, differences, relative risks, confidence-interval bounds, odds-ratios, and so on. Because it is a proper vector, it survives every `dplyr` verb, and you can read or rewrite any of its fields.

```{r}
gss_simple <- gss_cat_data_formatting()
tabs <- tab(gss_simple, race, marital, pct = "row", color = "difference")
```

## Getting the plain numbers out

The quickest way to recover the underlying numbers as ordinary numeric vectors is `get_num()`, which extracts whatever field is currently displayed:

```{r}
tabs |> mutate(across(where(is_fmt), get_num))
```

To get the character strings instead (formatted, but without colours), use `format()`:

```{r, eval = FALSE}
tabs |> mutate(across(where(is_fmt), format))
```

Individual fields are read most simply with `$` on the fmt column (see "Reading and changing fields" below):

```{r}
tabs$Married$pct
```

To hand the whole table to base R --- a correspondence analysis, `chisq.test()`, `mosaicplot()` --- `as.matrix()` and `as.table()` do it in one call. They keep only the data cells: the total row, the total columns and the display-time rows (the base count, `add_pct`, the p-values) go, because an analysis run on a table's own margins is wrong. Pass `totals = TRUE` to keep them.

```{r}
as.matrix(tab(gss_simple, race, marital))
```

## The fields of a cell

A `tabxplor_fmt` cell carries **21 fields** (many are `NA` when the relevant quantity was not requested). The user-facing ones are:

| Field | Meaning |
|:------|:--------|
| `n`         | unweighted count (integer) |
| `wn`        | weighted count |
| `pct`       | percentage |
| `mean`      | mean (numeric column variables) |
| `diff`      | difference from the total / reference cell |
| `ratio`     | ratio to the reference (relative risk, or a ratio of means) |
| `ci_inf`, `ci_sup` | confidence-interval bounds |
| `pvalue`    | per-cell significance p-value (feeds the stars) |
| `or`        | odds ratio / relative-risk ratio |
| `ctr`       | contribution to the chi-squared (`color = "contrib"`) |
| `var`       | variance (numeric columns; chi-squared variance with `pct`) |
| `tot_n`     | the cell's own base — the count its percentage is computed on |
| `n_eff`     | effective sample size used for the CI, with `options(tabxplor.design_effect = TRUE)` or a `survey` design |
| `obs`       | `tab_reg()` only: the observed (crude) effect the modelled one is compared to |
| `gap_se`    | `tab_reg()` only: the standard error of the gap between the estimate and `obs` |
| `digits`    | number of decimals to display (per cell) |
| `display`   | which field is shown (per cell) |
| `row_kind`  | what kind of row the cell sits in: `"data"`, `"total"`, or one of the synthetic display rows `"n"` / `"pct"` / `"pvalue"` / `"gof"` / `"blank"` |
| `in_tottab`, `in_refrow` | is the cell in a total table / a reference row (logical) |

(`row_kind` replaced a logical `in_totrow` field in 2.0.0. `x$in_totrow` still returns that logical, and `is_totrow()` / `as_totrow()` are unchanged.)

See the whole underlying data frame of a column with `vctrs::vec_data()`:

```{r}
vctrs::vec_data(tabs$Married)
```

## Reading and changing fields

Read a field with `$` (the friendliest way), or `vctrs::field()`:

```{r, eval = FALSE}
tabs$Married$pct
tabs |> mutate(across(where(is_fmt), ~ .$pct))
tabs |> mutate(across(where(is_fmt), ~ vctrs::field(., "pct")))
```

Confidence intervals are stored as their two bounds, the `ci_inf` and `ci_sup` fields:

```{r}
ci_tab <- tab(gss_simple, race, marital, pct = "row", ci = "cell")
ci_tab$Married$ci_inf
ci_tab$Married$ci_sup
```

Switch the displayed field with `set_display()`:

```{r, eval = FALSE}
tabs |> set_display("diff")
tabs |> mutate(across(where(is_fmt), ~ set_display(., "diff")))
```

To change a field, the easiest route is `dplyr::mutate()` **on the fmt vector itself** — a `tabxplor_fmt` behaves like a little data frame of its fields. For example, round every cell to two decimals:

```{r, eval = FALSE}
tabs |> mutate(across(where(is_fmt), ~ mutate(., digits = 2L)))
```

A fuller example: turn a means table's variance into a standard deviation and add it as a new, un-coloured column shown to one decimal:

```{r, eval = FALSE}
tab(gss_simple, race, c(age, tvhours), digits = 1L) |>
  mutate(across(
    c(age, tvhours),
    ~ mutate(., var = sqrt(var), display = "var", digits = 1L) |> set_color("no"),
    .names = "{.col}_sd"
  ))
```

## Composite display: combining fields

`set_display()` chooses *one* field to show. To show **several fields in one cell**, give it a `{}` **template** instead of a bare field name — the same template you can pass to `tab(display =)` when you build the table. The template is plain text with `{field}` tokens; each token is replaced by that field, formatted as usual:

```{r}
tabs |> set_display("{pct} ({diff})")
```

The rules:

- **Valid tokens** are `pct`, `n`, `wn`, `mean`, `est`, `base`, `diff`, `ratio`, `ci`, `moe`, `or`, `ctr`, `var`, `sd`, `cv`, `coef`, `resid`, `obs`, `gap` — the same set as the single-field displays. Most name a stored field; a few are **derived** and so read-only — `resid` (from the p-value and the sign of `ctr`), `sd` and `cv` from `var`, `gap` from `est` and `obs`.
- Two of them are **scale-relative**, and are what makes one template work on every table: `{est}` is whatever the column estimates (a percentage, a difference, an odds ratio) and `{base}` the level it sits on (a percentage, a mean, a count). `"{est} ({base})"` reads the same on a crosstab and on a regression.
- Common layouts have **names**: `"est"`, `"est_ci"`, `"est_base"`, `"base_est"`, `"base"`, `"base_ci"`, `"base_moe"`, `"mean_sd"`, `"mean_cv"`. They are the same names in `tab()`, in `tab_reg()` and in `set_display()`.
- The first token **outside brackets** is the *primary* one: it is what `get_num()` returns, the value Excel keeps, and the only part the colours paint. A template with no token outside brackets has no primary, and renders whole as an aside.
- A token may carry **its own precision** — `"{est} ({base:1})"` — which beats the column's `digits`.
- A **bare field name** is shorthand for its own template, so `set_display("ci")` is exactly `set_display("{ci}")`.
- It is a **display overlay** for text output (the console, `tab_html()`, `tab_md()`): the stored fields are untouched, so you can switch back at any time.
- The `{ci}` field already prints its own `[…;…]` brackets, so write `"{pct} {ci}"` — **not** `"{pct} [{ci}]"`, which would double them.

## Creating a column from another field

Because switching the display recomputes nothing — every field is already stored in the cell — you can spin off a **new column that shows a different field**. A percentage table has a reference by default (the Total row), so its `diff` field is already filled; a difference twin of every percentage column is then one `mutate()`:

```{r}
tab(gss_simple, race, marital, pct = "row") |>
  mutate(across(where(is_fmt), ~ set_display(., "diff"), .names = "{.col}_diff"))
```

`.names = "{.col}_diff"` keeps the original percentage columns and adds a `<name>_diff` twin next to each. (If a `diff` column comes out empty, the source table had no reference cell — build it with `ref =` / `comp =`, or with `color = "difference"`.)

The same recipe puts an **adjusted prediction** beside a regression effect. Two extra calls make the copy purely *descriptive*: `set_color("")` drops its colour (the ladder belongs to the effect, and the same ladder twice is noise), and `set_pvalue(NA_real_)` drops its stars — the stored p-value is their only source, so erasing it is the direct way to say "this copy tests nothing":

```{r, eval = FALSE}
tab_reg(gss_simple, "married", c("race", "rincome"), family = "binomial",
        display = "est_base") |>
  mutate(Model_pct = Model_OR |> set_display("{base}") |> set_color("") |>
                       set_pvalue(NA_real_),
         .after = Model_OR)
```

## Total rows, reference rows and columns

Helper predicates let you act on structural parts of a table: `is_totrow()` / `is_tottab()` / `is_refrow()` (cell-level, logical vectors), and `is_totcol()` / `is_refcol()` (column-level). They are what keeps totals in place when you re-order or when you write conditional formatting:

```{r, eval = FALSE}
# fewer decimals on the total row than on the body:
tab(gss_simple, race, marital, race, pct = "row") |>
  mutate(across(
    where(is_fmt),
    ~ if_else(is_totrow(.), mutate(., digits = 1L), mutate(., digits = 2L))
  ))
```

## Column attributes

Besides its per-cell fields, each `fmt` column carries a few **column-level attributes**, read and set with `get_*` / `set_*` (or `is_*` / `as_*` for the logical ones):

- `scale` — `get_scale()` / `set_scale()`: **what the column estimates**. A level (`"level_n"`, `"level_pct"`, `"level_mean"`), a difference (`"points"` for percentage points, `"mean_diff"`, `"raw_diff"` for a regression coefficient in the outcome's own units), a ratio (`"pct_ratio"`, `"mean_ratio"`, `"odds_ratio"`) or a link-scale coefficient (`"log_coef"`). It says which field holds the estimate, what its null value is, and which colour ladder it is read on.
- `pct_type` — `get_pct_type()` / `set_pct_type()`: for a percentage, what it is a percentage *of* — `"row"`, `"col"`, `"all"`, `"all_tabs"`, or `"none"` for counts, means and coefficients.
- `ci_method` — `get_ci_method()`: which interval engine built this column's bounds (`"wilson"`, `"newcombe"`, `"welch"`, `"katz"`, …; `""` when the column carries no interval).
- `color` — `get_color()` / `set_color()`: the colour measure of the column — `""`, `"no"`, or one of the measure names `"difference"`, `"ratio"`, `"odds_ratio"`, `"contrib"` (plus `"adjustment"` / `"between_groups"` on a `tab_reg()` table). The discipline's acronyms (`"diff"`, `"OR"`, `"or"`, `"RR"`, `"RD"`) are permanent shorthands you may *type*, but a built table always stores — and its legend always names — the full word.
- `col_var` — `get_col_var()`: the name of the column variable (a table can hold several).
- `col_group` — `get_col_group()`: which sub-population the column's block belongs to, after a `spread_vars` / `tab_spread()` pivot or a `tab_reg(tab_vars =)` split (`""` otherwise). Together with `col_var` it identifies a column *block*: two blocks can show the same variable for two sub-populations, and exports head them on two lines.
- `comp_all` — `get_comp_all()`: with `tab_vars`, is the comparison reference the sub-table (`FALSE`) or the whole table (`TRUE`)?
- `totcol` / `refcol` — `is_totcol()` / `is_refcol()`: is this a total column / a reference column?

## Knowing what you have before you touch it

Generic code cannot assume the shape of the table it is handed. Three accessors answer that, so a function can branch on facts rather than guess from column names.

`tab_structure()` says **what the object is** — a crosstab or a regression table, merged or not, grouped or not, and which variables sit on each axis:

```{r}
t <- tab(gss_simple, race, marital, pct = "row", ci = "ref", color = "difference")
tab_structure(t)
```

`tab_columns()` is the column-level view: one row per column, with everything the exporters and the colour engine read — what it estimates, its reference, its confidence-interval method, its role:

```{r}
tab_columns(t) |> dplyr::select(column, scale, pct_type, ref, ci_method, totcol)
```

And `fmt_attr()` reads or writes **any one of those facts by name**, which is the programmatic twin of the `get_*` / `set_*` family above — useful when the attribute is itself a variable:

```{r}
fmt_attr(t$Married, "scale")
fmt_attr(t$Married, "ci_method")
```

On a regression table, `reg_measures(data, outcome)` plays the same role for the model: it lists the `effect` × `measure` combinations that outcome can be asked for, and why the others are refused.

## Building cells from scratch

`fmt()` constructs an `fmt` vector directly from numbers — handy for tests, or to add a hand-computed column. Everything the display and colour engine needs can be supplied:

```{r}
fmt(n = c(10L, 20L, 30L), pct = c(0.1, 0.2, 0.7), display = "pct", digits = 0L)
```

## Tables from pre-aggregated counts

Sometimes the data already arrives **cross-tabulated** — a counts table from a report, a `table()`, a matrix of frequencies. `tab_counts()` builds the same coloured `tabxplor` table as `tab()`, but from those counts instead of the raw records; every calculation runs on the counts, so the result is identical to what `tab()` would have produced from individual records.

The commonest input is a **tidy counts** data frame (one row per combination, the count in a column):

```{r}
counts <- dplyr::count(gss_simple, marital, race)
tab_counts(counts, marital, race, counts = n, pct = "row", color = TRUE)
# identical to tab(gss_simple, marital, race, pct = "row", color = "difference")
```

It also melts a `table()` / `xtabs()` / matrix automatically, and reads a **wide** frame (one column per column-variable level) via `cols =` / `col_name =`:

```{r, eval = FALSE}
tab_counts(table(gss_simple$marital, gss_simple$race), pct = "row", color = "difference")

wide <- tidyr::pivot_wider(counts, names_from = race, values_from = n)
tab_counts(wide, row_var = marital, cols = c(White, Black, Other),
           col_name = "race", pct = "row", color = "difference")
```

Weighting works as in `tab()`: pass the unweighted count in `counts` and the weighted count in `wt_counts` — estimates use the weighted count, confidence intervals and Chi-2 use the unweighted N. (`options(tabxplor.design_effect = TRUE)` needs the individual weights, which pre-aggregated counts no longer carry, so it does not apply here — the CIs use the unweighted N, and the table's footer says so rather than claiming a correction it does not have.) When the only figures available are non-whole numbers (percentages × a base, or weighted-only counts), CIs and the Chi-2 test are disabled with a message.

## Pivoting a grouped table into columns

A **grouped** table (built with `tab_vars`) stacks one sub-table per group. `tab_spread()` pivots a grouping variable's levels into **side-by-side columns** — ideal to compare one measure across groups. Select the measure column you want first, then spread:

```{r}
tab(gss_simple, relig, marital, year, pct = "row", totaltab = "no", tot = "row") |>
  dplyr::select(year, relig, Married) |>
  tab_spread(year)
```

`tab(..., spread_vars = year)` does the same in one call. `tab_reg(..., tab_vars =)` produces a grouped table you can spread the same way — see `vignette("tabxplor-reg")`.

## Building many tables at once

`tab()` accepts **vectors of arguments** to build several differently-shaped tables in one call (`output_list = TRUE` keeps them as a list rather than merging them). For a fully data-driven batch, `purrr::pmap()` over a small specification table is the idiom — one row per table, one column per argument:

```{r, eval = FALSE}
specs <- tibble::tribble(
  ~row_var, ~col_var,  ~pct,
  "race",   "marital", "row",
  "relig",  "party3",  "row",
)
purrr::pmap(specs, \(row_var, col_var, pct)
            tab(gss_simple, all_of(row_var), all_of(col_var), pct = pct))
```


## Advanced options

Beyond the everyday defaults (`vignette("tabxplor")` &rarr; *Global R options*), these `options()` tune the exporters and the build. HTML / `tab_html()`:

- `tabxplor.tab_kable_css` — inline the stylesheet with each `tab_html()` *and* `tab_md()` table (`TRUE`); set `FALSE` in a many-table document and call `tab_css()` once instead (this vignette does), or in a plain-Markdown notebook that wants no stylesheet at all.
- `tabxplor.tab_kable_tooltips` — the per-cell hover tooltips (`TRUE`); set `FALSE` once per document to switch them off everywhere.
- `tabxplor.tab_kable_num_font` — the number CSS font stack (everything else is `tab_css()`'s business).
- `tabxplor.kable_popover` — click popovers instead of hover tooltips.

Excel / `tab_export("xl")`:

- `tabxplor.xl_font_text` / `tabxplor.xl_font_num` / `tabxplor.xl_font_num_stars` — the label, number and starred-number fonts.
- `tabxplor.xl_ratio_cells` — how a multiplicative cell reaches Excel: `"fold"` (default, the signed fold as a number), `"raw"` (the untransformed ratio) or `"text"` (the exact display string).

Console, stats and paths:

- `tabxplor.console_bold` — embolden reference / total / coloured cells (auto-detected per editor).
- `tabxplor.stars` — the significance ladder, as one named vector: `options(tabxplor.stars = c("*" = 0.05, "**" = 0.01))`.
- `tabxplor.export_dir` — the default export directory.

Performance and integration:

- `tabxplor.parallel` — build one worker per row variable on a background pool (needs `mirai`); `tabxplor.parallel_min` sets the smallest row-variable count worth dispatching. Release the pool with `tab_parallel_stop()`.

See `?tabxplor-options` for the full list and every default.

## See also

- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `vignette("tabxplor-reg")` — regression tables.
- `vignette("tabxplor-weights")` — weighted and survey data.
- `?fmt` and `?tabxplor-vctrs` for the cell type, `?tabxplor-options` for every option, `?tab` for the arguments and the display tokens.
