---
title: "Introduction to tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabxplor)
```

```{r hiddensetup, echo=FALSE}
library(tabxplor)

# The tables below are shown exactly as they appear in the R console --- colored. tabxplor builds
# its terminal colors with cli (which reads options("cli.num_colors")); the hook just below turns
# that ANSI into the colored HTML you see here (via fansi). In a report you would instead export
# with tab_kable() (HTML) or tab_xl() (Excel) --- see the "Exporting" section.
options(cli.num_colors = 256)
set_color_palette(theme = "light") # type = "text"
```



```{r, echo = FALSE, include = FALSE}
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

`tabxplor` helps you **explore data with cross-tables**: it makes them quickly and **colors the cells so you can read a table at a glance**. Over-represented cells use shades of blue, under-represented ones turn red/orange, so patterns jump out without you squinting at every number.

Everything is a `tibble`, so the result works with the usual `dplyr` verbs, and tables export to Excel, HTML and Markdown with their colors. Heavy computations run on `data.table`, so it stays fast on large survey data.

Throughout this vignette we use `gss_simple`, a cleaned-up version of the General Social Survey (`forcats::gss_cat`) with factors levels merged and reordered.

```{r, echo = TRUE}
gss_simple <- gss_cat_data_formatting()
```

## Your first cross-tables

`tab()` needs a data frame, a **row variable** and a **column variable**. By default it shows counts:

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

Add `pct = "row"` for row percentages (or `"col"` for column percentages). A **Total** row/column and a count column (`n`) are added automatically:

```{r, echo = TRUE}
tab(gss_simple, marital, race, pct = "row")
```

When the column variable is **numeric**, `tab()` shows its **mean** in each row instead of percentages:

```{r, echo = TRUE}
tab(gss_simple, marital, age)
```

You can pass **several row and column variables at once**. 

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), c(party3, tvhours), na = "drop", pct = "row")
```

`levels = "first"` keeps only the first level of each column factor, which is handy to display many binary factors/survey questions with multiple answers, all at once, in a compact way :

```{r, echo = TRUE}
tab(gss_simple, relig, c(married, black, income25k), pct = "row", levels = "first")
```

A few other everyday arguments: `wt =` for a weight variable, `na = "drop"` to drop missing values from the base, `digits =` for the number of decimals, and `cleannames = TRUE` to strip prefixes like `"1-"` from level names. See `?tab` for the full list.

## Sub-tables

Give `tab()` a third variable as `tab_vars` and it builds **one sub-table per group** (here, one per income group). The result is *grouped*: `dplyr` operations then run within each sub-table.

```{r, echo = TRUE}
tab(gss_simple, race, party3, rincome, na = "drop", pct = "row")
```

## colors: reading helpers

This is the main purpose of `tabxplor`. `color = "diff"` colors each cell by **how far it sits from its reference** — by default the Total of its row or column. Cells clearly above the average turn **blue**, cells clearly below turn **red/orange**, the more , and a color legend is printed underneath.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff")
```

`color = TRUE` picks a sensible scheme automatically for each column type (differences for percentages, ratios for means, …), check which one in the legend:

```{r, echo = TRUE}
tab(gss_simple, race, c(marital, partyid), pct = "row", color = TRUE)
```

**Which cell is the reference?** By default each cell is compared to its Total. Two useful alternatives:

- `ref = "first"` compares each row to the **first row** — perfect for reading an evolution over time.
- with sub-tables, `comp = "all"` compares against the **overall** Total instead of each sub-table's own Total.

```{r, echo = TRUE}
tab(gss_simple, year, marital, pct = "row", color = "diff", ref = 1)
```

color thresholds and the palette are set **once for the whole session** with `set_color_breaks()` and `set_color_palette()` — not to repeat on every table.

## colors that respect significance

The colors above show the *size* of a deviation, but not whether it is **statistically reliable**. On small samples a big-looking difference can be pure noise. The `color_signif` argument brings significance into the coloring:

- `"ignore"` (default): color every deviation by its observed size. Grey out small differences below a certain threshold.
- `"grey_non_signif"`: color by size of the effect, grey out small effects below a certain threshold, but also **grey out cells with important effects that are not significant**. Every colored cell is then guaranteed to be significantly different from its reference, without being bothered by very small significant differences. 
- `"guaranteed_effect"`: color only by the part of the effect you can be confident about (its confidence bound), with dimmer, conservative colors. Use it on **small samples** to **highlight all the differences you have the right to interpret**. Everything colored is significant ; nothing grey is. 

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff", color_signif = "grey_non_signif")
```

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = "diff", color_signif = "guaranteed_effect")
```

## Confidence intervals, tests and contributions

Print confidence intervals for each cell with `ci = "cell"` : 

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", ci = "cell") # by default, conf_level = 0.95
```

Print the confidence intervals of the difference with a reference, used to calculate significance (if 0 belongs to the confidence interval, the cell is not significantly different from the reference, here the Total row) : 
```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = TRUE, color_signif = "guaranteed_effect", display = "{ci}")
```

Add significance stars with `stars = TRUE`. They tell the same story than confidence intervals of the difference with the reference, but for different confidence levels (99%, 95%, 90%) :

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", display = "{ci}", stars = TRUE) 
```

`test = TRUE` adds a statistical test of independence per (sub-)table — **Chi-squared** for factor columns, **Welch's F ANOVA** for numeric variables:

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, tvhours), pct = "row", test = TRUE)
```

`color = "contrib"` colors cells by their **contribution to the Chi-squared** — the cells that would stand out in a correspondence analysis (here, being Black is statistically associated with *Never married* and *Separated*):

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib")
# tab(gss_simple, race, party3, pct="row", color = "contrib") # works with pct, but independent from rows/columns
```

See below for the detail of how confidence intervals and colors can be composed 


## Exporting

A finished table exports with its colors to Excel, HTML or Markdown:

```{r, echo = TRUE, eval = FALSE}
t <- tab(gss_simple, race, party3, pct = "row", color = "diff")
tab_xl(t, path = "table.xlsx")   # Excel, with colors
tab_kable(t)                     # HTML (for RStudio Viewer / R Markdown / Quarto)
tab_md(t)                        # Markdown (with pandoc color spans)
```

## Working with the result

`tab()` returns a `tibble` (of class `tabxplor_tab`), so `dplyr` verbs just work. Use the helper `is_totrow()` to keep the Total row in place when you re-order (it flags total rows, so sorting on it first sends them to the bottom):

```{r, echo = TRUE, message = FALSE}
library(dplyr)
tab(gss_simple, race, marital, pct = "row") |>
  arrange(desc(Married))
```


## Confidence intervals and colors composition : variable_type × `color` × `color_signif`

Below are, <to finish>

The type is determined by the column variable, which is either a factor (proportions/percentages are calculated) or a numeric (means are calculated). (The row variable is always coerced to a factor.)

The most important confidence intervals compare a cell to its reference, and are uses for colors : 

| type | color     | measure formula      | confidence interval: default | ci option 2 | ci option 3 |
|------|-----------|----------------------|------------------------------|-------------|-------------|
| pct  | `diff`    | $%_{cell} - %_{ref}$ | Newcombe                     | Agresti-Caffo |             |
| pct  | `ratio`   |                      |                              |             |             |
| pct  | `or`      |                      | *only `tab_reg()`*           |             |             |
| pct  | `contrib` |                      |                              |             |             |
|------|-----------|----------------------|------------------------------|-------------|-------------|
| mean | `diff`    |                      |                              |             |             |
| mean | `ratio`   |                      |                              |             |             |

The interval method can be tuned with `method_diff` / `method_ratio` / `method_mean_diff` / `method_mean_ratio` (see `?tab_ci`). Significance uses the confidence level `conf_level` (default 95%). 

And the equivalent for simple, cell-by-cell confidence intervals : 

| type | color     | confidence interval: default | ci option 2               |
|------|-----------|------------------------------|---------------------------|
| pct  | `cell`    | Wilson score interval        | Wald normal approximation |
| mean | `cell`    |                              |                           |

`method_cell` 



| type | color     | `color_signif="ignore"`    | `="grey_non_signif"`                                 | `="guaranteed_effect"`     |
|------|-----------|----------------------------|------------------------------------------------------|----------------------------|
| pct  | `diff`    | observed diff, all cells   | observed diff, grey where $diff-CI ∋ 0$              | diff CI-floor              |
| pct  | `ratio`   | observed ratio, all cells  | observed ratio, grey where $cell=ref$ not rejected   | ratio CI-floor             |
| pct  | `or`      | observed odds-ratio, all   | observed OR, grey where $OR-CI ∋ 1$                  | OR CI-floor                |
| pct  | `contrib` | χ² contribution share, all | contribution, grey where $\|resid\| < z(conf_level)$ | significant residuals only |
|------|-----------|----------------------------|------------------------------------------------------|----------------------------|
| mean | `diff`    | observed diff, all cells   | observed diff, grey where $diff-CI ∋ 0$              | diff CI-floor              |
| mean | `ratio`   | observed ratio, all cells  | observed ratio, grey where $cell=ref$ not rejected   | ratio CI-floor             |

Examples : 

```{r, eval=FALSE}
# type : pct / proportion
tab(gss_simple, race, party3, pct = "row", color = "diff" , color_signif = "grey_non_signif")
tab(gss_simple, race, party3, pct = "row", color = "ratio", color_signif = "guaranteed_effect")
tab(gss_simple, rincome, married, pct = "row", color = "OR", OR = TRUE, ref2 = 1)
tab(gss_simple, rincome, party3 , color = "contrib") # pct = "row" and pct = "col" works too

# type : mean
tab(gss_simple, rincome, tvhours, color = "diff" , color_signif = "guaranteed_effect")
tab(gss_simple, rincome, tvhours, color = "diff" , color_signif = "grey_non_signif", 
   color_breaks = list(mean_diff = c(0.4, 0.8, 1.6)), ref = 1
   ) # with a custom scale for differences
tab(gss_simple, rincome, tvhours, color = "ratio", color_signif = "grey_non_signif")
```

## Where to go next

- `vignette("tabxplor-reg")` — regression tables with `tab_reg()`, and comparing modelled to observed effects.
- `vignette("tabxplor-programming")` — the `tabxplor_fmt` cell type and how to program with its fields.
- `?tab` for every argument (grouped by purpose), `?tab_ci` for confidence-interval methods, and `?tabxplor-options` for the package-wide defaults.
