---
title: "Introduction to tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabxplor)

# The tables below are shown exactly as they appear in the R console --- coloured. tabxplor builds
# its terminal colours with cli (which reads options("cli.num_colors")); the hook just below turns
# that ANSI into the coloured HTML you see here (via fansi). In a report you would instead export
# with tab_kable() (HTML) or tab_xl() (Excel) --- see the "Exporting" section.
options(cli.num_colors = 256)
set_color_style(type = "text", theme = "light")
```

```{r, echo = FALSE, include = FALSE}
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

`tabxplor` helps you **explore data with cross-tables**: it makes them quickly, and — its main idea — it **colours the cells so you can read a table at a glance**. Over-represented cells turn blue/green, under-represented ones turn red/orange, so patterns jump out without you squinting at every number.

Everything is a `tibble`, so the result works with the usual `dplyr` verbs, and tables export to Excel, HTML and Markdown *with their colours*. Heavy computations run on `data.table`, so it stays fast on large survey data.

Throughout this vignette we use `gss_simple`, a cleaned-up version of the General Social Survey (`forcats::gss_cat`) shipped with the package. Its factors have merged, tidy levels, and their **first level is chosen to be a sensible reference** (which matters for colours and for `tab_reg()`).

```{r, echo = TRUE}
gss_simple <- gss_cat_data_formatting()
```

## Your first cross-tables

`tab()` needs a data frame, a **row variable** and a **column variable**. By default it shows counts:

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

Add `pct = "row"` for row percentages (or `"col"` for column percentages). A **Total** row/column and a count column (`n`) are added automatically:

```{r, echo = TRUE}
tab(gss_simple, marital, race, pct = "row")
```

When the column variable is **numeric**, `tab()` shows its **mean** in each row instead of percentages:

```{r, echo = TRUE}
tab(gss_simple, marital, age)
```

You can pass **several column variables at once** — factors and numerics mixed. `levels = "first"` keeps only the first level of each factor, which is handy for compact summary tables:

```{r, echo = TRUE}
tab(gss_simple, race, c(marital, age, tvhours), pct = "row", levels = "first")
```

A few other everyday arguments: `wt =` for a weight variable, `na = "drop"` to drop missing values from the base, `digits =` for the number of decimals, and `cleannames = TRUE` to strip prefixes like `"1-"` from level names. See `?tab` for the full list, grouped by purpose.

## Sub-tables

Give `tab()` a third variable as `tab_vars` and it builds **one sub-table per group** (here, one per race). The result is *grouped*: `dplyr` operations then run within each sub-table.

```{r, echo = TRUE}
tab(gss_simple, marital, partyid, race, pct = "row")
```

## Colours: reading helpers

This is what `tabxplor` is really for. `color = "diff"` colours each cell by **how far it sits from its reference** — by default the Total of its row or column. Cells clearly above the average turn **blue**, cells clearly below turn **red/orange**, and a colour legend is printed underneath.

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", color = "diff")
```

`color = TRUE` picks a sensible scheme automatically for each column type (differences for percentages, ratios for means, …), so you rarely need to think about it:

```{r, echo = TRUE}
tab(gss_simple, race, c(marital, partyid), pct = "row", color = TRUE)
```

**Which cell is the reference?** By default each cell is compared to its Total. Two useful alternatives:

- `ref = "first"` compares each row to the **first row** — perfect for reading an evolution over time.
- with sub-tables, `comp = "all"` compares against the **overall** Total instead of each sub-table's own Total.

```{r, echo = TRUE}
gss_year <- dplyr::mutate(gss_simple, year = as.factor(year))
tab(gss_year, year, marital, pct = "row", color = "diff", ref = "first")
```

Colour thresholds and the palette are set **once for the whole session** with `set_color_breaks()` and `set_color_palette()` — you do not repeat them on every table.

## Colours that respect significance

The colours above show the *size* of a deviation, but not whether it is **statistically reliable**. On small samples a big-looking difference can be pure noise. The `color_signif` argument brings significance into the colouring:

- `"ignore"` (default): colour every deviation by its observed size.
- `"grey_non_signif"`: colour by size, but **grey out cells that are not significant**. A coloured cell is then guaranteed to be significantly different from its reference.
- `"guaranteed_effect"`: colour only by the part of the effect you can be confident about (its confidence bound), with dimmer, conservative colours. Use it on **small samples** to highlight only the differences you have the right to interpret.

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", color = "diff",
    color_signif = "grey_non_signif")
```

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", color = "diff",
    color_signif = "guaranteed_effect")
```

## Confidence intervals, tests and contributions

Print confidence intervals for each cell with `ci = "cell"`, and add significance stars with `stars = TRUE`:

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", ci = "cell")
```

`test = TRUE` adds a statistical test of independence per (sub-)table — **Chi-squared** for factor columns, **Welch's F** for numeric ones:

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", test = TRUE)
```

`color = "contrib"` colours cells by their **contribution to the Chi-squared** — the cells that would stand out in a correspondence analysis (here, being Black is strongly associated with *Never married* and *Separated*):

```{r, echo = TRUE}
tab(gss_simple, race, marital, color = "contrib")
```

## Exporting

A finished table exports with its colours to Excel, HTML or Markdown:

```{r, echo = TRUE, eval = FALSE}
t <- tab(gss_simple, race, marital, pct = "row", color = "diff")
tab_xl(t, path = "table.xlsx")   # Excel, with colours
tab_kable(t)                     # HTML (for R Markdown / Quarto)
tab_md(t)                        # Markdown (with pandoc colour spans)
```

## Working with the result

`tab()` returns a `tibble` (of class `tabxplor_tab`), so `dplyr` verbs just work. Use the helper `is_totrow()` to keep the Total row in place when you re-order (it flags total rows, so sorting on it first sends them to the bottom):

```{r, echo = TRUE, message = FALSE}
library(dplyr)
tab(gss_simple, race, marital, pct = "row") %>%
  arrange(is_totrow(.), desc(Married))
```

## Expert reference: `color` × `color_signif`

For the record, here is what the two colour arguments combine to. `color` chooses **what** to colour; `color_signif` chooses **how significance gates it**.

| `color` measure | what it shows | works with |
|:----------------|:--------------|:-----------|
| `"diff"`  | difference from the reference (percentage points; Glass's Δ for means) | `pct`, means |
| `"ratio"` | relative risk (factors) or mean ratio (numerics) | `pct`, means |
| `"contrib"` | signed contribution to the Chi-squared (reference-free) | counts |
| `"OR"` | empirical odds ratio | `pct` |

| `color_signif` policy | effect on the colouring |
|:----------------------|:------------------------|
| `"ignore"` (default) | colour every deviation by its observed size |
| `"grey_non_signif"` | grey out non-significant cells; a coloured cell is significant |
| `"guaranteed_effect"` | colour only the confidence-bounded effect (conservative; for small samples) |

Significance uses the confidence level `conf_level` (default 95%); the interval method can be tuned with `method_cell` / `method_diff` (see `?tab_ci`).

## Where to go next

- `vignette("tabxplor-reg")` — regression tables with `tab_reg()`, and comparing modelled to observed effects.
- `vignette("tabxplor-programming")` — the `tabxplor_fmt` cell type and how to program with its fields.
- `?tab` for every argument (grouped by purpose), `?tab_ci` for confidence-interval methods, and `?tabxplor-options` for the package-wide defaults.
