---
title: "Introduction to tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option

# The tables below are tabxplor's real html tables, rendered live: the setup sets
# options(tabxplor.print = "html") --- the recommended everyday setting in RStudio/Positron ---
# so every bare tab() chunk knits as a colored html table. The shared stylesheet is emitted ONCE
# by the tab_css() chunk below (tab_kable_css = FALSE stops each table re-inlining it), and the
# hover tooltips are off document-wide (see the dedicated tooltips section, which re-enables them).
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

# The few console examples still show terminal colors: cli emits ANSI (options("cli.num_colors")),
# and the fansi hook below turns that ANSI into colored HTML.
options(cli.num_colors = 256)
set_color_palette(theme = "light") # type = "text"
```

```{r, echo = FALSE, results = "asis"}
cat(tab_css(theme = "light"))
```

The first time only, install the package : 

```{r, eval = FALSE}
install.packages("tabxplor", dependencies = TRUE)
```

At the start of each new R session, load it :

```{r, eval = FALSE}
library(tabxplor)
```


```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

`tabxplor` helps you **explore data with cross-tables, coloring the cells so you can read a table at a glance**. Over-represented cells use shades of blue, under-represented ones turn red/orange, so patterns jump out without you squinting at every number.

Everything is a `tibble`, so the result works with the usual `dplyr` verbs, and tables export to Excel, HTML and Markdown with their color helpers. Underlying heavy computations run on `data.table`

Throughout this vignette we use `gss_simple`, a cleaned-up version of the US General Social Survey (`forcats::gss_cat`) with factors levels merged and reordered.

```{r, echo = TRUE}
gss_simple <- gss_cat_data_formatting()
```

## Your first cross-tables

`tab()` needs a data frame, a **row variable** and a **column variable**. By default it shows counts:

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

This is tabxplor's **html table** — what you get in the RStudio or Positron Viewer pane with the recommended session option, used throughout this vignette:

```{r, eval = FALSE}
options(tabxplor.print = "html")
```

Without the option, the same table prints in the **console**, as a colored `tibble` — same information, lighter display:

```{r, include = FALSE}
options(tabxplor.print = "console")
```

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

```{r, include = FALSE}
options(tabxplor.print = "html")
```

Add `pct = "row"` for row percentages (or `"col"` for column percentages). A **Total** row/column and a count column (`n`) are added automatically:

```{r, echo = TRUE}
tab(gss_simple, marital, race, pct = "row")
```

When the column variable is **numeric**, `tab()` shows its **mean** in each row instead of percentages:

```{r, echo = TRUE}
tab(gss_simple, marital, age)
```

You can pass **several row and column variables at once**. 

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), c(party3, tvhours), na = "drop", pct = "row")
```

`levels = "first"` keeps only the first level of each column factor, which is handy to display many binary factors, like survey questions with multiple answers, all at once, in a compact way :

```{r, echo = TRUE}
tab(gss_simple, relig, c(married, black, income25k), pct = "row", levels = "first", na = "drop", cleannames = TRUE)
```

A few other everyday arguments: `na = "drop"` to drop missing values from the base, `digits =` for the number of decimals, and `cleannames = TRUE` to strip prefixes like `"1-"` from level names. See `?tab` for the full list.

## Weights

Most survey files come with a **weight**: a number saying how many people in the population each respondent stands for. The argument `wt =` uses it, and every percentage and mean in the table becomes an estimate of the *population* rather than of the people you happened to interview.

That is the easy half. The harder half is the **margin of error** around those percentages — the confidence intervals, the significance stars, the colours and the tests. There, tabxplor has **three weighting levels**, and a table's footer always tells you which one it is on.

### Level 1 — weighted percentages, plain margins of error (the default)

```{r, echo=TRUE, eval=FALSE}
tab(my_survey, education, job, wt = weight, pct = "row")
#> Weighted by weight; confidence intervals and tests use the unweighted sample size.
```

The percentages are weighted; the margins of error simply count your real respondents. This is the convention almost every teaching text and every point-and-click program uses, and it is why it is the default here. But when the weights are **unequal** it is optimistic: a sample where some people count for four others carries less information than its size suggests, and the intervals come out **usually a bit too narrow**.

### Level 2 — `design_effect = TRUE`

```{r, echo=TRUE, eval=FALSE}
tab(my_survey, education, job, wt = weight, pct = "row", design_effect = TRUE)
options(tabxplor.design_effect = TRUE)   # or once, for the whole session
#> Weighted by weight; confidence intervals and tests account for the weighting.
```

Now the unequal weighting is taken into account in **every** interval, star, colour threshold and whole-table test. This is not a rule of thumb: a column of weights *is* a survey design — the simplest one, with no clusters and no strata — and tabxplor computes that design's variance exactly, reproducing the reference [survey](https://CRAN.R-project.org/package=survey) package to the last digit.

Here it is on real data. We give `gss_simple` a made-up weight (never-married people are under-covered in most surveys, so they count for more), and build the same table twice:

```{r, echo = TRUE}
gss_w <- dplyr::mutate(gss_simple, w = ifelse(marital %in% "Never married", 2.5, 0.8))

tab(gss_w, race, party3, wt = w, pct = "row", ci = "cell", na = "drop")
```

```{r, echo = TRUE}
tab(gss_w, race, party3, wt = w, pct = "row", ci = "cell", na = "drop",
    design_effect = TRUE)
```

Same percentages, wider brackets, and a footer that says what changed. Behind them, the 16 292 White respondents are now worth about 11 800 — the price of the unequal weights.

If you weight at all, level 2 is the honest reading of your own numbers, and it costs nothing. It is not the default only because switching it on moves every figure in every table anyone has already produced, so it should be a deliberate, visible act rather than a silent one. One more reason to turn it on: a regression table's observed columns are *always* at level 2 (see `vignette("tabxplor-reg")`), so this is what makes a `tab()` percentage and a `tab_reg()` observed percentage on the same data directly comparable.

### Level 3 — a `survey` design

Real surveys are not drawn by picking names at random from the whole country. They are usually **stratified** (drawn separately by region, city size…), often **clustered** (a few neighbourhoods are picked first, then several households inside each), and their weights are usually **calibrated** so the sample matches known population totals for age, sex, region. Those three facts are not in the weight column; they live in extra variables. If your file has them, build a design with the [survey](https://CRAN.R-project.org/package=survey) package and pass it as `data`:

```{r, echo=TRUE, eval=FALSE}
library(survey)
d <- svydesign(ids = ~psu, strata = ~stratum, weights = ~w, data = my_survey, nest = TRUE)
tab(d, race, marital, pct = "row", color = TRUE, test = TRUE)
#> Design-based (survey): weighted estimates, intervals and tests account for the sample design.
```

Everything then follows the design: the estimates, the p-values, and every confidence interval, star and colour threshold.

| you pass | level | what the intervals and tests carry |
|---|---|---|
| `wt = w` | 1 | the weighted estimate, on the raw *n* — no design effect at all |
| `wt = w` + `design_effect = TRUE` | 2 | the unequal weighting, exactly — blind to clustering and calibration |
| a `survey` design as `data` | 3 | the full design: strata, clusters, `fpc`, calibration |

### What level 2 can and cannot do for you

Level 2 sees the weights, and nothing else. That matters, because the three things it cannot see do **not** pull in the same direction:

| | seen at level 2 | needs a design (level 3) |
|---|---|---|
| unequal weights | **yes, exactly** | |
| strata, calibration, finite population | | would **narrow** your intervals — by a few percent at most |
| **clusters** | | can make them **much wider** |

So the two corrections you are missing at level 2 are wildly unequal. Ignoring the strata and the calibration costs a few percent, and in the *safe* direction — your intervals are slightly too cautious. Ignoring **clusters** goes the other way and can be enormous: on a real clustered school survey, an average that level 2 reports with a margin of error of ±3 points really deserves ±9. On the American NHANES health survey, the distribution of ethnicity comes out **nine times** too precise, because that survey is built around exactly that variable. These do not cancel out: when clustering is present, it wins.

**Which surveys are clustered?** Not the ones drawn from a register, or run by web, telephone or mail — there level 2 is simply the right answer, however unequal the weights. But **face-to-face** household surveys are clustered by construction, because sending interviewers to scattered addresses is unaffordable. The French *Enquête Emploi*, for instance, is drawn as groups of about **twenty neighbouring dwellings**; neighbours resemble each other, so twenty interviews in one street tell you less than twenty interviews spread across the country.

**Does it matter for *your* table?** Usually much less than those numbers suggest, and there is a simple test. The cluster effect largely cancels out in a *comparison*, as long as both compared groups are found inside the same neighbourhoods. So ask one question about your row variable:

> Does it **vary within** a neighbourhood, or does it **define** one?

Sex, age, diploma, occupation, income, opinions — these vary within any neighbourhood, so a comparison between their categories is close to right at level 2. Region, urban/rural, city size, type of neighbourhood, local immigrant density — these *are* the neighbourhood, and there level 2 is as wrong as it is for a single percentage. The measurements back this up: on the same clustered file where an average is three times too precise, the *difference* between two groups is only 13 % too precise, and a regression coefficient 8 % — and differences are exactly what tabxplor's colours and stars test.

**Check what your file actually contains before reaching for level 3.** Look for a stratum, cluster or calibration variable (`strate`, `grappe`, `psu`, `nomen`…). Many released research files ship one calibrated weight and nothing else — the cluster is a small geographic area, so it is withheld for confidentiality — and on such a file a `survey` design would carry only the unequal weighting, which is exactly what level 2 already gives you. The European Social Survey is a useful middle case: it does publish its clusters and strata, but in a *separate* file you have to download and merge.

### The fine print

- **Sometimes level 2 makes an interval *narrower*.** That is correct, not a bug. If the weights happen to line up with what you are measuring, unequal selection can carry *more* information than equal selection, and the effective sample size comes out above your number of respondents. (The older approximation tabxplor used could only ever widen; that is why it was replaced by the exact computation.)
- **Degrees of freedom.** A clustered survey is usually built on few clusters — 15 to 60 is common — and level 3 refers each interval to that number, which widens it again by roughly 8 %. Level 2 has no clusters to count, so it cannot do this: a second, separate reason a face-to-face survey's true margin of error is wider than what level 2 prints.
- **When level 3 cannot be computed.** If a design's variance fails for a table, it falls back to the level-2 correction and says so in its footer, rather than claiming a design its numbers do not carry.
- **Exact for a cell, cautious for a difference.** Level 3 is exact for a single percentage; for a cell-vs-reference *difference* it cannot carry the covariance between the two cells, so the difference comes out slightly too wide rather than too narrow. It will never hand you a star the design does not support.
- **Pre-aggregated counts and cost.** Level 2 needs the individual weights, so a table built with `tab_counts()` from published counts cannot use it — such a table says so in its footer. And a level-3 table costs roughly three times a weighted one.


## Sub-tables

Give `tab()` a third variable as `tab_vars` and it builds **one sub-table per group** (here, one per income group). The result is *grouped*: `dplyr` operations then run within each sub-table.

```{r, echo = TRUE}
tab(gss_simple, race, party3, rincome, na = "drop", pct = "row")
```

When you pass several **row variables** *without* `tab_vars`, `tab()` merges the mirror tables into a single table by default. `output_list = TRUE` returns instead a **list with one table per row variable** (with `tab_vars`, the result is always a list):

```{r, echo = TRUE, eval = FALSE}
tab(gss_simple, c(married, income25k), race, pct = "row", output_list = TRUE)
```

## colors: reading helpers

One you the main purposes of `tabxplor` is to provide a full palette of color helpers for data exploration. `color = "diff"` colors each cell by **how far it sits from its reference** — by default the Total of its row or column. Cells clearly above the average turn **blue**, cells clearly below turn **red/orange** — the further a cell sits from its reference, the stronger the shade — and a color legend is printed underneath.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff")
```

`color = TRUE` picks a sensible scheme automatically for each column type (both differences and ratio for percentages, only ratios for means, …), check which one in the legend:

```{r, echo = TRUE}
tab(gss_simple, rincome, c(party3, marital), pct = "row", color = TRUE)
```

Numeric columns are colored the same way, on their **means** (here, hours of TV per day by income):

```{r, echo = TRUE}
tab(gss_simple, rincome, tvhours, color = "diff")
```

**Which cell is the reference for comparison ?** By default each cell is compared to the relevant Total (Total row for row percentages and Total column for column percentages) to highlight over-representations and under-representations. Two useful alternatives:

- `ref = 1` compares each row to the **first row** — perfect for reading an evolution over time or an ordinal factor.
- with sub-tables, `comp = "all"` compares against the **overall** Total instead of each sub-table's own Total.

```{r, echo = TRUE}
tab(gss_simple, year, marital, pct = "row", color = "diff", ref = 1)
```

```{r, echo = TRUE}
tab(gss_simple, rincome, party3, race, na = "drop", pct = "row", color = TRUE, comp="all")
```

**A different reference for each variable.** `ref` is reinterpreted by `pct`. Under **row** percentages (or means) it picks a reference **row**, so a *named* vector gives each row variable its own — here `race` is read against its first row, `relig` against its Total:

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), party3, pct = "row", color = "diff",
    ref = c(race = 1, relig = "tot"), na = "drop")
```

Under **column** percentages `ref` picks a reference **column** instead, vectorised over the column variables — either named (`ref = c(party3 = "first", marital = "tot")`) or positional, one value per column variable:

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, marital), pct = "col", color = "diff",
    ref = c("first", "tot"), na = "drop")
```

Color thresholds and the palette can be customised : set them **once for the whole session** with `set_color_breaks()` and `set_color_palette()`.

## colors that respect significance

The colors above show the *size* of a deviation, but not whether it is **statistically reliable**. On small samples a big-looking difference can be pure noise. The `color_signif` argument brings significance into the coloring:

- `"ignore"` (default): color every deviation by its observed size. Grey out small differences below a certain threshold.
- `"grey_non_signif"`: color by size of the effect, grey out small effects below a certain threshold, but also **grey out cells with important effects that are not significant**. Every colored cell is then guaranteed to be significantly different from its reference, without being bothered by very small significant differences. 
- `"guaranteed_effect"`: color only by the part of the effect you can be confident about (its confidence bound), with dimmer, conservative colors. Use it on **small samples** to **highlight all the differences you have the right to interpret**. Everything colored is significant ; nothing grey is.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff", color_signif = "grey_non_signif")
```

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = "diff", color_signif = "guaranteed_effect")
```

On **small samples** a big-looking percentage can rest on a handful of respondents. `n_min =` is a purely visual filter, applied last: it hides cells whose (unweighted) base is below the threshold and drops a row entirely when its largest base falls short. Here the rarest religions drop out:

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row", n_min = 400)
```

An alternative is to keep the small rows and cols but group them all in a "Other" level : 

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row",  other_if_less_than = 400)
```



## Confidence intervals, tests and contributions

Print confidence intervals for the percentage or mean of each cell with `ci = "cell"` :

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", ci = "cell") # by default, conf_level = 0.95
```

Print the confidence intervals of the **difference** with a reference, used to calculate significance (if 0 belongs to the confidence interval, the cell is not significantly different from the reference, here the Total row) :
```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = TRUE, color_signif = "guaranteed_effect",
      display = "num_ci" # "{pct} {ci}"
  )
```

`display = "num_ci"` is a type-adaptive shorthand for this: it shows each value with whatever confidence interval the table computes — `{pct} {ci}` on percentage columns and `{mean} {ci}` on numeric columns, chosen per column — so it works for a mix of factors and numbers in one call:


Add significance stars with `stars = TRUE`. They tell the same story than confidence intervals of the difference with the reference, but for different confidence levels (99%, 95%, 90%) :

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(rincome, c(party3, tvhours), pct = "row", display = "num_ci", stars = TRUE)
```

**Show two numbers in one cell.** `display` is not just for confidence intervals: it takes a **`{}` template** that combines any cell fields. For example `display = "{pct} ({diff})"` prints each percentage followed by its difference from the reference, and `"{pct} (n={n})"` follows it with the count:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff", display = "{pct} ({diff})")
```

The first field in the template is the *primary* one — the value Excel keeps and the one the colours read. `vignette("tabxplor-programming")` explains the full grammar and lists every field you can combine.

`test = TRUE` adds a statistical test of independence per (sub-)table — **Chi-squared** for factor columns, **Welch's F ANOVA** for numeric variables (`options(tabxplor.anova = "classic")` switches to the pooled F):

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, tvhours), pct = "row", test = TRUE)
```

See below for the detail of how confidence intervals and colors can be composed.

## Which cells make the table interesting? (`color = "contrib"`)

Every colour you have seen so far compares a cell to a **reference** you chose — the Total row, the
first row. `color = "contrib"` asks a different question, and needs no reference at all: **which cells
depart from what we would expect if the two variables were unrelated?** It is the natural way to
colour a table of raw counts (a count has no reference), and it is what `color = TRUE` picks for one.

There are two useful ways to answer that question, and `color_signif` chooses between them.

### 1. "Which cells build *this* association?" — the default

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib")
# tab(gss_simple, race, party3, pct = "row", color = "contrib") # works with pct too
```

Each cell is coloured by its **share of the table's Chi-squared**, written as a multiple of the average
cell: `×1` means "this cell carries an average share", `×5` means "five times the average". The strong
cells are, literally, the ones that make the table's association what it is.

This scale is *relative to this table*: it says how the association is distributed inside it, not how
big it is. That is exactly what a **correspondence analysis** reads, and it is why this is the default
reading — but it also means you cannot compare a `×2` here with a `×2` in another table.

If you have met **log-linear models** for contingency tables (Goodman's tradition, familiar from social
mobility research), this colouring is their descriptive core: the chi-squared *is* the log-linear model
of independence, and each cell's contribution is its departure from it — so a `color = "contrib"` table
is a heatmap of the association pattern, read at a glance. For the specialist models built on top of
that — quasi-independence, RC association models, UNIDIFF for comparing fluidity across groups or
cohorts — use the [logmult](https://cran.r-project.org/package=logmult) package, which also supports
complex survey designs. (Note that "log-linear" in this sense means models of *cell counts* in a table;
it is unrelated to `tab_reg(family = "poisson")`, which models an individual outcome.)

### 2. "Which cells are notably off, on a scale I can compare?"

Add `color_signif = "guaranteed_effect"` and the colour switches to the **standardized residual** — how
many standard errors each cell sits away from what independence predicts:

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib", color_signif = "guaranteed_effect")
```

Read it with the rule you may know from SPSS: beyond **±2** a cell is notable, beyond **±3** strongly
so. Positive means over-represented, negative under-represented. Unlike the share above, ±3 means the
same thing in *every* table, so you can compare two tables — and any cell that is not significant stays
grey.

The middle option, `color_signif = "grey_non_signif"`, keeps the contribution scale of (1) but greys out
cells that are not significant.

**In one line:** use the default to see *where the association lives*, `guaranteed_effect` to see *which
cells are individually notable*.

### Showing the numbers

The residual can be printed as well as coloured, with the `display` template (like the
`"{pct} ({diff})"` example above):

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "contrib",
    display = "{pct} ({resid})")
```

Hovering an html table shows both numbers (`contrib:` and `std. residual:`) in the tooltip, whatever
you display.

**For the record** (the vocabulary, if you need to cite it): the residual is the *adjusted standardized
residual* of Haberman — SPSS's "adjusted residual", R's `chisq.test()$stdres`. Careful, several tools
call `(o − e)/√e` "standardized residual" too, but that one is **not** on a ±2 scale: its variance is
below 1, so it under-states. tabxplor always uses the adjusted one, both for the colours and for the
greying. Thresholds come from `conf_level`, and with a weight the residual uses the unweighted *n*
(or the table's effective *n*, see [Weights](#weights)) while the contribution itself stays weighted. Cells whose
expected count is below 1 are left uncoloured — no approximation holds there. And, as everywhere in
tabxplor, the test is per cell: in a 30-cell table, expect one or two false positives at 5 %
(±3 is roughly a Bonferroni correction at that size).

## Hover tooltips (html tables)

Every html table carries per-cell hover **tooltips** with the numbers behind the cell: the unweighted count, the difference from the reference, the ratio, the confidence interval... They are on by default in the Viewer and in reports — this vignette only switched them off document-wide with `options(tabxplor.tab_kable_tooltips = FALSE)`, to keep the page light. Hover the cells of the table below, where they are switched back on with `tooltips = TRUE`:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff") |>
  tab_html(tooltips = TRUE)
```

**A note on weights.** With a weight (`wt =`), every proportion or mean is weighted, but by default the sample size behind the confidence intervals and tests stays the real, **unweighted** number of observations — the footer says so. Under unequal weights that carries no design effect, so it runs a little too narrow : `options(tabxplor.design_effect = TRUE)` (see [Weights](#weights)) widens every interval by exactly the weighting's own design effect and switches the whole-table Chi2 / F tests to their design-based counterparts.

## Exporting

A finished table exports with its colors to Excel, HTML or Markdown:

```{r, echo = TRUE, eval = FALSE}
tabs <- tab(gss_simple, race, party3, pct = "row", color = "diff")
tab_export(tabs) # default : html table (RStudio Viewer, .Rmd/.qmd, etc.)
tab_export(tabs, format = "xl", path = "table") # Excel export 
tab_export(tabs, format = "md", path = "table") # flat markdown file (pipes tables)
```

Two options are worth knowing:

- `theme = "auto"` lets an HTML or Markdown export **follow the reader's light/dark mode** (it flips live). For the console, `set_color_palette(theme = "auto")` detects the editor (RStudio, Positron, etc.) and picks the matching palette — it is applied automatically when the package loads.

```{r, echo = TRUE, eval = FALSE}
tab_export(tabs, theme = "auto") # HTML that follows the reader's light/dark modes
```

- Since numeric variables can only be passed in columns, some complex layout with numeric variables in rows need to transpose the table during export using `transpose = TRUE` :

```{r, echo = TRUE}
tab(gss_simple, party3, c(race, tvhours), pct = "row") |>
  tab_html(transpose = TRUE)
```

- **One stylesheet for a whole document.** In an `.Rmd`/`.qmd` report, `tab_css()` writes the colour CSS once and every later table emits only classes, so a single `theme` — including `"auto"`, which follows the reader's light/dark mode — styles every table at once. This very vignette does exactly that (with `theme = "light"`):

```{r, echo = TRUE, eval = FALSE}
options(tabxplor.tab_kable_css = FALSE)
tab_css(theme = "auto")   # emit once, near the top of the document
```

Nothing is written inline on a cell, so any look is overridable with plain CSS afterwards (column widths, fonts…); see `?tab_css` for the role classes (`.tx-rv`, `.tx-tot`, `.tx-num`).

### Black and white, for publication

Colors are for exploring. For a journal, `theme = "print"` renders the same reading in black and white: **over-represented cells in bold**, *under-represented ones in italic*, an underline for the strongest threshold, and a grey fill for a second measure. The legend names the typography instead of the hues.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff") |>
  tab_html(theme = "print")
```

This is not merely a taste: converted to lightness, the two directions of the color palette come out as the *same* shades of grey, so a grayscale print loses over-vs-under entirely. It works in every export — `tab_html()`, `tab_md()` and `tab_xl()` (real bold/italic/underline in the spreadsheet) — and the typography is written as `<b>`/`<i>`/`<u>` markup as well as CSS, so it also survives a paste into Word.

**You rarely need to ask for it.** Every stylesheet already carries this palette in an `@media print` block, so a colored html table *prints* — or saves to PDF from your browser — publication-ready on its own. Set `options(tabxplor.print_rules = FALSE)` if your printer is a color one and the colors are the point.

## A chart of the table: `forest_plot()`

A colored table already reads at a glance. When the pattern is what matters, `forest_plot()` draws the same numbers: one point per cell, its confidence interval as a whisker, its colour as the table's colour.

```{r, echo = TRUE, eval = FALSE}
tab(gss_simple, race, party3, pct = "row", ci = "cell") |>
  forest_plot()
```

It draws **whatever the table computed** — no argument to learn: with `ci = "cell"` each cell's percentage with its interval, with `ci = "diff"` the difference from the reference with a zero line, with `OR = TRUE` the odds ratios on a log axis. `what = "level"` and `what = "effect"` override it.

Two things are worth knowing:

* **The gridlines are your colour breaks.** Change them with `set_color_breaks()` and the axis moves with the table. The colour key beside the plot is the table's own legend, with the same thresholds and the same shades.
* **It reads the table, and computes nothing.** It returns a plain `ggplot`, so `+ ggplot2::labs(...)`, `+ ggplot2::theme(...)` and `ggsave()` work; `theme = "print"` gives the greyscale version, `guide = "bands"` shades the panel behind the points with the colour scale itself (a good way to *show* a class what the colours mean).

Do not confuse it with `tab_plot()`, which makes a picture *of the table* (useful to paste a coloured table where nothing else renders). `forest_plot()` is the chart; it also draws regression tables — see `vignette("tabxplor-reg")`.

## Working with the result

`tab()` returns a `tibble` (of class `tabxplor_tab`), so `dplyr` verbs just work. Use the helper `is_totrow()` to keep the Total row in place when you re-order (it flags total rows, so sorting on it first sends them to the bottom):

```{r, echo = TRUE, message = FALSE}
library(dplyr)
tab(gss_simple, race, marital, pct = "row") |>
  arrange(desc(Married))
```

**Titling and annotating.** `subtext =` prints one or more legend lines under a table (a data source, a note). `set_caption()` gives a table a **title that survives a dplyr pipeline**, and every exporter uses it as the table title:

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", subtext = "Source: GSS, 2000-2014") |>
  set_caption("Custom title")
```


## Confidence intervals and colors composition : variable_type × `color` × `color_signif`

This section is the reference behind the two color sections above. It shows how the parameters fit together — the **variable type**, the **measure** you color (`color`), and the **significance policy** (`color_signif`).

The **type** is set by the column variable — a **factor** (percentages are computed, `pct`) or a **numeric** (means are computed, `mean`). The row variable is always turned into a factor.

Every colored table answers three questions:

- **How to measure deviation?** `color =` picks what a color *means*: `"diff"` (distance from the reference), `"ratio"` (relative risk for percentages, mean ratio for means), `"contrib"` (departure from independence — weight in the Chi-squared, or the standardized residual, see below), `"OR"` (odds ratio). `color = TRUE` picks a sensible one per column type.
- **How confident are we in this measure?** every color reads one **confidence interval** at `conf_level` (95% by default). A cell is *significant* when that interval excludes its neutral value — **0** for a difference, **1** for a ratio or odds ratio. The printed bracket, the significance stars and the greying all read that same interval, so they can never disagree.
- **How to show significance?** `color_signif` — `"ignore"`, `"grey_non_signif"` or `"guaranteed_effect"` ; `stars = TRUE` to use significance stars instead of, or stacked with, colors.

**The confidence interval used for the colors and stars** compares each cell to its reference cell (by default, the corresponding cell in the Total row or Total column):

| type | color     | what the color measures                  | confidence interval (default)      |
|------|-----------|------------------------------------------|------------------------------------|
| pct  | `diff`    | cell % - reference % (percentage points) | Newcombe hybrid-score              |
| pct  | `ratio`   | cell % / reference % (relative risk)     | Katz log-risk-ratio                |
| pct  | `OR`      | empirical odds ratio                     | Woolf log-odds-ratio               |
| pct  | `contrib` | signed χ² contribution (no reference)    | — (adjusted standardized residual) |
| mean | `diff`    | cell mean - reference mean (SD units)    | Welch *t*                          |
| mean | `ratio`   | cell mean / reference mean               | robust ratio-of-means              |

Alternative interval methods go in ONE argument, `ci_method` — a named vector, partial like `ref` or `pct`, so a kind you do not name keeps its default (see `?tab`): 

- `ci_method = c(diff = "ac")` (Agresti-Caffo) or `"wald"` for a percentage difference; 
- `ci_method = c(mean_diff = "student")` (pooled, the OLS two-group interval) for a mean difference; 
- `ci_method = c(mean_ratio = "quasipoisson")` or `"poisson"` for a mean ratio. 
- Several at once: `ci_method = c(cell = "beta", diff = "ac")`.
- The relative-risk (Katz) and odds-ratio (Woolf) intervals have no alternative, so they are not a choice.
- A mean **difference** is colored **standardized** — Glass's Δ, the difference divided by the reference's standard deviation — so the `mean_diff` color breaks are read in SD units, unless the user provides a custom break scale.
- For a factor with **3 or more levels**, the odds ratio (and its interval) compares each level to the `ref2` baseline level : it is a relative risk ratio (RRR) (the same observed quantity that is modelised by a multinomial logistic regression). The OR interval is computed only when `color_signif` or `stars` needs it.

**Simple, cell-by-cell confidence intervals** (`ci = "cell"`) compare each cell to 0 % (or a mean of 0), *not* to a reference:

| type | color  | confidence interval (default) | other method                    |
|------|--------|-------------------------------|---------------------------------|
| pct  | `cell` | Wilson score interval         | `ci_method = c(cell = "wald")`  |
| mean | `cell` | one-sample Student *t* (n-1)  | —                               |

Because they compare to 0 and not to a reference, cell intervals are purely descriptive: they carry **no significance and no stars**. `ci_method = c(cell = )` chooses `"wilson"` (default), `"wald"` or `"beta"` (Korn-Graubard) for percentages; a mean cell interval is always the one-sample Student *t*.

**`color_signif` turns that interval into a coloring policy.** 

- `"ignore"` colors **every** cell by its **observed effect** size, for example the observed difference compared to the Total row. Grey cells have an observed effect below the threshold (for example, differences of less than ±5 points of percentages). 
- `"grey_non_signif"` and `"guaranteed_effect"` both color **only significant cells**, but differ in the *intensity basis*: 
  + `grey_non_signif` colors by the **observed** effect, like "ignore", greying out small deviations, but it also **greys out any large deviation that turns to be non-significant**. Ideal for large samples.
  + `guaranteed_effect` colors by the **guaranteed** effect — the confidence-bound (CI-floor), the **smaller deviation that is assured at a given confidence level** (default 95%) — so its colors are dimmer and conservative, but **all significative differences are colored**, which is ideal for small samples.

| type | color     | `="ignore"`     | `="grey_non_signif"`                | `="guaranteed_effect"`                     |
|------|-----------|-----------------|-------------------------------------|--------------------------------------------|
| pct  | `diff`    | observed diff   | grey if the diff-CI contains 0      | diff CI-floor                              |
| pct  | `ratio`   | observed ratio  | grey if the ratio-CI contains 1     | ratio CI-floor                             |
| pct  | `OR`      | observed OR     | grey if the OR-CI contains 1        | OR CI-floor                                |
| pct  | `contrib` | χ² contribution | idem, grey if \|residual\| < z(95%) | **the residual itself**, `zscore` breaks |
| mean | `diff`    | observed diff   | grey if the diff-CI contains 0      | diff CI-floor                              |
| mean | `ratio`   | observed ratio  | grey if the ratio-CI contains 1     | ratio CI-floor                             |

`contrib` is the one measure whose **unit changes with the policy**, because a contribution has no
confidence interval to take a floor of. `ignore` and `grey_non_signif` colour the *relative*
contribution (a share of this table's χ², in multiples of the mean cell — the correspondence-analysis
reading); `guaranteed_effect` colours the **adjusted standardized residual** on the absolute
`zscore` break scale, whose default `c(1.96, 2.58, 3.89, 6)` is written
`conf_level_to_z(c(0.95, 0.99, 0.9999, 1 - 2e-9))`. Its *first* value is re-anchored to the
significance threshold at print time (so every significant cell is coloured, whatever `conf_level`
says) and the others are read as spacings from it. Both readings share one significance source:

- the **adjusted standardized residual** (Haberman; SPSS's "adjusted residual", `chisq.test()$stdres`),
  never the Pearson residual `(o − e)/√e`, whose variance is below 1 and which therefore under-rejects;
- computed the way every interval in tabxplor is: weighted estimate, and a base that follows the
  inference basis — the raw *n* (the reading a correspondence analysis expects), or that *n* divided
  by the **association's** design effect under `options(tabxplor.design_effect = TRUE)` or a `survey`
  design. That design effect is Rao-Scott's, the very one the table's own Chi-2 line reports, so the
  colours and the p-value describe one design effect and not two. One base per table, so a counts
  table and a percentage table of the same data give the **same** residuals. The contribution itself
  stays **weighted** and is identical at every basis — it estimates the population table's structure,
  while the residual is the test, which is what keeps the correspondence-analysis reading safe;
- with cells whose expected count is below 1 left uncoloured (the normal approximation fails there);
- and per cell, with no multiple-comparison correction — as everywhere else in tabxplor. In a 30-cell
  table expect one or two false positives at 5 %; the ±3 convention is roughly Bonferroni at that size.

The residual is also readable as a number: `display = "{resid}"` or `"{pct} ({resid})"`, and it always
appears in the html tooltip.

Colours are computed **per column** at print time, so they cannot see a table-level attribute — but each
column now records the confidence level it was built at, so every significance threshold inside them
(the greying, and the `zscore` anchor) follows the call's `conf_level`, like the printed intervals and
the stars. `options(tabxplor.conf_level)` remains the fallback for a column that never recorded one (a
hand-built `fmt()`).

Examples :

```{r, echo=TRUE, eval=FALSE}
# --- factors: percentages -------------------------------------------------
tab(gss_simple, race, party3, pct = "row", color = "diff",  color_signif = "grey_non_signif")
tab(gss_simple, race, party3, pct = "row", color = "ratio", color_signif = "guaranteed_effect")
tab(gss_simple, rincome, married, pct = "row", color = "OR", OR = TRUE, ref2 = 1)
tab(gss_simple, rincome, party3, color = "contrib")   # works with pct = "row"/"col" too
#    the same table read as standardized residuals (the +-2 / +-3 scale), with the numbers shown
tab(gss_simple, rincome, party3, color = "contrib", color_signif = "guaranteed_effect",
    display = "{n} ({resid})")
#    a stricter residual ladder, written in confidence levels
tab(gss_simple, rincome, party3, color = "contrib", color_signif = "guaranteed_effect",
    color_breaks = list(zscore = conf_level_to_z(c(0.99, 0.999, 0.9999))))

# --- numerics: means ------------------------------------------------------
tab(gss_simple, rincome, tvhours, color = "diff",  color_signif = "guaranteed_effect")
tab(gss_simple, rincome, tvhours, color = "ratio", color_signif = "grey_non_signif")

#    a custom scale for differences in means, and a "first row" reference
tab(gss_simple, rincome, tvhours, color = "diff", color_signif = "grey_non_signif",
    color_breaks = list(mean_diff = c(0.4, 0.8, 1.6)), ref = 1)
```

## Session options

A handful of `options()` set your preferred defaults once for the whole session — put them at the top of a script, or in your `.Rprofile`. Each one has a per-call argument too; the option just changes the default. The everyday ones:

- `options(tabxplor.print = "html")` — print tables not in console, but as html in RStudio or Positron Viewer Pane by default (recommended)
- `options(tabxplor.cleannames = TRUE)` — strip `"1-"`-style prefixes from level names everywhere.
- `options(tabxplor.parallel = 8)` — parallelise tables with multiples variables on different CPU cores by default (needs `mirai`)
- `options(tabxplor.var_labels = TRUE)` — in exports, show a variable's label (from `haven`/`labelled` data) instead of its bare name.
- `options(tabxplor.theme = "auto")` — the export theme (`"light"`/`"dark"`/`"auto"`); `set_color_palette(theme = "auto")` does the same for the console.
- `options(tabxplor.stars = TRUE)` — show significance stars in every table (like `stars = TRUE`).
- `options(tabxplor.conf_level = 0.9)` — the confidence level for intervals and tests (default `0.95`).
- `options(tabxplor.design_effect = TRUE)` — on weighted data, make every interval, star, colour threshold and test account for the unequal weighting (see [Weights](#weights)).
- `options(tabxplor.ci_print = "moe")` — print a confidence interval as a `pct ± margin of error` instead of a `[low; high]` bracket.
- `options(tabxplor.lang = "fr")` — the language of the colour legends and footers (`"auto"`/`"en"`/`"fr"`).

Colour thresholds and palettes have their own helpers, `set_color_breaks()` and `set_color_palette()`. `?tabxplor-options` documents every option, and `vignette("tabxplor-programming")` covers the more advanced ones (export fonts, parallel builds…).

## A point-and-click interface (jamovi)

Everything above is also available **without writing R code**, through a [jamovi](https://www.jamovi.org/download) module. jamovi is a free, open-source statistical software : install it, open the modules menu (the **`+`** at the top-right), choose **jamovi library**, and install *tabxplor*. It adds a **Crosstables** analysis — and a **Regressions** analysis powered by `tab_reg()` — with the same coloured, exportable tables, driven entirely by menus. Handy for teaching, or for colleagues who do not use R.

## Where to go next

- `vignette("tabxplor-reg")` — regression tables with `tab_reg()`, and comparing modelled to observed effects.
- `vignette("tabxplor-programming")` — the `tabxplor_fmt` cell type and how to program with its fields.
- `?tab` for every argument (grouped by purpose), `?tab_ci` for confidence-interval methods, and `?tabxplor-options` for the package-wide defaults.
