---
title: "Introduction to tabxplor"
description: >
  Build a first coloured cross-table, read its colours and its legend, then choose what a cell shows and how it exports.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option

# The tables below are tabxplor's real html tables, rendered live: the setup sets
# options(tabxplor.print = "html") --- the recommended everyday setting in RStudio/Positron ---
# so every bare tab() chunk knits as a colored html table. The shared stylesheet is emitted ONCE
# by the tab_css() chunk below (tab_kable_css = FALSE stops each table re-inlining it), and the
# hover tooltips are off document-wide (see the dedicated tooltips section, which re-enables them).
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

# The few console examples still show terminal colors: cli emits ANSI (options("cli.num_colors")),
# and the fansi hook below turns that ANSI into colored HTML.
options(cli.num_colors = 256)
set_color_palette(theme = "light") # type = "text"
```

```{r, echo = FALSE, results = "asis"}
# The website carries a light/dark switch and tab_css("auto") follows it; a shipped vignette
# is always read on a light page, so there it stays light.
cat(tab_css(theme = if (Sys.getenv("IN_PKGDOWN") == "true") "auto" else "light"))
```

*Une version française de ce document est disponible : [Introduction à tabxplor](https://bricenocenti.github.io/tabxplor/articles/tabxplor-fr.html).*

The first time only, install the package : 

```{r, eval = FALSE}
install.packages("tabxplor", dependencies = TRUE)
```

At the start of each new R session, load it :

```{r, eval = FALSE}
library(tabxplor)
```


```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
# Escape the three HTML specials before fansi turns the ANSI codes into markup.
esc_html <- function(x) gsub(">", "&gt;", gsub("<", "&lt;", gsub("&", "&amp;", x, fixed = TRUE),
                                               fixed = TRUE), fixed = TRUE)
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = esc_html(x), warn = FALSE),
         '</code></pre>')
})
# A cli message or warning is its own kind of condition, so knitr routes each through its own hook,
# not `output`: without these two it would land in the collapsed source block, ANSI codes and all.
for (hook in c("message", "warning")) {
  knitr::knit_hooks$set(stats::setNames(list(function(x, options) {
    paste0('<pre class="r-output"><code>',
           fansi::sgr_to_html(x = esc_html(x), warn = FALSE),
           '</code></pre>')
  }), hook))
}
```

`tabxplor` helps you **explore data with cross-tables, coloring the cells so you can read a table at a glance**. Over-represented cells use shades of blue, under-represented ones turn red/orange, so patterns jump out without you squinting at every number.

Everything is a `tibble`, so the result works with the usual `dplyr` verbs, and tables export to Excel, HTML and Markdown with their color helpers. Underlying heavy computations run on `data.table`

**No R needed, if you prefer menus.** Everything below is also available point-and-click, through a module for [jamovi](https://www.jamovi.org/download) — a free, open-source statistical software. Install it, open the modules menu (the **`+`** at the top right), choose **jamovi library**, and install *tabxplor*: it adds a **Crosstables** analysis and a **Regressions** analysis, with the same coloured, exportable tables. The options carry the same names as the arguments taught here, so this vignette reads as its manual.

Throughout this vignette we use `gss_simple`, a cleaned-up version of the US General Social Survey (`forcats::gss_cat`) with factors levels merged and reordered.

```{r, echo = TRUE}
gss_simple <- gss_cat_data_formatting()
```

## Your first cross-tables

`tab()` needs a data frame, a **row variable** and a **column variable**. By default it shows counts:

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

This is tabxplor's **html table** — what you get in the RStudio or Positron Viewer pane with the recommended session option, used throughout this vignette:

```{r, eval = FALSE}
options(tabxplor.print = "html")
```

Without the option, the same table prints in the **console**, as a colored `tibble` — same information, lighter display:

```{r, include = FALSE}
options(tabxplor.print = "console")
```

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

```{r, include = FALSE}
options(tabxplor.print = "html")
```

Add `pct = "row"` for row percentages (or `"col"` for column percentages). A **Total** row/column and a count column (`n`) are added automatically:

```{r, echo = TRUE}
tab(gss_simple, marital, race, pct = "row")
```

When the column variable is **numeric**, `tab()` shows its **mean** in each row instead of percentages:

```{r, echo = TRUE}
tab(gss_simple, marital, age)
```

You can pass **several row and column variables at once**. 

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), c(party3, tvhours), na = "drop", pct = "row")
```

A few other everyday arguments: `na = "drop"` to drop missing values from the base, `digits =` for the number of decimals, and `cleannames = TRUE` to strip prefixes like `"1-"` from level names. See `?tab` for the full list.

**If your numbers are already counted** — a published table, a `table()`, a `count()` — start from `tab_counts()` instead. It builds the same object, with the same percentages, colours and tests:

```{r, echo = TRUE}
counts <- dplyr::count(gss_simple, marital, race)          # or a published table
tab_counts(counts, marital, race, counts = n, pct = "row", color = "difference")
```

## Colours: reading a table at a glance

One of the main purposes of `tabxplor` is its palette of colour helpers for data exploration. `color = "difference"` colors each cell by **how far it sits from its reference** — by default the Total of its row or column. Cells clearly above the average turn **blue**, cells clearly below turn **red/orange** — the further a cell sits from its reference, the stronger the shade — and a color legend is printed underneath.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "difference")
```

`color = TRUE` picks a sensible scheme automatically for each column type (both differences and ratio for percentages, only ratios for means, …), check which one in the legend:

```{r, echo = TRUE}
tab(gss_simple, rincome, c(party3, marital), pct = "row", color = TRUE)
```

Numeric columns are colored the same way, on their **means** (here, hours of TV per day by income):

```{r, echo = TRUE}
tab(gss_simple, rincome, tvhours, color = "difference")
```

**Which cell is the reference for comparison ?** By default each cell is compared to the relevant Total (Total row for row percentages and Total column for column percentages) to highlight over-representations and under-representations. Two useful alternatives:

- `ref = 1` compares each row to the **first row** — perfect for reading an evolution over time or an ordinal factor.
- with sub-tables, `comp = "all"` compares against the **overall** Total instead of each sub-table's own Total.

```{r, echo = TRUE}
tab(gss_simple, year, marital, pct = "row", color = "difference", ref = 1)
```

```{r, echo = TRUE}
tab(gss_simple, rincome, party3, race, na = "drop", pct = "row", color = TRUE, comp="all")
```

**A different reference for each variable.** `ref` is reinterpreted by `pct`. Under **row** percentages (or means) it picks a reference **row**, so a *named* vector gives each row variable its own — here `race` is read against its first row, `relig` against its Total:

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), party3, pct = "row", color = "difference",
    ref = c(race = 1, relig = "tot"), na = "drop")
```

Under **column** percentages `ref` picks a reference **column** instead, vectorised over the column variables — either named (`ref = c(party3 = "first", marital = "tot")`) or positional, one value per column variable:

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, marital), pct = "col", color = "difference",
    ref = c("first", "tot"), na = "drop")
```

Color thresholds and the palette can be customised : set them **once for the whole session** with `set_color_breaks()` and `set_color_palette()`.

## Colours that respect significance

The colors above show the *size* of a deviation, but not whether it is **statistically reliable**. On small samples a big-looking difference can be pure noise. The `color_signif` argument brings significance into the coloring:

- `"ignore"` (default): color every deviation by its observed size. Grey out small differences below a certain threshold.
- `"grey_non_signif"`: color by size of the effect, grey out small effects below a certain threshold, but also **grey out cells with important effects that are not significant**. Every colored cell is then guaranteed to be significantly different from its reference, without being bothered by very small significant differences. 
- `"guaranteed_effect"`: color only by the part of the effect you can be confident about (its confidence bound), with dimmer, conservative colors. Use it on **small samples** to **highlight all the differences you have the right to interpret**. Everything colored is significant ; nothing grey is.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "difference", color_signif = "grey_non_signif")
```

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = "difference", color_signif = "guaranteed_effect")
```

On **small samples** a big-looking percentage can rest on a handful of respondents. `n_min =` is a purely visual filter, applied last: it hides cells whose (unweighted) base is below the threshold and drops a row entirely when its largest base falls short. Here the rarest religions drop out:

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row", n_min = 400)
```

An alternative is to keep the small rows and cols but group them all in a "Other" level : 

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row",  other_if_less_than = 400)
```

## Confidence intervals, tests and stars

Print confidence intervals for the percentage or mean of each cell with `ci = "cell"` :

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", ci = "cell") # by default, conf_level = 0.95
```

Print the confidence intervals of the **difference** with a reference, used to calculate significance (if 0 belongs to the confidence interval, the cell is not significantly different from the reference, here the Total row) :
```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = TRUE, color_signif = "guaranteed_effect",
      display = "base_ci" # "{base} {ci}"
  )
```

`display = "base_ci"` is the named layout for this. `{base}` means "the level this column shows" — a percentage on a factor column, a mean on a numeric one — so one call works for a mix of factors and numbers, and each column answers with its own quantity. The next section covers the rest of the layouts.


Add significance stars with `stars = TRUE`. They tell the same story than confidence intervals of the difference with the reference, but for different confidence levels (99%, 95%, 90%) :

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(rincome, c(party3, tvhours), pct = "row", display = "base_ci", stars = TRUE)
```

`test = TRUE` adds a statistical test of independence per (sub-)table — **Chi-squared** for factor columns, **Welch's F ANOVA** for numeric variables (`options(tabxplor.anova = "classic")` switches to the pooled F):

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, tvhours), pct = "row", test = TRUE)
```

## What the cell shows: the `display` grammar

Every cell already holds far more than the one number it prints — its count, its percentage, its difference from the reference, its confidence interval. `display` chooses which of them you see. **Nothing is recomputed**: a display is picked after the table is built, so changing it never changes a number.

The quickest route is a **named layout**. The everyday ones:

| `display =` | the cell shows |
|---|---|
| `"base"` | the level alone — the percentage, the mean or the count (the default) |
| `"base_ci"` | the level with its confidence interval, `48.6 [45.1; 52.1]` |
| `"base_moe"` | the level with its margin of error, `48.6 ± 3.5` |
| `"base_diff"` | the level and, in brackets, its difference from the reference |
| `"base_ratio"` | the level and the same comparison as a ratio |
| `"mean_sd"` | a mean and its standard deviation (numeric columns) |
| `"mean_cv"` | a mean and its coefficient of variation — the spread as a percentage of the level, so two columns measured in different units become comparable (the default on numeric columns) |

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, tvhours), pct = "row", display = "base_moe")
```

**Or write your own**, with a `{}` template naming the fields you want. `display = "{pct} ({diff})"` prints each percentage followed by its difference from the reference; `"{pct} (n={n})"` follows it with the count:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "difference", display = "{pct} ({diff})")
```

Two rules are worth knowing. The **first field outside brackets is the primary one**: it carries the significance stars, it is what an Excel export keeps as a real number, and it is the only part the colours paint. And a field can carry **its own precision**, `{pct:1} ({n:0})`, which beats the table's `digits` — useful when an aside would otherwise be printed to as many decimals as the estimate.

**On a table you have already built**, `set_display()` does the same thing:

```{r, echo = TRUE}
tabs <- tab(gss_simple, race, party3, pct = "row")
set_display(tabs, "{pct} (n={n})")
```

`?tab` lists every field a template can name; `vignette("tabxplor-programming")` explains the record behind them.


## Which cells make the table interesting? (`color = "contrib"`)

Every colour you have seen so far compares a cell to a **reference** you chose — the Total row, the
first row. `color = "contrib"` asks a different question, and needs no reference at all: **which cells
depart from what we would expect if the two variables were unrelated?** It is the natural way to
colour a table of raw counts (a count has no reference), and it is what `color = TRUE` picks for one.

There are two useful ways to answer that question, and `color_signif` chooses between them.

### 1. "Which cells build *this* association?" — the default

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib")
# tab(gss_simple, race, party3, pct = "row", color = "contrib") # works with pct too
```

Each cell is coloured by its **share of the table's Chi-squared**, written as a multiple of the average
cell: `×1` means "this cell carries an average share", `×5` means "five times the average". The strong
cells are, literally, the ones that make the table's association what it is.

This scale is *relative to this table*: it says how the association is distributed inside it, not how
big it is. That is exactly what a **correspondence analysis** reads, and it is why this is the default
reading — but it also means you cannot compare a `×2` here with a `×2` in another table.

If you have met **log-linear models** for contingency tables (Goodman's tradition, familiar from social
mobility research), this colouring is their descriptive core: the chi-squared *is* the log-linear model
of independence, and each cell's contribution is its departure from it — so a `color = "contrib"` table
is a heatmap of the association pattern, read at a glance. For the specialist models built on top of
that — quasi-independence, RC association models, UNIDIFF for comparing fluidity across groups or
cohorts — use the [logmult](https://cran.r-project.org/package=logmult) package, which also supports
complex survey designs. (Note that "log-linear" in this sense means models of *cell counts* in a table;
it is unrelated to `tab_reg(family = "poisson")`, which models an individual outcome.)

### 2. "Which cells are notably off, on a scale I can compare?"

Add `color_signif = "guaranteed_effect"` and the colour switches to the **standardized residual** — how
many standard errors each cell sits away from what independence predicts:

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib", color_signif = "guaranteed_effect")
```

Read it with the rule you may know from SPSS: beyond **±2** a cell is notable, beyond **±3** strongly
so. Positive means over-represented, negative under-represented. Unlike the share above, ±3 means the
same thing in *every* table, so you can compare two tables — and any cell that is not significant stays
grey.

The middle option, `color_signif = "grey_non_signif"`, keeps the contribution scale of (1) but greys out
cells that are not significant.

**In one line:** use the default to see *where the association lives*, `guaranteed_effect` to see *which
cells are individually notable*.

### Showing the numbers

The residual can be printed as well as coloured, with the `display` template (like the
`"{pct} ({diff})"` example above):

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "contrib",
    display = "{pct} ({resid})")
```

Hovering an html table shows both numbers (`ctr:` and `resid:`) in the tooltip, whatever you
display: every tooltip line is named after the field it shows, so the hover uses the same words as
the column headers and as `$`.

**If you need to cite it**: this is Haberman's *adjusted standardized residual* — SPSS's "adjusted
residual", R's `chisq.test()$stdres` — not the `(o − e)/√e` that several tools also call
"standardized", which sits below a ±2 scale and under-states. As everywhere in tabxplor the test is
per cell, so in a 30-cell table expect one or two false positives at 5 % (±3 is roughly a Bonferroni
correction at that size). See `?tab`.

## Sub-tables, and the most compact table

Give `tab()` a third variable as `tab_vars` and it builds **one sub-table per group** (here, one per income group). The result is *grouped*: `dplyr` operations then run within each sub-table.

```{r, echo = TRUE}
tab(gss_simple, race, party3, rincome, na = "drop", pct = "row")
```

When you pass several **row variables** *without* `tab_vars`, `tab()` merges the mirror tables into a single table by default. `output_list = TRUE` returns instead a **list with one table per row variable** (with `tab_vars`, the result is always a list):

```{r, echo = TRUE, eval = FALSE}
tab(gss_simple, c(married, income25k), race, pct = "row", output_list = TRUE)
```

### The most compact table: one block of columns per group

`spread_vars` shows a sub-table variable **across** the page instead of down it: each of its levels becomes a *block of columns*. Put it together with `levels = "first"` (one column per variable, see above) and you get the most condensed table the package can draw --- several variables, several groups, one screen:

```{r, echo = TRUE}
tab(gss_simple, rincome, c(married, tvhours), tab_vars = race, spread_vars = race,
    pct = "row", na = "drop", levels = "first", comp = "all",
    color = TRUE, color_signif = "grey_non_signif")
```

The layout follows from the shape, and the table says so:

- there is **one `Total` row** for the whole table; each block answers in its own columns.
- the base count takes **one `n` column per block**, gathered at the right, so the counts can be read against each other. A `Total` column per block would only repeat `100%`.
- `comp = "all"` compares every cell against the **overall** total rather than its own group's --- here the `Ensemble` block's `Total` cell, the one number every shade in the table is measured from. The color legend names it.
- a total *line* cannot become a block of columns, so `totaltab = "line"` (the default) is promoted to `"table"`, and the message says so. Say `totaltab = "no"` if you want no overall block at all.

A variable named in `spread_vars` alone is added to `tab_vars` for you, so `tab(gss_simple, rincome, party3, spread_vars = race, pct = "row")` is a complete call.

## Batteries of yes/no items, and a score

**Multiple-answer questions** — "which of these apply to you?" — reach the data as a *battery* of yes/no factors, one per item. `levels = "first"` keeps only the first level of each column factor, so a whole battery fits in one compact table, one column per item.

`facto_tea`, which ships with tabxplor (300 tea drinkers, from the **FactoMineR** package of François Husson, Julie Josse, Sébastien Lê and Jérémy Mazet — with thanks; see `?facto_tea`), has such a battery: *when do you drink tea?*, asked as six yes/no items. Everything below depends on one thing — the "yes" answer has to be each factor's **first** level, which is how the shipped copy stores it:

```{r, echo = TRUE}
tea_when_vars <- c("breakfast", "tea.time", "evening", "lunch", "dinner", "always")
levels(facto_tea$breakfast)     # the "yes" answer first: what every table below reads
```

`score_from_lv1()` reduces that same battery to a single **summed score**: for each person it counts the factors sitting at their first level — here, at how many of the six moments they drink tea (missing values never count). The score is an ordinary numeric variable, so it takes its place in the same table, as a mean:

```{r, echo = TRUE}
tea <- facto_tea |> score_from_lv1("tea_when", vars_list = tea_when_vars) # score variable
tab(tea, SPC, all_of(c(tea_when_vars, "tea_when")), pct = "row", levels = "first", na = "drop", color = TRUE)
```

Each percentage reads down its own column: 63 % of senior managers drink tea at breakfast, against 25 % of manual workers. There is no Total column, and rightly so — the six items do not add up to 100 %, since one person can tick several. The score plays that role instead: 2.1 moments a day on average for senior managers, 1.5 for manual workers.

`levels = "auto"` makes that choice **per variable**: it keeps the first level only of two-level factors, and every level of the others. It is what you want when a battery of yes/no items sits beside an ordinary factor — here two tea moments and the social category, in one table:

```{r, echo = TRUE}
tab(tea, sex, c(breakfast, evening, SPC), pct = "row", levels = "auto", na = "drop", tot = "row")
```

Put a battery together with `spread_vars` (above) and you get the most condensed table the package draws — six items, two sexes, one screen:

```{r, echo = TRUE}
tab(tea, SPC, all_of(c(tea_when_vars, "tea_when")), tab_vars = sex, spread_vars = sex,
    pct = "row", levels = "first", na = "drop", comp = "all", color = TRUE)
```

A summed score is also exactly what a **grouped-binomial** regression models — see `trials` in `vignette("tabxplor-reg")`.

## Weights, in one minute

Give `wt =` a weight column and every percentage and mean becomes an estimate of the *population* rather than of the people you happened to interview:

```{r, echo = TRUE}
gss_w <- dplyr::mutate(gss_simple, w = ifelse(marital %in% "Never married", 2.5, 0.8))
tab(gss_w, race, party3, wt = w, pct = "row", na = "drop")
```

That is the easy half. The **margins of error** around those percentages have three levels, and a table's footer always says which one it is on:

1. **weighted percentages, plain margins of error** — the default, and the convention almost every textbook uses. Under unequal weights it runs a little too narrow.
2. **`design_effect = TRUE`** — every interval, star, colour threshold and test accounts for the unequal weighting, exactly.
3. **a `survey` design passed as `data`** — everything follows the full design: strata, clusters, calibration.

Which one you need, what each costs you, and whether it matters for your table: `vignette("tabxplor-weights")`.


## Exporting

A finished table exports with its colors to Excel, HTML or Markdown:

```{r, echo = TRUE, eval = FALSE}
tabs <- tab(gss_simple, race, party3, pct = "row", color = "difference")
tab_export(tabs) # default : html table (RStudio Viewer, .Rmd/.qmd, etc.)
tab_export(tabs, format = "xl", path = "table") # Excel export 
tab_export(tabs, format = "md", path = "table") # flat markdown file (pipes tables)
```

**Getting a table into Word** goes through Excel, and that is the recommended route rather than a workaround: the workbook holds the *real* numbers, not rounded strings, so you can still fix a decimal or a label there. Then copy the cells and paste them into Word — **from the desktop app, not the browser version**, which drops the formatting. Colours, bold and borders all survive.

Two options are worth knowing:

- `theme = "auto"` lets an HTML or Markdown export **follow the reader's light/dark mode** (it flips live). For the console, `set_color_palette(theme = "auto")` detects the editor (RStudio, Positron, etc.) and picks the matching palette — it is applied automatically when the package loads.

```{r, echo = TRUE, eval = FALSE}
tab_export(tabs, theme = "auto") # HTML that follows the reader's light/dark modes
```

- Since numeric variables can only be passed in columns, some complex layout with numeric variables in rows need to transpose the table during export using `transpose = TRUE` :

```{r, echo = TRUE}
tab(gss_simple, party3, c(race, tvhours), pct = "row") |>
  tab_html(transpose = TRUE)
```

- **One stylesheet for a whole document.** In an `.Rmd`/`.qmd` report, `tab_css()` writes the colour CSS once and every later table emits only classes, so a single `theme` — including `"auto"`, which follows the reader's light/dark mode — styles every table at once. This very vignette does exactly that (with `theme = "light"`):

```{r, echo = TRUE, eval = FALSE}
options(tabxplor.tab_kable_css = FALSE)
tab_css(theme = "auto")   # emit once, near the top of the document
```

Nothing is written inline on a cell, so any look is overridable with plain CSS afterwards (column widths, fonts…); see `?tab_css` for the role classes (`.tx-rv`, `.tx-tot`, `.tx-num`).

### Black and white, for publication

Colors are for exploring. For a journal, `theme = "print_minimalistic"` renders the same reading in black and white: **over-represented cells in bold**, *under-represented ones in italic*, an underline for the strongest threshold, and a grey fill for a second measure. The legend names the typography instead of the hues.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "difference") |>
  tab_html(theme = "print_minimalistic")
```

This is not merely a taste: converted to lightness, the two directions of the color palette come out as the *same* shades of grey, so a grayscale print loses over-vs-under entirely. It works in every export — `tab_html()`, `tab_md()` and `tab_xl()` (real bold/italic/underline in the spreadsheet) — and the typography is written as `<b>`/`<i>`/`<u>` markup as well as CSS, so it also survives a paste into Word.

**You rarely need to ask for it.** Every stylesheet already carries this palette in an `@media print` block, so a colored html table *prints* — or saves to PDF from your browser — publication-ready on its own. Set `options(tabxplor.print_rules = FALSE)` if your printer is a color one and the colors are the point.

## Hover tooltips (html tables)

Every html table carries per-cell hover **tooltips** with the numbers behind the cell: the unweighted count, the difference from the reference, the ratio, the confidence interval... They are on by default in the Viewer and in reports — this vignette only switched them off document-wide with `options(tabxplor.tab_kable_tooltips = FALSE)`, to keep the page light. Hover the cells of the table below, where they are switched back on with `tooltips = TRUE`:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "difference") |>
  tab_html(tooltips = TRUE)
```

**A note on weights.** With a weight (`wt =`), every proportion or mean is weighted, but by default the sample size behind the confidence intervals and tests stays the real, **unweighted** number of observations — the footer says so. Under unequal weights that carries no design effect, so it runs a little too narrow : `options(tabxplor.design_effect = TRUE)` (see [Weights](#weights)) widens every interval by exactly the weighting's own design effect and switches the whole-table Chi2 / F tests to their design-based counterparts.

## A chart of the table: `forest_plot()`

A colored table already reads at a glance. When the pattern is what matters, `forest_plot()` draws the same numbers: one whisker per cell, in the cell's own colour, with the percentage or mean printed just above it.

```{r, echo = TRUE, eval = FALSE}
tab(gss_simple, race, party3, pct = "row", ci = "ref", color = "diff") |>
  forest_plot()
```

What the position shows is the **deviation** — the comparison your `color =` grades: a difference from the reference in percentage points, a ratio or an odds ratio on a log axis. The **level** it sits on is the number above the whisker, so the two say different things. Ask for `ci = "ref"` when you build the table, or there is no interval to draw. `what = "level"` swaps them.

Three things are worth knowing:

* **The gridlines are your colour breaks** — one dashed rule per threshold, in that threshold's own colour, continued beyond the last one (each step twice the one before) as far as your data goes. Change them with `set_color_breaks()` and the axis moves with the table.
* **Significance is the whisker.** It is coloured whole, so an interval that crosses the reference line is visible without stars — and there are none.
* **It reads the table, and computes nothing.** It returns a plain `ggplot`, so `+ ggplot2::labs(...)`, `+ ggplot2::theme(...)` and `ggsave()` work; `theme = "print_minimalistic"` gives the greyscale version, `guide = "bands"` shades the panel behind the whiskers with the colour scale itself (a good way to *show* a class what the colours mean).

`forest_plot()` draws regression tables too — see `vignette("tabxplor-reg")`.

## Working with the result

`tab()` returns a `tibble` (of class `tabxplor_tab`), so `dplyr` verbs just work. Use the helper `is_totrow()` to keep the Total row in place when you re-order (it flags total rows, so sorting on it first sends them to the bottom):

```{r, echo = TRUE, message = FALSE}
library(dplyr)
tab(gss_simple, race, marital, pct = "row") |>
  arrange(desc(Married))
```

**Titling and annotating.** `subtext =` prints one or more legend lines under a table (a data source, a note). `set_caption()` gives a table a **title that survives a dplyr pipeline**, and every exporter uses it as the table title:

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", subtext = "Source: GSS, 2000-2014") |>
  set_caption("Custom title")
```

## Session options

A handful of `options()` set your preferred defaults once for the whole session — put them at the top of a script, or in your `.Rprofile`. Each one has a per-call argument too; the option just changes the default. The everyday ones:

- `options(tabxplor.print = "html")` — print tables not in console, but as html in RStudio or Positron Viewer Pane by default (recommended)
- `options(tabxplor.cleannames = TRUE)` — strip `"1-"`-style prefixes from level names everywhere.
- `options(tabxplor.parallel = 8)` — parallelise tables with multiples variables on different CPU cores by default (needs `mirai`)
- `options(tabxplor.var_labels = TRUE)` — in exports, show a variable's label (from `haven`/`labelled` data) instead of its bare name.
- `options(tabxplor.theme = "auto")` — the export theme (`"light"`/`"dark"`/`"auto"`); `set_color_palette(theme = "auto")` does the same for the console.
- `options(tabxplor.stars = TRUE)` — show significance stars in every table (like `stars = TRUE`).
- `options(tabxplor.conf_level = 0.9)` — the confidence level for intervals and tests (default `0.95`).
- `options(tabxplor.design_effect = TRUE)` — on weighted data, make every interval, star, colour threshold and test account for the unequal weighting (see `vignette("tabxplor-weights")`).
- `options(tabxplor.lang = "fr")` — the language of the colour legends and footers (`"auto"`/`"en"`/`"fr"`).

Colour thresholds and palettes have their own helpers, `set_color_breaks()` and `set_color_palette()`. `?tabxplor-options` documents every option, and `vignette("tabxplor-programming")` covers the more advanced ones (export fonts, parallel builds…).

## Where to go next

- `vignette("tabxplor-all-else-equal")` — regressions, taught the way this vignette teaches cross-tables: one analysis walked from a first table to a finished sentence.
- `vignette("tabxplor-reg")` — the regression reference: every family, every argument, the model checks and the plots.
- `vignette("tabxplor-weights")` — weighted and survey data, for both producers.
- `vignette("tabxplor-programming")` — the `tabxplor_fmt` cell type and how to program with its fields.
- `?tab` for every argument (grouped by purpose, including the confidence-interval methods) and `?tabxplor-options` for the package-wide defaults.
