---
title: "Regression tables with tab_reg()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression tables with tab_reg()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tabxplor)
library(dplyr)

# As in the introduction vignette, tables are shown as the coloured console output turned to HTML.
options(cli.num_colors = 256)
set_color_palette(theme = "light")
```

```{r, echo = FALSE, include = FALSE}
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

For the most common regression models, `tab_reg()` builds a **regression table** that looks and behaves like a `tabxplor` cross-table: one row per predictor level, significance stars, colours that grey out non-significant effects, and the same Excel, html or markdown exports. You give it a data frame, an **outcome** (`dependent`) and some **predictors**, and it tries to pick the right kind of model from the outcome's type. Its distinctive feature is `empirical = TRUE`, which shows the **observed / crude / empirical effect right next to the model's adjusted one**, so you can see what "controlling for the other variables" actually changed.

We use a formatted version of the `forcats::gss_cat` data, from the US General Social Survey:

```{r}
gss_simple <- gss_cat_data_formatting()
```


## The outcome's type chooses the model and the observed quantity to compare with

You rarely set `family` by hand — `tab_reg()` mostly detects it: 

| Outcome | Detected model | Effect measure |
|:--------|:---------------|:---------------|
| 2-level factor | binomial (logistic) | odds ratio (OR) |
| numeric (continuous) | gaussian (linear) | coefficient β |
| count | poisson | incidence-rate ratio (IRR) |
| 3+ level unordered factor | multinomial | one OR column per category vs. the reference |
| 3+ level ordered factor | ordinal (proportional odds) | cumulative OR |

With `empirical = TRUE`, each model column is joined by a crude/observed companion column showing the *empirical, unadjusted, single-predictor* effect — the effect you would see with no controls at all, "all things being unequal", for that predictor.
- 2 levels logistic (binomial) → observed %, and observed odd-ratios (not modelised: calculated from percentages only)
- 3+ levels logistic (multinomial) → the observed ORs are shown as a tooltip on the model cells in html exports
- gaussian (linear) → group means and their difference
- poisson (counts) → observed rate and observed rate ratio


## Logistic regression (a binary factor)

When the outcome is a two-level factor, here "married" versus "not married", `tab_reg()` chooses a **binomial** family to fit a logistic regression and reports **odds ratios** (the reference level of each predictor shows the neutral value `1`). 
- Like in every regression model, the effect of a predictor level on the dependent variable reads "all the others chosen predictors being equals".
- Colors read like any `tabxplor` table: an odds ratio above 1 (blue) means *more likely to be married* than the reference level, below 1 (red) means *less likely*; stars and colors both flag significance.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"))
```

For a simple logistic regression, `empirical = TRUE` add **raw percentages** and **crude unmodelised odds-ratios** for each predictor.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), empirical = TRUE)
```

The raw percentages `Obs_%` are a summary of the base observed results everything in the model is derived from :"28% of black americans are married, compared to 51% for white americans"
- The plain difference from reference is colored (28% - 51% = -23%).
- Cells that are non-significantly different from the reference are greyed out (based on a Newcombe confidence interval for differences of proportions).

The **modelised odds-ratios** `Model_OR` directly compared to the **observed odds-ratios** `Obs_OR` :
- Comparing the two tells you what adjustment the model did: if a predictor's **model** OR is much closer to 1 than its **crude** OR, the raw association was largely explained by the other predictors.
- Here, "all things being unequal", black americans have 2.68 lower odds to be married than white americans. "All things being equal" (more precisely : income, age and religion being equal), black americans still have 2.4 lower odds to be married than white americans. The result holds, it’s not explained by income differences or religious differences.

The **observed odds-ratios** `Obs_OR` are the same as those you can calculate from percentages only in a crosstable :
- Colors and significance stars use a **Woolf OR confidence interval** that matches what is done in the regression model. 
- The population of the table must match the complete-cases population of the model, filtering out indivivuals with `NA` at any involved variable.

```{r}
gss_simple |> 
  dplyr::filter(dplyr::if_all(all_of(c("race", "age", "rincome", "relig")), ~ !is.na(.) )) |> 
  tab(race, married, pct = "row", na = "drop", 
    OR = "OR", color = "OR", color_signif = "grey_non_signif"
   )
```

### Average marginal effects (AME) of a logistic regression

Another way to interpret a logistic regression is to use **average marginal effects** (AME) with `effect = "ame"` :
- Instead of directly reading the modelised odd-ratio we can infer, from it, the average difference of percentage of each level compared to the reference.
- Here `Model_AME` reads like this : "Comparing Black and White respondents who are alike in income, age and religion, being Black is associated with a marriage rate 19.8 points lower, on average."

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), effect = "ame", empirical = TRUE)
```

Then we can use `empirical = TRUE` to compare this new modelised quantity to the **raw, observed difference of percentages** : 
- In the sample (`Obs_diff`), we observe a rate of marriages 23 points lower for black americans compared to white american (`28% - 51% = -23%`).
- The modelised difference `Model_AME` is very close, `-19.8%`, so very little of this difference between black and white can be explained by age, income or religion.

We can also compare the raw percentage `Obs_%` with the **model-adjusted probability** — the value in parentheses in the `Model_AME (adjusted %)` column :
- Adjusted percentage, which is a result of the model, reads this way : "if the whole sample kept its income/age/religion distribution but everyone were Black, an estimated 31.5% would be married, vs 51.3% if everyone were White". 
- Blacks marry at 30.9% in this sample (observed); standardizing their income/age/religion to the population mix moves that only to 31.5% — so almost none of the Black–White gap is accounted for by differences in those three variables; it persists after adjustment.

We can, again, get the same empirical differences and confidence intervals with a plain table : 

```{r}
dependent  <- "married"
predictors <- c("race", "age", "rincome", "relig")
factor_predictors <- c("race", "rincome", "relig")
gss_simple |>
  dplyr::filter(dplyr::if_all(all_of(c(predictors, dependent)), ~ !is.na(.) )) |> 
  tab(all_of(factor_predictors), all_of(dependent), pct = "row", na = "drop", 
      color = "diff", ref = 1, color_signif = "grey_non_signif", stars = TRUE, 
      method_diff = "wald"
      ) |>
  dplyr::mutate(diff = set_display(`01-Married`, "diff"))
```



## Grouped-binomial outcomes (a summed score)

When the outcome is a **summed score** — how many of several yes/no items a respondent answered one way — you model the number of "successes" out of a fixed number of items with `trials =`. `tab_reg()` then fits `cbind(score, trials - score)` as a binomial, so the odds ratios read on the *per-item* probability. `score_from_lv1()` (see `vignette("tabxplor-programming")`) builds such a score by counting factors at their first level:

```{r}
gss_simple |>
  score_from_lv1("score", vars_list = c("married", "income25k")) |>   # a 0–2 score
  tab_reg("score", c("race", "age"), family = "binomial", trials = 2)
```

Grouped-binomial (like Poisson) models report a Pearson **dispersion** check in the footer, flagging over-dispersed counts.

## Ordinal and nominal outcomes (a factor with 3+ levels)

An **ordered** factor outcome is fit as a proportional-odds (cumulative) logistic model:

```{r}
tab_reg(gss_simple, "rincome", c("race", "age", "relig"))
```

A nominal outcome with three or more unordered levels is fit as one **multinomial** logistic model, giving one odds-ratio column per outcome category versus the reference category (also called relative risks ratios) :

```{r}
tab_reg(gss_simple, "party3", c("race", "age", "rincome", "relig"))
```

Relative risk ratios may be quite difficult to read because they are relative to **two** reference levels : not only the reference chosen for the predictor, but also the reference level chosen for the dependent variable. It’s specially difficult when it’s difficult to find a good reference level corresponding to the most common situation (like "married" for matrimonial status).

Most of the time, average marginal effects (AME) are easier to interpret, because they make the second reference level disappear, and directly modelise, for each level of the dependent variable, difference of percentages of each predictor compared to their reference level (a less abstract quantity than odds ratios). 

```{r}
tab_reg(gss_simple, "party3", c("race", "age", "rincome", "relig"), effect = "ame", empirical = TRUE) # |> tab_export()
```

With `empirical = TRUE`, the crude/observed/unajusted differences and raw percentages appear in html tooltips, available at mouse hover of a cell.


## Linear regression (a numeric outcome)

A continuous outcome gives plain linear regression coefficients (here we set `family` explicitly, because an integer like `age` is ambiguous — it could also be modelled as a count):

```{r}
tab_reg(gss_simple, "age", c("race", "marital", "relig", "rincome"), family = "gaussian")
```

In the case of a linear regression, the **empirical/observed counterpart of the model coefficient** for a categorical predictor is simply the **difference of means** : here, the difference of mean age, per level of the predictor, compared to the reference level. 
```{r}
tab_reg(gss_simple, "age", c("race", "marital", "relig", "rincome"), family = "gaussian", empirical = TRUE)
```

The empirical mean ages, and differences of mean ages from reference, can be computed in a simple table with : 

```{r}
tab(gss_simple, "race", "age", pct = "row", digits = 2, na = "drop",
    color = "diff", ref = 1,  method_mean_diff = "student"
) |> 
  mutate(diff = set_display(age, "diff"))
# student : confidence intervals with pooled variance, to match those computed by linear regression
```


## Poisson regression (a count outcome)

```{r}
tab_reg(gss_simple, "tvhours", c("race", "marital", "relig", "rincome"), family = "poisson")
```

An **incidence-rate ratio** (IRR) of 1.5 means "50% more hours of TV per day". Unweighted Poisson models automatically use dispersion-scaled (quasi-Poisson) standard errors, so over-dispersed counts get honest, wider intervals. Concretely: with an over-dispersed outcome, `family = "poisson"` returns CIs and p-values **identical to `family = "quasipoisson"`** and emits a warning saying so (the footer reports the dispersion); at equidispersion (≈ 1) the scaling is a no-op and the result matches a plain `glm(family = poisson)` — so a comparison to a hand-fit Poisson `glm` never surprises you.

In the case of a poisson regression, the **empirical/observed counterpart of the model’s exponentiated coefficient** for a categorical predictor is the **ratio of means** : here, the ratio of average TV hours compared to the reference level.

```{r}
tab_reg(gss_simple, "tvhours", c("race", "marital", "relig", "rincome"), family = "poisson", empirical = TRUE)
```

The empirical average TV hours per day, and differences of rate of television watching from reference, can be computed in a simple table with : 

```{r}
tab(gss_simple, "race", "tvhours", pct = "row", digits = 2, na = "drop",
    color = "ratio", ref = 1,  method_mean_ratio = "quasipoisson"
) |> 
  mutate(IRR = set_display(tvhours, "ratio"))
# default method for confidence intervals is robust ratio of means with unequal variable ;
#  we use "quasipoisson" to match those computed by quasi-poisson regression 
#  (assumption : variance is proportional to mean). 
```

## Weighted and survey data

Passing a weight column with `wt =` switches estimation to a **survey design** (`survey::svyglm()`), which gives correct *design-based* standard errors. The framework `tabxplor` uses is simple: the **estimate is weighted**, and the model's standard errors come from the survey design, so unequal weights widen the confidence intervals honestly. (This is the design-based counterpart to `tab()`'s cheaper `options(tabxplor.kish_neff = TRUE)` approximation — see `vignette("tabxplor")`.)

```{r, eval = FALSE}
# `weight` is a survey weight column in your data, using mostly a basic  :
tab_reg(data, "dependent", c("pred1", "pred2"), wt = "weight")

# A stratified / clustered design (strata usually *narrow* the intervals, clusters widen them):
tab_reg(data, "dependent", c("pred1", "pred2"),
        wt = "weight", ids = "psu", strata = "stratum")

# Or pass a design you built yourself with survey::svydesign() as `data` (see ?survey::svydesign
# for calibration, replicate weights and the more complex cases).
```

If you only have simple weights, `wt =` is all you need. Give `strata =` when you have a stratified survey: it typically buys a little precision (narrower confidence intervals) when the strata variables are related to the outcome.

The `empirical = TRUE` observed companions columns (the unadjusted `Obs_*` effects) are *descriptive*, so on weighted data their confidence intervals honour `options(tabxplor.kish_neff = TRUE)` exactly like `tab()`. Set it to `TRUE` when using `tab_reg()` with weights for a more like-for-like uncertainty comparison (the model columns themselves stay design-based regardless, so there’s always a remainnig discrepency between the two ways to calculate confidence intervals). 


## Comparing several models

Pass a **named list** of predictor sets instead of a vector to fit and show several models side by side, and `compare =` adds a comparison Likelihood ratio test in the footer (`"baseline"` = each model vs. the first; `"sequential"` = each vs. the previous):

```{r}
tab_reg(gss_simple,
        "married",
        list("Race only"    = "race",
             "+ age"        = c("race", "age"),
             "+ party"      = c("race", "age", "party3")),
        compare = "sequential")
```

## The same model within sub-populations

`split_var =` is the regression analogue of `tab()`'s `tab_vars`: the same model is fitted **within each level** of a grouping variable, and the per-group tables are stacked into one grouped table. It answers "does this effect hold in every subgroup?". With just one dependent variable, it uses `tab_spread()` internally to lay the groups out as side-by-side columns for an easy comparison : 

```{r}
tab_reg(gss_simple, "married", c("race", "rincome"), split_var = "year")
```

## Fine-tuning the display, and scaling a predictor

A few arguments change *what each cell shows*, or *how a predictor is scaled* — without changing the model itself.

**`estimate_display`** folds a second quantity into each estimate cell. `"ci"` shows the confidence interval beside every estimate (any family):

```{r}
tab_reg(gss_simple, "married", c("race", "age"), estimate_display = "ci")
```

For a logistic model you can instead fold the model-adjusted predicted probability (`"prob"`) or the average marginal effect (`"ame"`) into the odds-ratio cell (these two need the `marginaleffects` package). Note the difference: `estimate_display = "ame"` *keeps* the odds-ratio column and adds the AME beside it, whereas `effect = "ame"` (seen above) turns the **whole** column into an AME.

**`multiplier`** rescales a continuous predictor's effect. A one-year change in `age` barely moves the odds, so its odds ratio sits near 1; `multiplier = c(age = 10)` reports the effect **per 10 years** instead (the odds ratio raised to the power 10, with its interval scaled to match):

```{r}
tab_reg(gss_simple, "married", c("race", "age"), multiplier = c(age = 10))
```

**`at = "reference"`** evaluates the effect at the reference profile (the abstract population combined the reference level of all predictors) instead of averaging it over the sample — a marginal effect *at the reference* for `effect = "ame"`, or a "category vs. the rest" odds ratio at that profile for a multinomial model:

```{r, eval = FALSE}
tab_reg(gss_simple, "married", c("race", "age"), effect = "ame", at = "reference")
```


## Reading the footer statistics

The footer summarises model fit; which statistics appear depends on the family:

- **N** — the number of observations used.
- **LR vs null** — a likelihood-ratio test of the whole model against the intercept-only model (is the model worth anything?).
- **McFadden R²** — a pseudo-R² for binomial and poisson models (higher = the predictors explain more; values are much smaller than a linear R²).
- **AIC / BIC** — information criteria for comparing models on the same data (lower = better; BIC penalises complexity more).
- Linear models (gaussian) add the usual **R² / adjusted R² / F / σ**; Poisson adds a **dispersion** check.

For weighted models a reduced set is reported (a Wald test against the null, Nagelkerke / Cox–Snell pseudo-R², a Rao–Scott AIC), because the likelihood-based quantities do not apply to a design-based fit.

## Plots

`or_plot()` draws a forest plot of the odds ratios of a finished table, and `lm_plots()` draws the standard 2×2 linear-model diagnostics:

```{r, eval = FALSE}
t <- tab_reg(gss_simple, "married", c("race", "age"))
or_plot(t)

m <- tab_reg(gss_simple, "age", c("race", "marital"), family = "gaussian")
lm_plots(m)
```

## Where to go next

- `tab_reg()` is also available **without writing R code**, as a point-and-click **Regressions** analysis in the [jamovi](https://www.jamovi.org/download) module — install *tabxplor* from jamovi's module library (see `vignette("tabxplor")`).
- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `?tab_reg` for every argument (grouped by purpose), and `?tab_reg` &rarr; *Details* for the modelling choices.
- `survey::svydesign()` for complex designs (calibration, replicate weights, finite-population corrections).
