---
title: "Introduction to tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option

# The tables below are tabxplor's real html tables, rendered live: the setup sets
# options(tabxplor.print = "html") --- the recommended everyday setting in RStudio/Positron ---
# so every bare tab() chunk knits as a colored html table. The shared stylesheet is emitted ONCE
# by the tab_css() chunk below (tab_kable_css = FALSE stops each table re-inlining it), and the
# hover tooltips are off document-wide (see the dedicated tooltips section, which re-enables them).
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

# The few console examples still show terminal colors: cli emits ANSI (options("cli.num_colors")),
# and the fansi hook below turns that ANSI into colored HTML.
options(cli.num_colors = 256)
set_color_palette(theme = "light") # type = "text"
```

```{r, echo = FALSE, results = "asis"}
cat(tab_css(theme = "light"))
```

The first time only, install the package : 

```{r, eval = FALSE}
install.packages("tabxplor", dependencies = TRUE)
```

At the start of each new R session, load it :

```{r, eval = FALSE}
library(tabxplor)
```


```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

`tabxplor` helps you **explore data with cross-tables, coloring the cells so you can read a table at a glance**. Over-represented cells use shades of blue, under-represented ones turn red/orange, so patterns jump out without you squinting at every number.

Everything is a `tibble`, so the result works with the usual `dplyr` verbs, and tables export to Excel, HTML and Markdown with their color helpers. Underlying heavy computations run on `data.table`

Throughout this vignette we use `gss_simple`, a cleaned-up version of the US General Social Survey (`forcats::gss_cat`) with factors levels merged and reordered.

```{r, echo = TRUE}
gss_simple <- gss_cat_data_formatting()
```

## Your first cross-tables

`tab()` needs a data frame, a **row variable** and a **column variable**. By default it shows counts:

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

This is tabxplor's **html table** — what you get in the RStudio or Positron Viewer pane with the recommended session option, used throughout this vignette:

```{r, eval = FALSE}
options(tabxplor.print = "html")
```

Without the option, the same table prints in the **console**, as a colored `tibble` — same information, lighter display:

```{r, include = FALSE}
options(tabxplor.print = "console")
```

```{r, echo = TRUE}
tab(gss_simple, marital, race)
```

```{r, include = FALSE}
options(tabxplor.print = "html")
```

Add `pct = "row"` for row percentages (or `"col"` for column percentages). A **Total** row/column and a count column (`n`) are added automatically:

```{r, echo = TRUE}
tab(gss_simple, marital, race, pct = "row")
```

When the column variable is **numeric**, `tab()` shows its **mean** in each row instead of percentages:

```{r, echo = TRUE}
tab(gss_simple, marital, age)
```

You can pass **several row and column variables at once**. 

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), c(party3, tvhours), na = "drop", pct = "row")
```

`levels = "first"` keeps only the first level of each column factor, which is handy to display many binary factors, like survey questions with multiple answers, all at once, in a compact way :

```{r, echo = TRUE}
tab(gss_simple, relig, c(married, black, income25k), pct = "row", levels = "first", na = "drop", cleannames = TRUE)
```

A few other everyday arguments: `na = "drop"` to drop missing values from the base, `digits =` for the number of decimals, and `cleannames = TRUE` to strip prefixes like `"1-"` from level names. See `?tab` for the full list.

## Weights

The argument `wt =` adds a survey weight. Every percentage and mean is then **weighted**, while the sample size behind the confidence intervals stays the real, **unweighted** number of cases — the honest default basis for uncertainty.

```{r, echo=TRUE, eval=FALSE}
data(hdv2003, package = "questionr")
tab(hdv2003, nivetud, occup, wt = poids, pct = "row", na="drop", digits = 1)
```

Under **unequal** weights that default interval carries no design effect, so it runs a little too narrow. Kish's *effective sample size* fixes this : it counts each observation by how much it really contributes, `n_eff = (sum of w)² / (sum of w²)` (always at most the real *n*), and uses `n_eff` in place of the raw *n*. Turn it on and every weighted confidence interval in the table — proportions and means alike — widens honestly:

```{r, echo=TRUE, eval=FALSE}
options(tabxplor.kish_neff = TRUE)
```

This is a simple single-stage approximation (it needs the individual weights, so it is not available for tables built from pre-aggregated counts). A **fully design-based** result — clusters, strata, exact standard errors — is available for Chi2 pvalues (factors) and ANOVA F pvalues (numeric column variables) only with `test = "survey"` and the related arguments (see the `test =` argument) : confidence intervals are not covered. See the [survey](https://CRAN.R-project.org/package=survey) package for more informations about survey design. `tab_reg()` regressions tables with weights always use design-effect for standard errors (`vignette("tabxplor-reg")`), but the base version only use Kish's effective sample size.


## Sub-tables

Give `tab()` a third variable as `tab_vars` and it builds **one sub-table per group** (here, one per income group). The result is *grouped*: `dplyr` operations then run within each sub-table.

```{r, echo = TRUE}
tab(gss_simple, race, party3, rincome, na = "drop", pct = "row")
```

When you pass several **row variables** *without* `tab_vars`, `tab()` merges the mirror tables into a single table by default. `output_list = TRUE` returns instead a **list with one table per row variable** (with `tab_vars`, the result is always a list):

```{r, echo = TRUE, eval = FALSE}
tab(gss_simple, c(married, income25k), race, pct = "row", output_list = TRUE)
```

## colors: reading helpers

One you the main purposes of `tabxplor` is to provide a full palette of color helpers for data exploration. `color = "diff"` colors each cell by **how far it sits from its reference** — by default the Total of its row or column. Cells clearly above the average turn **blue**, cells clearly below turn **red/orange** — the further a cell sits from its reference, the stronger the shade — and a color legend is printed underneath.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff")
```

`color = TRUE` picks a sensible scheme automatically for each column type (both differences and ratio for percentages, only ratios for means, …), check which one in the legend:

```{r, echo = TRUE}
tab(gss_simple, rincome, c(party3, marital), pct = "row", color = TRUE)
```

Numeric columns are colored the same way, on their **means** (here, hours of TV per day by income):

```{r, echo = TRUE}
tab(gss_simple, rincome, tvhours, color = "diff")
```

**Which cell is the reference for comparison ?** By default each cell is compared to the relevant Total (Total row for row percentages and Total column for column percentages) to highlight over-representations and under-representations. Two useful alternatives:

- `ref = 1` compares each row to the **first row** — perfect for reading an evolution over time or an ordinal factor.
- with sub-tables, `comp = "all"` compares against the **overall** Total instead of each sub-table's own Total.

```{r, echo = TRUE}
tab(gss_simple, year, marital, pct = "row", color = "diff", ref = 1)
```

```{r, echo = TRUE}
tab(gss_simple, rincome, party3, race, na = "drop", pct = "row", color = TRUE, comp="all")
```

**A different reference for each variable.** `ref` is reinterpreted by `pct`. Under **row** percentages (or means) it picks a reference **row**, so a *named* vector gives each row variable its own — here `race` is read against its first row, `relig` against its Total:

```{r, echo = TRUE}
tab(gss_simple, c(race, relig), party3, pct = "row", color = "diff",
    ref = c(race = 1, relig = "tot"), na = "drop")
```

Under **column** percentages `ref` picks a reference **column** instead, vectorised over the column variables — either named (`ref = c(party3 = "first", marital = "tot")`) or positional, one value per column variable:

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, marital), pct = "col", color = "diff",
    ref = c("first", "tot"), na = "drop")
```

Color thresholds and the palette can be customised : set them **once for the whole session** with `set_color_breaks()` and `set_color_palette()`.

## colors that respect significance

The colors above show the *size* of a deviation, but not whether it is **statistically reliable**. On small samples a big-looking difference can be pure noise. The `color_signif` argument brings significance into the coloring:

- `"ignore"` (default): color every deviation by its observed size. Grey out small differences below a certain threshold.
- `"grey_non_signif"`: color by size of the effect, grey out small effects below a certain threshold, but also **grey out cells with important effects that are not significant**. Every colored cell is then guaranteed to be significantly different from its reference, without being bothered by very small significant differences. 
- `"guaranteed_effect"`: color only by the part of the effect you can be confident about (its confidence bound), with dimmer, conservative colors. Use it on **small samples** to **highlight all the differences you have the right to interpret**. Everything colored is significant ; nothing grey is.

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff", color_signif = "grey_non_signif")
```

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = "diff", color_signif = "guaranteed_effect")
```

On **small samples** a big-looking percentage can rest on a handful of respondents. `n_min =` is a purely visual filter, applied last: it hides cells whose (unweighted) base is below the threshold and drops a row entirely when its largest base falls short. Here the rarest religions drop out:

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row", n_min = 400)
```

An alternative is to keep the small rows and cols but group them all in a "Other" level : 

```{r, echo = TRUE}
tab(gss_simple, relig, race, pct = "row",  other_if_less_than = 400)
```



## Confidence intervals, tests and contributions

Print confidence intervals for the percentage or mean of each cell with `ci = "cell"` :

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", ci = "cell") # by default, conf_level = 0.95
```

Print the confidence intervals of the **difference** with a reference, used to calculate significance (if 0 belongs to the confidence interval, the cell is not significantly different from the reference, here the Total row) :
```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(race, party3, pct = "row", color = TRUE, color_signif = "guaranteed_effect",
      display = "num_ci" # "{pct} {ci}"
  )
```

`display = "num_ci"` is a type-adaptive shorthand for this: it shows each value with whatever confidence interval the table computes — `{pct} {ci}` on percentage columns and `{mean} {ci}` on numeric columns, chosen per column — so it works for a mix of factors and numbers in one call:


Add significance stars with `stars = TRUE`. They tell the same story than confidence intervals of the difference with the reference, but for different confidence levels (99%, 95%, 90%) :

```{r, echo = TRUE}
gss_simple |>
  dplyr::filter(year == "2012") |> # n=1 974
  tab(rincome, c(party3, tvhours), pct = "row", display = "num_ci", stars = TRUE)
```

**Show two numbers in one cell.** `display` is not just for confidence intervals: it takes a **`{}` template** that combines any cell fields. For example `display = "{pct} ({diff})"` prints each percentage followed by its difference from the reference, and `"{pct} (n={n})"` follows it with the count:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff", display = "{pct} ({diff})")
```

The first field in the template is the *primary* one — the value Excel keeps and the one the colours read. `vignette("tabxplor-programming")` explains the full grammar and lists every field you can combine.

`test = TRUE` adds a statistical test of independence per (sub-)table — **Chi-squared** for factor columns, **Welch's F ANOVA** for numeric variables (`options(tabxplor.anova = "classic")` switches to the pooled F):

```{r, echo = TRUE}
tab(gss_simple, race, c(party3, tvhours), pct = "row", test = TRUE)
```

`color = "contrib"` colors cells by their **contribution to the Chi-squared** — the cells that would stand out in a correspondence analysis :

```{r, echo = TRUE}
tab(gss_simple, race, party3, color = "contrib")
# tab(gss_simple, race, party3, pct="row", color = "contrib") # works with pct, but independent from rows/columns
```

See below for the detail of how confidence intervals and colors can be composed.

## Hover tooltips (html tables)

Every html table carries per-cell hover **tooltips** with the numbers behind the cell: the unweighted count, the difference from the reference, the ratio, the confidence interval... They are on by default in the Viewer and in reports — this vignette only switched them off document-wide with `options(tabxplor.tab_kable_tooltips = FALSE)`, to keep the page light. Hover the cells of the table below, where they are switched back on with `tooltips = TRUE`:

```{r, echo = TRUE}
tab(gss_simple, race, party3, pct = "row", color = "diff") |>
  tab_html(tooltips = TRUE)
```

**A note on weights.** With a weight (`wt =`), every proportion or mean is weighted, but by default the sample size behind the confidence intervals and tests stays the real, **unweighted** number of observations. Under unequal weights it carries no design effect, so it runs a little too narrow : opt in to Kish's effective sample size with `options(tabxplor.kish_neff = TRUE)` (see [Weights](#weights)) to widen every interval honestly and switch the whole-table Chi2 tests to a Rao–Scott correction.

## Exporting

A finished table exports with its colors to Excel, HTML or Markdown:

```{r, echo = TRUE, eval = FALSE}
tabs <- tab(gss_simple, race, party3, pct = "row", color = "diff")
tab_export(tabs) # default : html table (RStudio Viewer, .Rmd/.qmd, etc.)
tab_export(tabs, format = "xl", path = "table") # Excel export 
tab_export(tabs, format = "md", path = "table") # flat markdown file (pipes tables)
```

Two options are worth knowing:

- `theme = "auto"` lets an HTML or Markdown export **follow the reader's light/dark mode** (it flips live). For the console, `set_color_palette(theme = "auto")` detects the editor (RStudio, Positron, etc.) and picks the matching palette — it is applied automatically when the package loads.

```{r, echo = TRUE, eval = FALSE}
tab_export(tabs, theme = "auto") # HTML that follows the reader's light/dark modes
```

- Since numeric variables can only be passed in columns, some complex layout with numeric variables in rows need to transpose the table during export using `transpose = TRUE` :

```{r, echo = TRUE}
tab(gss_simple, party3, c(race, tvhours), pct = "row") |>
  tab_html(transpose = TRUE)
```

- **One stylesheet for a whole document.** In an `.Rmd`/`.qmd` report, `tab_css()` writes the colour CSS once and every later table emits only classes, so a single `theme` — including `"auto"`, which follows the reader's light/dark mode — styles every table at once. This very vignette does exactly that (with `theme = "light"`):

```{r, echo = TRUE, eval = FALSE}
options(tabxplor.tab_kable_css = FALSE)
tab_css(theme = "auto")   # emit once, near the top of the document
```

Nothing is written inline on a cell, so any look is overridable with plain CSS afterwards (column widths, fonts…); see `?tab_css` for the role classes (`.tx-rv`, `.tx-tot`, `.tx-num`).

## Working with the result

`tab()` returns a `tibble` (of class `tabxplor_tab`), so `dplyr` verbs just work. Use the helper `is_totrow()` to keep the Total row in place when you re-order (it flags total rows, so sorting on it first sends them to the bottom):

```{r, echo = TRUE, message = FALSE}
library(dplyr)
tab(gss_simple, race, marital, pct = "row") |>
  arrange(desc(Married))
```

**Titling and annotating.** `subtext =` prints one or more legend lines under a table (a data source, a note). `set_caption()` gives a table a **title that survives a dplyr pipeline**, and every exporter uses it as the table title:

```{r, echo = TRUE}
tab(gss_simple, race, marital, pct = "row", subtext = "Source: GSS, 2000-2014") |>
  set_caption("Custom title")
```


## Confidence intervals and colors composition : variable_type × `color` × `color_signif`

This section is the reference behind the two color sections above. It shows how the parameters fit together — the **variable type**, the **measure** you color (`color`), and the **significance policy** (`color_signif`).

The **type** is set by the column variable — a **factor** (percentages are computed, `pct`) or a **numeric** (means are computed, `mean`). The row variable is always turned into a factor.

Every colored table answers three questions:

- **How to measure deviation?** `color =` picks what a color *means*: `"diff"` (distance from the reference), `"ratio"` (relative risk for percentages, mean ratio for means), `"contrib"` (weight in the Chi-squared), `"OR"` (odds ratio). `color = TRUE` picks a sensible one per column type.
- **How confident are we in this measure?** every color reads one **confidence interval** at `conf_level` (95% by default). A cell is *significant* when that interval excludes its neutral value — **0** for a difference, **1** for a ratio or odds ratio. The printed bracket, the significance stars and the greying all read that same interval, so they can never disagree.
- **How to show significance?** `color_signif` — `"ignore"`, `"grey_non_signif"` or `"guaranteed_effect"` ; `stars = TRUE` to use significance stars instead of, or stacked with, colors.

**The confidence interval used for the colors and stars** compares each cell to its reference cell (by default, the corresponding cell in the Total row or Total column):

| type | color     | what the color measures                  | confidence interval (default) |
|------|-----------|------------------------------------------|-------------------------------|
| pct  | `diff`    | cell % - reference % (percentage points) | Newcombe hybrid-score         |
| pct  | `ratio`   | cell % / reference % (relative risk)     | Katz log-risk-ratio           |
| pct  | `OR`      | empirical odds ratio                     | Woolf log-odds-ratio          |
| pct  | `contrib` | signed χ² contribution (no reference)    | — (standardized residual)     |
| mean | `diff`    | cell mean - reference mean (SD units)    | Welch *t*                     |
| mean | `ratio`   | cell mean / reference mean               | robust ratio-of-means         |

Alternative interval methods (see `?tab_ci`): 

- `method_diff = "ac"` (Agresti-Caffo) or `"wald"` for a percentage difference; 
- `method_mean_diff = "student"` (pooled, the OLS two-group interval) for a mean difference; 
- `method_mean_ratio = "quasipoisson"` or `"poisson"` for a mean ratio. 
- The relative-risk (`method_ratio = "katz"`) and odds-ratio (Woolf) intervals have no alternative.
- A mean **difference** is colored **standardized** — Glass's Δ, the difference divided by the reference's standard deviation — so the `mean_diff` color breaks are read in SD units, unless the user provides a custom break scale.
- For a factor with **3 or more levels**, the odds ratio (and its interval) compares each level to the `ref2` baseline level : it is a relative risk ratio (RRR) (the same observed quantity that is modelised by a multinomial logistic regression). The OR interval is computed only when `color_signif` or `stars` needs it.

**Simple, cell-by-cell confidence intervals** (`ci = "cell"`) compare each cell to 0 % (or a mean of 0), *not* to a reference:

| type | color  | confidence interval (default) | other method                    |
|------|--------|-------------------------------|---------------------------------|
| pct  | `cell` | Wilson score interval         | `method_cell = "wald"` (normal) |
| mean | `cell` | one-sample Student *t* (n-1)  | —                               |

Because they compare to 0 and not to a reference, cell intervals are purely descriptive: they carry **no significance and no stars**. `method_cell` chooses `"wilson"` (default) or `"wald"` for percentages; a mean cell interval is always the one-sample Student *t*.

**`color_signif` turns that interval into a coloring policy.** 

- `"ignore"` colors **every** cell by its **observed effect** size, for example the observed difference compared to the Total row. Grey cells have an observed effect below the threshold (for example, differences of less than ±5 points of percentages). 
- `"grey_non_signif"` and `"guaranteed_effect"` both color **only significant cells**, but differ in the *intensity basis*: 
  + `grey_non_signif` colors by the **observed** effect, like "ignore", greying out small deviations, but it also **greys out any large deviation that turns to be non-significant**. Ideal for large samples.
  + `guaranteed_effect` colors by the **guaranteed** effect — the confidence-bound (CI-floor), the **smaller deviation that is assured at a given confidence level** (default 95%) — so its colors are dimmer and conservative, but **all significative differences are colored**, which is ideal for small samples.

| type | color     | `="ignore"`     | `="grey_non_signif"`               | `="guaranteed_effect"`      |
|------|-----------|-----------------|------------------------------------|-----------------------------|
| pct  | `diff`    | observed diff   | grey if the diff-CI contains 0     | diff CI-floor               |
| pct  | `ratio`   | observed ratio  | grey if the ratio-CI contains 1    | ratio CI-floor              |
| pct  | `OR`      | observed OR     | grey if the OR-CI contains 1       | OR CI-floor                 |
| pct  | `contrib` | χ² contribution | grey if residual < 1.96 (conf 95%) | residual >= 1.96 (conf 95%) |
| mean | `diff`    | observed diff   | grey if the diff-CI contains 0     | diff CI-floor               |
| mean | `ratio`   | observed ratio  | grey if the ratio-CI contains 1    | ratio CI-floor              |

Examples :

```{r, echo=TRUE, eval=FALSE}
# --- factors: percentages -------------------------------------------------
tab(gss_simple, race, party3, pct = "row", color = "diff",  color_signif = "grey_non_signif")
tab(gss_simple, race, party3, pct = "row", color = "ratio", color_signif = "guaranteed_effect")
tab(gss_simple, rincome, married, pct = "row", color = "OR", OR = TRUE, ref2 = 1)
tab(gss_simple, rincome, party3, color = "contrib")   # works with pct = "row"/"col" too

# --- numerics: means ------------------------------------------------------
tab(gss_simple, rincome, tvhours, color = "diff",  color_signif = "guaranteed_effect")
tab(gss_simple, rincome, tvhours, color = "ratio", color_signif = "grey_non_signif")

#    a custom scale for differences in means, and a "first row" reference
tab(gss_simple, rincome, tvhours, color = "diff", color_signif = "grey_non_signif",
    color_breaks = list(mean_diff = c(0.4, 0.8, 1.6)), ref = 1)
```

## Session options

A handful of `options()` set your preferred defaults once for the whole session — put them at the top of a script, or in your `.Rprofile`. Each one has a per-call argument too; the option just changes the default. The everyday ones:

- `options(tabxplor.print = "html")` — print tables not in console, but as html in RStudio or Positron Viewer Pane by default (recommended)
- `options(tabxplor.cleannames = TRUE)` — strip `"1-"`-style prefixes from level names everywhere.
- `options(tabxplor.parallel = 8)` — parallelise tables with multiples variables on different CPU cores by default (needs `mirai`)
- `options(tabxplor.var_labels = TRUE)` — in exports, show a variable's label (from `haven`/`labelled` data) instead of its bare name.
- `options(tabxplor.theme = "auto")` — the export theme (`"light"`/`"dark"`/`"auto"`); `set_color_palette(theme = "auto")` does the same for the console.
- `options(tabxplor.stars = TRUE)` — show significance stars in every table (like `stars = TRUE`).
- `options(tabxplor.conf_level = 0.9)` — the confidence level for intervals and tests (default `0.95`).
- `options(tabxplor.ci_print = "moe")` — print a confidence interval as a `pct ± margin of error` instead of a `[low; high]` bracket.
- `options(tabxplor.lang = "fr")` — the language of the colour legends and footers (`"auto"`/`"en"`/`"fr"`).

Colour thresholds and palettes have their own helpers, `set_color_breaks()` and `set_color_palette()`. `?tabxplor-options` documents every option, and `vignette("tabxplor-programming")` covers the more advanced ones (export fonts, parallel builds…).

## A point-and-click interface (jamovi)

Everything above is also available **without writing R code**, through a [jamovi](https://www.jamovi.org/download) module. jamovi is a free, open-source statistical software : install it, open the modules menu (the **`+`** at the top-right), choose **jamovi library**, and install *tabxplor*. It adds a **Crosstables** analysis — and a **Regressions** analysis powered by `tab_reg()` — with the same coloured, exportable tables, driven entirely by menus. Handy for teaching, or for colleagues who do not use R.

## Where to go next

- `vignette("tabxplor-reg")` — regression tables with `tab_reg()`, and comparing modelled to observed effects.
- `vignette("tabxplor-programming")` — the `tabxplor_fmt` cell type and how to program with its fields.
- `?tab` for every argument (grouped by purpose), `?tab_ci` for confidence-interval methods, and `?tabxplor-options` for the package-wide defaults.
