---
title: "Regression tables with tab_reg()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression tables with tab_reg()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tabxplor)

# As in the introduction vignette, tables are shown as the coloured console output turned to HTML.
options(cli.num_colors = 256)
set_color_style(type = "text", theme = "light")
```

```{r, echo = FALSE, include = FALSE}
knitr::knit_hooks$set(output = function(x, options) {
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

`tab_reg()` builds a **regression table** that looks and behaves like a `tabxplor` cross-table: one row per predictor level, significance stars, colours that grey out non-significant effects, and the same Excel, html or markdown exports. You give it a data frame, an **outcome** (`dependent`) and some **predictors**, and it tries to picks the right kind of model from the outcome's type. Its distinctive feature is `empirical = TRUE`, which shows the **observed / crude / empirical effect right next to the model's adjusted one**, so you can see what "controlling for the other variables" actually changed.

We use a formatted version of the `forcats::gss_cat` data, from the US General Social Survey:

```{r}
gss_simple <- gss_cat_data_formatting()
```


## The outcome's type chooses the model and the observed quantity to compare with

You rarely set `family` by hand — `tab_reg()` mostly detects it: 

| Outcome | Detected model | Effect measure |
|:--------|:---------------|:---------------|
| 2-level factor | binomial (logistic) | odds ratio (OR) |
| numeric (continuous) | gaussian (linear) | coefficient β |
| count | poisson | incidence-rate ratio (IRR) |
| 3+ level unordered factor | multinomial | one OR column per category vs. the reference |
| 3+ level ordered factor | ordinal (proportional odds) | cumulative OR |

With `empirical = TRUE`, each model column is joined by a crude/observed companion column showing the *empirical, unadjusted, single-predictor* effect — the effect you would see with no controls at all, "all things being unequal", for that predictor.
- 2 levels logistic (binomial) → observed %, and observed odd-ratios (not modelised: calculated from percentages only)
- 3+ levels logistic (multinomial) → the observed ORs are shown as a tooltip on the model cells in html exports
- gaussian (linear) → group means and their difference
- poisson (counts) → observed rate and observed rate ratio


## Logistic regression

When the outcome is a two-levels factor, here "married" versus "not married", `tab_reg()` chooses a **binomial** family to fit a logistic regression and reports **odds ratios** (the reference level of each predictor shows the neutral value `1`). 
- Like in every regression model, the effect of a predictor level on the dependent variable reads "all the others chosen predictors being equals".
- Colors read like any `tabxplor` table: an odds ratio above 1 (blue) means *more likely to be married* than the reference level, below 1 (red) means *less likely*; stars and colors both flag significance.

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"))
```

For a simple logistic regression, `empirical = TRUE` add raw percentages and crude unmodelised odds-ratios 

```{r}
tab_reg(gss_simple, "married", c("race", "age", "rincome", "relig"), empirical = TRUE)
```


 With `empirical = TRUE`, each model column is joined by a **crude** companion column showing the *unadjusted, single-predictor* effect — the effect you would see with no controls at all:
"all things being unequal"


The **"Emp." column is the same descriptive quantity you would get from a cross-table**. For a single predictor, the model's crude odds ratio equals `tab()`'s empirical odds ratio:

```{r}
tab(gss_simple, race, married, pct = "row", OR = "OR")
```

Comparing the two tells you what adjustment did: if a predictor's **model** OR is much closer to 1 than its **crude** OR, the raw association was largely explained by the other predictors. The empirical columns carry their own confidence intervals and stars, computed the same way as the model's, so the comparison is fair. Each family has its matching crude measure:



### Nominal and ordinal outcomes (3+ levels)

A nominal outcome with three or more unordered levels is fit as one **multinomial** model, giving one odds-ratio column per outcome category (versus the reference category):

```{r, eval = requireNamespace("nnet", quietly = TRUE)}
tab_reg(gss_simple, "party3", c("race", "age"))
```

An **ordered** outcome is fit as a proportional-odds (cumulative) model:

```{r, eval = requireNamespace("MASS", quietly = TRUE)}
tab_reg(gss_simple, "rincome", "race")
```

### Linear regression (a numeric outcome)

A continuous outcome gives plain regression coefficients (here we set `family` explicitly, because an integer like `age` is ambiguous — it could also be modelled as a count):

```{r}
tab_reg(gss_simple, "age", c("race", "marital"), family = "gaussian")
```

### Poisson regression (a count outcome)

```{r}
tab_reg(gss_simple, "tvhours", c("race", "marital"), family = "poisson")
```

An **incidence-rate ratio** of 1.2 means "20% more hours of TV". Unweighted Poisson models automatically use dispersion-scaled (quasi-Poisson) standard errors, so over-dispersed counts get honest, wider intervals.



## Weighted and survey data

Passing a weight column with `wt =` switches estimation to a **survey design** (`survey::svyglm()`), which gives correct *design-based* standard errors. The framework `tabxplor` uses is simple: the **estimate is weighted**, the number of observations in the sample in the basis for confidence intervals and pvalues calculations, and inference uses the **survey design** so that unequal weights widen the intervals honestly.

```{r, eval = FALSE}
# `weight` is a survey weight column in your data:
tab_reg(gss_simple, "married", c("race", "age"), wt = "weight")

# A stratified / clustered design (strata usually *narrow* the intervals, clusters widen them):
tab_reg(gss_simple, "married", c("race", "age"),
        wt = "weight", ids = "psu", strata = "stratum")

# Or pass a design you built yourself with survey::svydesign() as `data` (see ?survey::svydesign
# for calibration, replicate weights and the more complex cases):
```

If you only have simple weights, `wt =` is all you need. Give `strata =` when you have them: stratification typically buys a little precision (narrower confidence intervals) when the strata variables are related to the outcome.

## Comparing several models

Pass a **named list** of predictor sets instead of a vector to fit and show several models side by side, and `compare =` adds a comparison Likelihood ratio test in the footer (`"baseline"` = each model vs. the first; `"sequential"` = each vs. the previous):

```{r}
tab_reg(gss_simple,
        "married",
        list("Race only"    = "race",
             "+ age"        = c("race", "age"),
             "+ party"      = c("race", "age", "party3")),
        compare = "sequential")
```


## Reading the footer statistics

The footer summarises model fit; which statistics appear depends on the family:

- **N** — the number of observations used.
- **LR vs null** — a likelihood-ratio test of the whole model against the intercept-only model (is the model worth anything?).
- **McFadden R²** — a pseudo-R² for binomial and poisson models (higher = the predictors explain more; values are much smaller than a linear R²).
- **AIC / BIC** — information criteria for comparing models on the same data (lower = better; BIC penalises complexity more).
- Linear models (gaussian) add the usual **R² / adjusted R² / F / σ**; Poisson adds a **dispersion** check.

For weighted models a reduced set is reported (a Wald test against the null, Nagelkerke / Cox–Snell pseudo-R², a Rao–Scott AIC), because the likelihood-based quantities do not apply to a design-based fit.

## Plots

`or_plot()` draws a forest plot of the odds ratios of a finished table, and `lm_plots()` draws the standard 2×2 linear-model diagnostics:

```{r, eval = FALSE}
t <- tab_reg(gss_simple, "married", c("race", "age"))
or_plot(t)

m <- tab_reg(gss_simple, "age", c("race", "marital"), family = "gaussian")
lm_plots(m)
```

## Where to go next

- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `?tab_reg` for every argument (grouped by purpose), and `?tab_reg` &rarr; *Details* for the modelling choices.
- `survey::svydesign()` for complex designs (calibration, replicate weights, finite-population corrections).
