---
title: "Programming with tabxplor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with tabxplor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tabxplor)

# Pin the legend language: it defaults to "auto" = the ambient locale, so building this English
# vignette on a French machine silently renders French legends and captions (the -fr articles pin
# "fr" for the same reason). Output must not depend on where it is built.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")   # the test-summary / model-fit row labels go through gettext, not this option
library(dplyr)

# Tables render as tabxplor's real html tables (the recommended everyday setting); the shared
# stylesheet is emitted once by tab_css() below, and the hover tooltips are kept off here.
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)

# Console outputs (vectors, fields...) keep their terminal colors, turned to html by fansi.
options(cli.num_colors = 256)
set_color_palette(theme = "light")
```

```{r, echo = FALSE, results = "asis"}
cat(tab_css(theme = "light"))
```

```{r, echo = FALSE, include = FALSE}
# Colour the console outputs (ANSI -> html, via fansi), but hand as-is results (the html tables,
# marked by knitr with an ASIS token) back to knitr's default hook untouched.
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (grepl("KNITR_ASIS_OUTPUT_TOKEN", x, fixed = TRUE)) return(default_output_hook(x, options))
  paste0('<pre class="r-output"><code>',
         fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
         '</code></pre>')
})
```

This vignette is for users who want to **program with** `tabxplor` — write their own helpers, pull the underlying numbers out, or reshape a table cell by cell. If you only want to make and read tables, `vignette("tabxplor")` is the place to start.

Every numeric column of a `tabxplor` table is a single vector of class **`tabxplor_fmt`** ("formatted number"). It is a [`vctrs` record](https://vctrs.r-lib.org/reference/new_rcrd.html): behind the one value you see printed, each cell stores **all the data needed to compute the displayed number, its format and its colour** — counts, percentages, differences, relative risks, confidence-interval bounds, odds-ratios, and so on. Because it is a proper vector, it survives every `dplyr` verb, and you can read or rewrite any of its fields.

```{r}
gss_simple <- gss_cat_data_formatting()
tabs <- tab(gss_simple, race, marital, pct = "row", color = "diff")
```

## Getting the plain numbers out

The quickest way to recover the underlying numbers as ordinary numeric vectors is `get_num()`, which extracts whatever field is currently displayed:

```{r}
tabs |> mutate(across(where(is_fmt), get_num))
```

To get the character strings instead (formatted, but without colours), use `format()`:

```{r, eval = FALSE}
tabs |> mutate(across(where(is_fmt), format))
```

Individual fields are read most simply with `$` on the fmt column (see "Reading and changing fields" below):

```{r}
tabs$Married$pct
```

## The fields of a cell

A `tabxplor_fmt` cell carries **19 fields** (many are `NA` when the relevant quantity was not requested). The user-facing ones are:

| Field | Meaning |
|:------|:--------|
| `n`         | unweighted count (integer) |
| `wn`        | weighted count |
| `pct`       | percentage |
| `mean`      | mean (numeric column variables) |
| `diff`      | difference from the total / reference cell |
| `ratio`     | ratio to the reference (the "×2 rule"; relative risk) |
| `ci_inf`, `ci_sup` | confidence-interval bounds |
| `pvalue`    | per-cell significance p-value (feeds the stars) |
| `or`        | odds ratio / relative-risk ratio |
| `ctr`       | contribution to the chi-squared (`color = "contrib"`) |
| `resid`     | adjusted standardized residual (derived from `pvalue`, read-only) |
| `var`       | variance (numeric columns; chi-squared variance with `pct`) |
| `tot_n`     | the cell's own base — the count its percentage is computed on |
| `n_eff`     | effective sample size used for the CI, with `options(tabxplor.design_effect = TRUE)` or a `survey` design |
| `obs`       | `tab_reg()` only: the observed (crude) effect the modelled one is compared to |
| `digits`    | number of decimals to display (per cell) |
| `display`   | which field is shown (per cell) |
| `in_totrow`, `in_tottab`, `in_refrow` | is the cell in a total row / total table / reference row (logical) |

See the whole underlying data frame of a column with `vctrs::vec_data()`:

```{r}
vctrs::vec_data(tabs$Married)
```

## Reading and changing fields

Read a field with `$` (the friendliest way), or `vctrs::field()`:

```{r, eval = FALSE}
tabs$Married$pct
tabs |> mutate(across(where(is_fmt), ~ .$pct))
tabs |> mutate(across(where(is_fmt), ~ vctrs::field(., "pct")))
```

Confidence intervals are stored as their two bounds, the `ci_inf` and `ci_sup` fields:

```{r}
ci_tab <- tab(gss_simple, race, marital, pct = "row", ci = "cell")
ci_tab$Married$ci_inf
ci_tab$Married$ci_sup
```

Switch the displayed field with `set_display()`:

```{r, eval = FALSE}
tabs |> set_display("diff")
tabs |> mutate(across(where(is_fmt), ~ set_display(., "diff")))
```

To change a field, the easiest route is `dplyr::mutate()` **on the fmt vector itself** — a `tabxplor_fmt` behaves like a little data frame of its fields. For example, round every cell to two decimals:

```{r, eval = FALSE}
tabs |> mutate(across(where(is_fmt), ~ mutate(., digits = 2L)))
```

A fuller example: turn a means table's variance into a standard deviation and add it as a new, un-coloured column shown to one decimal:

```{r, eval = FALSE}
tab_num(gss_simple, race, c(age, tvhours), digits = 1L) |>
  mutate(across(
    c(age, tvhours),
    ~ mutate(., var = sqrt(var), display = "var", digits = 1L) |> set_color("no"),
    .names = "{.col}_sd"
  ))
```

## Composite display: combining fields

`set_display()` chooses *one* field to show. To show **several fields in one cell**, give it a `{}` **template** instead of a bare field name — the same template you can pass to `tab(display =)` when you build the table. The template is plain text with `{field}` tokens; each token is replaced by that field, formatted as usual:

```{r}
tabs |> set_display("{pct} ({diff})")
```

The rules:

- **Valid fields** are `pct`, `n`, `wn`, `mean`, `diff`, `ratio`, `ci`, `or`, `ctr`, `var`, `resid`, `obs` — the same set as the single-field displays.
- The **first** token is the *primary* field: it is what `get_num()` returns, the value Excel keeps (Excel shows the primary only), and the field the colours read.
- A **bare field name** is shorthand for its own template, so `set_display("ci")` is exactly `set_display("{ci}")`.
- It is a **display overlay** for text output (the console, `tab_kable()`, `tab_md()`): the stored fields are untouched, so you can switch back at any time.
- The `{ci}` field already prints its own `[…;…]` brackets, so write `"{pct} {ci}"` — **not** `"{pct} [{ci}]"`, which would double them.

## Creating a column from another field

Because switching the display recomputes nothing — every field is already stored in the cell — you can spin off a **new column that shows a different field**. A percentage table has a reference by default (the Total row), so its `diff` field is already filled; a difference twin of every percentage column is then one `mutate()`:

```{r}
tab(gss_simple, race, marital, pct = "row") |>
  mutate(across(where(is_fmt), ~ set_display(., "diff"), .names = "{.col}_diff"))
```

`.names = "{.col}_diff"` keeps the original percentage columns and adds a `<name>_diff` twin next to each. (If a `diff` column comes out empty, the source table had no reference cell — build it with `ref =` / `comp =`, or with `color = "diff"`.)

## Total rows, reference rows and columns

Helper predicates let you act on structural parts of a table: `is_totrow()` / `is_tottab()` / `is_refrow()` (cell-level, logical vectors), and `is_totcol()` / `is_refcol()` (column-level). They are what keeps totals in place when you re-order or when you write conditional formatting:

```{r, eval = FALSE}
# more decimals on the total row than on the body:
tab(gss_simple, race, marital, race, pct = "row") |>
  mutate(across(
    where(is_fmt),
    ~ if_else(is_totrow(.), mutate(., digits = 1L), mutate(., digits = 2L))
  ))
```

## Column attributes

Besides its per-cell fields, each `fmt` column carries a few **column-level attributes**, read and set with `get_*` / `set_*` (or `is_*` / `as_*` for the logical ones):

- `scale` — `get_scale()` / `set_scale()`: **what the column estimates**. A level (`"level_n"`, `"level_pct"`, `"level_mean"`), a difference (`"points"` for percentage points, `"mean_diff"`, `"raw_diff"` for a regression coefficient in the outcome's own units), a ratio (`"pct_ratio"`, `"mean_ratio"`, `"odds_ratio"`) or a link-scale coefficient (`"log_coef"`). It says which field holds the estimate, what its null value is, and which colour ladder it is read on.
- `pct_base` — `get_pct_base()` / `set_pct_base()`: for a percentage, what it is a percentage *of* — `"row"`, `"col"`, `"all"`, `"all_tabs"`, or `"none"` for counts, means and coefficients.
- `ci_method` — `get_ci_method()`: which interval engine built this column's bounds (`"wilson"`, `"newcombe"`, `"welch"`, `"katz"`, …; `""` when the column carries no interval).
- `color` — `get_color()` / `set_color()`: the colour measure of the column (`""`, `"no"`, `"diff"`, `"ratio"`, `"contrib"`, `"OR"`).
- `col_var` — `get_col_var()`: the name of the column variable (a table can hold several).
- `comp_all` — `get_comp_all()`: with `tab_vars`, is the comparison reference the sub-table (`FALSE`) or the whole table (`TRUE`)?
- `totcol` / `refcol` — `is_totcol()` / `is_refcol()`: is this a total column / a reference column?

## Building cells from scratch

`fmt()` constructs an `fmt` vector directly from numbers — handy for tests, or to add a hand-computed column. Everything the display and colour engine needs can be supplied:

```{r}
fmt(n = c(10L, 20L, 30L), pct = c(0.1, 0.2, 0.7), display = "pct", digits = 0L)
```

## Tables from pre-aggregated counts

Sometimes the data already arrives **cross-tabulated** — a counts table from a report, a `table()`, a matrix of frequencies. `tab_counts()` builds the same coloured `tabxplor` table as `tab()`, but from those counts instead of the raw records; every calculation runs on the counts, so the result is identical to what `tab()` would have produced from individual records.

The commonest input is a **tidy counts** data frame (one row per combination, the count in a column):

```{r}
counts <- dplyr::count(gss_simple, marital, race)
tab_counts(counts, marital, race, counts = n, pct = "row", color = TRUE)
# identical to tab(gss_simple, marital, race, pct = "row", color = "diff")
```

It also melts a `table()` / `xtabs()` / matrix automatically, and reads a **wide** frame (one column per column-variable level) via `cols =` / `col_name =`:

```{r, eval = FALSE}
tab_counts(table(gss_simple$marital, gss_simple$race), pct = "row", color = "diff")

wide <- tidyr::pivot_wider(counts, names_from = race, values_from = n)
tab_counts(wide, row_var = marital, cols = c(White, Black, Other),
           col_name = "race", pct = "row", color = "diff")
```

Weighting works as in `tab()`: pass the unweighted count in `counts` and the weighted count in `wt_counts` — estimates use the weighted count, confidence intervals and Chi-2 use the unweighted N. (`options(tabxplor.design_effect = TRUE)` needs the individual weights, which pre-aggregated counts no longer carry, so it does not apply here — the CIs use the unweighted N, and the table's footer says so rather than claiming a correction it does not have.) When the only figures available are non-whole numbers (percentages × a base, or weighted-only counts), CIs and the Chi-2 test are disabled with a message.

## Pivoting a grouped table into columns

A **grouped** table (built with `tab_vars`) stacks one sub-table per group. `tab_spread()` pivots a grouping variable's levels into **side-by-side columns** — ideal to compare one measure across groups. Select the measure column you want first, then spread:

```{r}
tab(gss_simple, relig, marital, year, pct = "row", totaltab = "no", tot = "row") |>
  dplyr::select(year, relig, Married) |>
  tab_spread(year)
```

`tab(..., spread_vars = year)` does the same in one call. The regression vignette's `split_var` produces a grouped table you can spread the same way — see `vignette("tabxplor-reg")`.

## A score from several factors

`score_from_lv1()` turns a battery of yes/no (or agree/disagree) factors into a single **summed score**: for each row it counts how many of the listed factors sit at their **first level**. Here we score how many of two markers — being married, and being in the higher income bracket — each respondent has:

```{r}
gss_simple |>
  score_from_lv1("score", vars_list = c("married", "income25k")) |>
  tab(relig, score)
```

The score ranges from 0 to the number of factors (missing values never count). Such a summed score is exactly the input a **grouped-binomial** regression models — see `trials` in `vignette("tabxplor-reg")`.

## Building many tables at once

`tab_many()` — the engine behind `tab()` — accepts **lists of arguments** to build several differently-shaped tables in one call. For a fully data-driven batch, `purrr::pmap()` over a small specification table is the idiom — one row per table, one column per argument:

```{r, eval = FALSE}
specs <- tibble::tribble(
  ~row_var, ~col_var,  ~pct,
  "race",   "marital", "row",
  "relig",  "party3",  "row",
)
purrr::pmap(specs, \(row_var, col_var, pct)
            tab(gss_simple, all_of(row_var), all_of(col_var), pct = pct))
```


## Advanced options

Beyond the everyday defaults (`vignette("tabxplor")` &rarr; *Session options*), these `options()` tune the exporters and the build. HTML / `tab_kable()`:

- `tabxplor.tab_kable_engine` — `"html"` (dependency-free, the default) or `"kableExtra"` (legacy).
- `tabxplor.tab_kable_css` — inline the stylesheet with each table (`TRUE`); set `FALSE` in a many-table document and call `tab_css()` once instead (this vignette does).
- `tabxplor.tab_kable_tooltips` — the per-cell hover tooltips (`TRUE`); set `FALSE` once per document to switch them off everywhere.
- `tabxplor.tab_kable_num_font` / `tabxplor.kable_html_font` — the number and text CSS font stacks.
- `tabxplor.kable_popover` — click popovers instead of hover tooltips.

Excel / `tab_export("xl")`:

- `tabxplor.xl_font_text` / `tabxplor.xl_font_num` / `tabxplor.xl_font_num_stars` — the label, number and starred-number fonts.
- `tabxplor.xl_or_numeric` — keep odds ratios as numbers rather than `1/x` text.

Console, stats and paths:

- `tabxplor.console_bold` — embolden reference / total / coloured cells (auto-detected per editor).
- `tabxplor.signif_levels` / `tabxplor.signif_labels` — the p-value cut-offs and the star labels.
- `tabxplor.plot_num_font` — `tab_plot()`'s number font; `tabxplor.export_dir` — the default export directory.

Performance and integration:

- `tabxplor.parallel` — build one worker per row variable on a background pool (needs `mirai`); `tabxplor.parallel_min` sets the smallest row-variable count worth dispatching. Release the pool with `tab_parallel_stop()`.

See `?tabxplor-options` for the full list and every default.

## See also

- `vignette("tabxplor")` — cross-tables and the colour helpers.
- `vignette("tabxplor-reg")` — regression tables.
- `?fmt`, `?tab_num`, `?tab_ci` for the reference documentation of the type and its building blocks.
