---
output:
  github_document:
    html_preview: false
    # a lone image stays `![alt](src)`, not a pandoc <figure> whose alt text becomes a caption
    md_extensions: -implicit_figures
always_allow_html: true
---

<!-- pkgdown/index.md is generated from pkgdown/index.Rmd. Please edit that file.
     This page IS README.Rmd in colour -- same prose, same examples, the tables rendered in the
     real palette instead of the black-and-white one GitHub forces. Edit both, or neither. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/index-",
  out.width = "100%"
)
library(tabxplor)
options(tabxplor.print = "html")
options(tabxplor.tab_kable_css = FALSE)
options(tabxplor.tab_kable_tooltips = FALSE)
# Pin the language, BOTH ways -- this English page must not depend on who builds it (knitting it
# on the maintainer's fr_FR.UTF-8 box otherwise yields French legends, captions and GOF labels).
# Two levers are needed because they drive different strings:
#   - options(tabxplor.lang) drives the colour legend / footer / reg caption;
#   - the test-summary and model-fit row labels (reg_footer_spec / test_pvalue_descriptor /
#     test_es_measure, R/tab-test-display.R) resolve through gettext on the AMBIENT locale, which
#     only LANGUAGE reaches. Without it a model-check row heads "Linéarité", not "Linearity".
#     (Notation is deliberately NOT translated -- "LR vs null", OR/IRR/beta stay as they are.)
# Same discipline as set_color_palette(theme=) below and tests/testthat/setup.R.
options(tabxplor.lang = "en")
Sys.setenv(LANGUAGE = "en")
set_color_palette(theme = "light")
```

# tabxplor

`tabxplor` makes cross-tables and regression models readable at a glance for data exploration. It builds a table with percentages, weighted counts, confidence intervals, tests — and colors highlight the cells that stand out from the total or reference, only when the difference is statistically solid, to spot the structure of your data immediately.

- **Colors encode effect size and significance**: the stronger the difference, the deeper the color; non-significant cells are greyed-out.
- Html, Excel and markdown/Quarto exports are available.
- It comes with a point-and-click [jamovi](https://www.jamovi.org/) graphical interface: no code needed.
- A black-and-white `theme = "print_ready"` renders the same reading for journals.
- **Regression models** are presented with the same visual language, next to their observed effect.
- In R the tables **are `tibble`s you can keep working on with `dplyr`**. Cells are rich values, each one carries its count, percentage, confidence interval and reference behind the displayed number.
- Weighted data and survey design are supported.

```{r css, echo = FALSE, results = "asis"}
# ONE stylesheet for the whole page (tab_kable_css = FALSE, above, stops each table re-inlining it).
# "auto" follows the site's light/dark switch through the `[data-bs-theme]` hooks bootstrap sets.
cat(tab_css(theme = "auto"))
```

## Installation

``` r
install.packages("tabxplor", dependencies = TRUE)
```

## A quick look

A simple cross-table with row percentages: shades of blue mean the cell is over-represented compared to the total row, shades of yellow to red mean it is under-represented.

```{r quick-look}
gss <- gss_cat_data_formatting() # cleaned-up version of forcats::gss_cat

tab(gss, race, party3, pct = "row", color = "difference")
```

Several column variables can be crossed at once for series of Yes/No survey questions. With `color_signif = "grey_non_signif"`, cells that are not significantly different from the total are greyed out, so every colored figure is a solid one. Use `wt =` for weighted or survey data. Example with [FactoMineR](http://factominer.free.fr/index_fr.html) tea data :

```{r quick-look-signif}
tea_when_vars <- c("breakfast", "tea.time", "evening", "lunch", "dinner", "always")
tab(facto_tea, SPC, all_of(tea_when_vars), pct = "row", 
    levels = "first", na = "drop", 
    color = "difference", ref = "first", color_signif = "grey_non_signif")
```

The same visual language extends to regression models: `tab_reg()` detects a binary outcome and fits a logistic regression, coloring odds ratios by strength and greying the non-significant ones, with a default comparison between the modelised deviations and their crude/observed counterparts.

```{r quick-look-reg, message = FALSE}
tab_reg(gss, outcome = "married", predictors = c("race", "age", "rincome"))
```

Or as a black and white table ready for publication:

```{r print-ready-reg, message = FALSE}
options(tabxplor.theme = "print_ready")
tab_reg(gss, outcome = "married", predictors = c("race", "age", "rincome"))
```

```{r include=FALSE}
options(tabxplor.theme = "light")
```


## Export your tables

Any table exports with its colors to Excel, html or markdown (for Word, copy-paste from Excel) :

``` r
tab(gss, marital, race, pct = "row", color = "difference") |> tab_html()
tab(gss, marital, race, pct = "row", color = "difference") |> tab_xl()
tab(gss, marital, race, pct = "row", color = "difference") |> tab_xl(theme = "print_ready")
```



## Learn more

- [Introduction to tabxplor](articles/tabxplor.html) — the place to start (*aussi disponible [en français](articles/tabxplor-fr.html)*).
- [Regression tables with tab_reg()](articles/tabxplor-reg.html) (*aussi disponible [en français](articles/tabxplor-reg-fr.html)*).
- [Reading a regression without losing sight of the percentages](articles/tabxplor-reading-a-regression.html) — a single analysis walked from a first cross-table to a finished sentence (*aussi disponible [en français](articles/tabxplor-reading-a-regression-fr.html)*).
- [Weighted and survey data](articles/tabxplor-weights.html) — the three levels of margin of error, and which one your file deserves (*aussi disponible [en français](articles/tabxplor-weights-fr.html)*).
- [Programming with tabxplor](articles/tabxplor-programming.html) — many tables at once, custom workflows, options (*aussi disponible [en français](articles/tabxplor-programming-fr.html)*).
